---
title: "HD vs MM – scRNA-seq All T Cell Analysis"
subtitle: "Integrated multi-sample reanalysis"
author: 
  - name: Jordan Dutel
    affiliation: 
      - CRCT – CHU Toulouse
      - GENIM Team
    email: jordan.dutel@inserm.fr
date: today
date-format: "DD MMMM YYYY"

lang: en
title-block-banner: true
title-block-style: default

format:
  html:
    theme:
      light: cosmo
      dark: darkly
    # theme: superhero         # Bootstrap
    css: styles.css
    toc: true
    toc-title: Table of Contents
    toc-depth: 7
    toc-location: left
    toc-expand: 2
    number-sections: true
    number-depth: 7
    page-layout: full
    smooth-scroll: true
    anchor-sections: true
    section-divs: true
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    code-copy: hover
    code-overflow: wrap
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    highlight-style: github
    embed-resources: true
    fig-responsive: true
    fig-format: retina
    fig-align: center
    fig-dpi: 300
    df-print: paged
    other-links:
      - text: Thomas's PPT
        icon: filetype-ppt
        href: ../00_context/PPT_Thomas_BCL6_09012026.pdf 
    
execute:
  echo: true
  warning: false
  message: false
  error: false
  # cache: true
  freeze: auto
  eval: true

editor: visual
---

```{r Setup, echo=FALSE}
############
# PACKAGES #
############

# BiocManager::install("SingleCellExperiment")
# install.packages("devtools")
# BiocManager::install("DropletUtils")
# devtools::install_github("immunogenomics/presto")
# BiocManager::install("scDblFinder")

library(DropletUtils)
library(SingleCellExperiment)
library(scater)
library(scran)
library(qs)
library(Seurat)
library(ggplot2)
library(patchwork)
library(clustree)
library(harmony)
library(patchwork)
library(dplyr)
library(ggridges)
library(readr)
library(knitr)
library(kableExtra)
library(SingleR)
library(celldex)
library(ComplexHeatmap)
library(circlize)
library(presto)
library(tidyverse)
library(tibble)
library(matrixStats)
library(viridis)
library(scuttle)
library(scDblFinder)
library(ggbeeswarm)
library(here)
library(scales)
library(Polychrome)
library(future)
library(glmGamPoi)
library(lisi)
library(pheatmap)
library(cluster)
library(ggrepel)
library(rstatix)
library(ggpubr)
library(pals)
library(IRanges)
library(sp)
library(SummarizedExperiment)
library(furrr)
library(fgsea)
library(msigdbr)
library(Seurat)
library(edgeR)
library(gt)
library(DT)
library(mclust)

###########
# OPTIONS #
###########

options(
  stringsAsFactors = FALSE,
  scipen = 999
)

options(future.globals.maxSize = 4 * 1024^3)  # 4 Go

plan(sequential)

#########
# PATHS #
#########

here::i_am("04_report/HD_vs_MM_scRNAseq_CD4_Tcells_analysis.qmd")
# here::i_am("02_script/HD_vs_MM_scRNAseq_reanalysis.qmd")

# knitr::opts_knit$set(
#   root.dir = "~/Documents/Project/Thomas/20260116_scRNAseq_HD_MM_reanalysis")

seurat_path_Tcell_obj <- "/mnt/crct-share/crct13/PPA/PPA_work/project/01_HCvsMM/01A_HCvsMM_10x/results/Tcells_analysis/seurat_objects/QCed_seurat_Tcells_clean.qs"

# DIR_RAW    <- here("01_raw")
DIR_DATA   <- here("01_data")
# DIR_PROCESSED_DATA <- here("02_processed_data")
DIR_SCRIPT <- here("02_script")
DIR_RESULT <- here("03_result")
DIR_REPORT <- here("04_report")
DIR_ENV <- here("05_environment")
DIR_LOG <- here("06_log")
# DIR_QC     <- here("03_result", "QC")

dir.create(here(DIR_DATA, "Processed_data"),
           recursive = TRUE, 
           showWarnings = FALSE)



##########
# COLORS #
##########

colors_30 <- c(
  "#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00",
  "#A65628", "#F781BF", "#999999", "#66C2A5", "#FC8D62",
  "#8DA0CB", "#E78AC3", "#A6D854", "#FFD92F", "#E5C494",
  "#B3B3B3", "#1B9E77", "#D95F02", "#7570B3", "#E7298A",
  "#66A61E", "#E6AB02", "#A6761D", "#666666", "#8DD3C7",
  "#BEBADA", "#FB8072", "#80B1D3", "#FDB462", "#B3DE69"
)
colors_55 <- c(
  "#E41A1C","#377EB8","#4DAF4A","#984EA3","#FF7F00",
  "#A65628","#F781BF","#999999","#66C2A5","#FC8D62",
  "#8DA0CB","#E78AC3","#A6D854","#FFD92F","#E5C494",
  "#1B9E77","#D95F02","#7570B3","#E7298A","#66A61E",
  "#E6AB02","#A6761D","#666666","#8DD3C7","#BEBADA",
  "#FB8072","#80B1D3","#FDB462","#B3DE69","#FCCDE5",
  "#D9D9D9","#BC80BD","#CCEBC5","#FFED6F","#1F78B4",
  "#33A02C","#FB9A99","#E31A1C","#FDBF6F","#CAB2D6",
  "#6A3D9A","#FFFF99","#B15928","#8E0152","#DE77AE",
  "#7FBC41","#4EB265","#4393C3","#2166AC","#B2182B",
  "#F4A582","#92C5DE","#FDDBC7","#D1E5F0","#542788"
)

# 22 col palette
pal <-c('#e6194B', '#3cb44b', '#ffe119', '#4363d8', '#f58231', '#911eb4', '#42d4f4', '#f032e6', '#bfef45', '#fabed4', '#9A6324', '#469990', '#dcbeff', '#fffac8', '#800000', '#aaffc3', '#808000', '#ffd8b1', '#000075', '#a9a9a9', '#ffffff', '#000000')

Glasbey = glasbey.colors(32)
# swatch(Glasbey)


# Define batch variable, if modalities in batch var are <= 22 use pal, if < 30 use colors_30, otherwise use colors_55
batch_var <- "emulsion"
num_batches <- 42 # replace with actual number of batches in your dataset
if (num_batches <= 22) {
  pal <- pal
} else if (num_batches <= 30) {
  pal <- colors_30
} else {
  pal <- colors_55
}
  

log_time <- format(Sys.time(), "%Y%m%d_%H%M%S")




# --- Sanity check ---
cat("Project root:", here(), "\n")
```

<br>

------------------------------------------------------------------------

------------------------------------------------------------------------

<br>

# Load Tcells {.tabset}

## Load raw data

If not already available in memory, the T cell was loaded for analysis.

```{r Load raw data}
# Load qs object
seu <- qread(seurat_path_Tcell_obj)
```

## Save a copy of the raw data {.tabset}

```{r Save raw data}
# Save qs object
qs::qsave(
  seu,
  file = file.path(
    DIR_DATA,
    "Processed_data",
    "01_hd_mm_scrnaseq_tcell_raw_seurat.qs")
)
```

## Statistics and object size {.tabset}

```{r Statistic object}
# Get statistic from this object
size_table <- tibble(
  Slot = c("Total object", "RNA assay", "ADT assay",
           "Metadata", "RNA@data", "RNA@counts", "RNA@scale.data"),
  Size_GB = c(
    object.size(seu),
    object.size(seu@assays$RNA),
    object.size(seu@assays$ADT),
    object.size(seu@meta.data),
    object.size(seu@assays$RNA@data),
    object.size(seu@assays$RNA@counts),
    object.size(seu@assays$RNA@scale.data)
  )
) %>%
  mutate(Size_GB = format(Size_GB, units = "GB"))

size_table

sizes <- list(
  cells        = length(Cells(seu)),
  object       = format(object.size(seu), units = "GB"),
  RNA_assay    = format(object.size(seu@assays$RNA), units = "GB"),
  ADT_assay    = format(object.size(seu@assays$ADT), units = "GB"),
  metadata     = format(object.size(seu@meta.data), units = "GB"),
  RNA_data     = format(object.size(seu@assays$RNA@data), units = "GB"),
  RNA_counts   = format(object.size(seu@assays$RNA@counts), units = "GB"),
  RNA_scale    = format(object.size(seu@assays$RNA@scale.data), units = "GB")
)

cat(sprintf(
"Dataset summary
----------------
Number of T cells : %s
Total object size : %s

Assay sizes:
- RNA assay       : %s
- ADT assay       : %s
- Metadata        : %s
- RNA@data        : %s
- RNA@counts      : %s
- RNA@scale.data  : %s
",
sizes$cells,
sizes$object,
sizes$RNA_assay,
sizes$ADT_assay,
sizes$metadata,
sizes$RNA_data,
sizes$RNA_counts,
sizes$RNA_scale
))

DefaultAssay(seu) <- "RNA"
seu[["RNA"]]@scale.data <- matrix(0, 0, 0)  # supprime l’ancien scaled data
invisible(gc())  # force garbage collection pour libérer la RAM
```

## Quick processing

```{r Quick processing}
seu <- seu %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData(features = rownames(seu)) %>%
  RunPCA(features = VariableFeatures(.)) %>%
  RunUMAP(dims = 1:20)

# --- Cleanup: remove scale.data to save memory ---
DefaultAssay(seu) <- "RNA"
seu[["RNA"]]@scale.data <- matrix(0, 0, 0)  # removes scaled data
invisible(gc())  # forces garbage collection to free RAM

qs::qsave(
  seu,
  file = file.path(
    DIR_DATA,
    "Processed_data",
    "02_hd_mm_scrnaseq_tcell_quick_processed_seurat.qs")
)
```

<br>

------------------------------------------------------------------------

------------------------------------------------------------------------

<br>

# QC {.tabset}

## Strategy {.tabset}

### Pre-filtering Metrics Assessment {.tabset}

For each QC metric, two batch-aware annotations should be added to the metadata (computed per emulsion):

-   `isOutlier_metric`: logical classification obtained from `isOutlier()`. A categorical variable (*low*, *normal*, *high*) based on ±3×MAD from the batch-specific median.
-   `??? An other technic` Maybe try to add a second annotation to have a continuous measure of how much a cell is an outlier for a given metric.

This design separates **metric computation**, **annotation**, and **filtering**, ensuring reproducibility and flexible downstream thresholding.

::: callout-important
All QC metrics and outlier detection procedures must be computed on the **raw counts matrix (`counts` slot)** !

Performing QC on normalized or scaled data introduces bias and invalidates threshold interpretation.
:::

### Outlier Detection Method

Outlier identification relies on the `isOutlier()` function from the scuttle package. This approach:

-   Uses robust statistics (median and MAD)
-   Accounts for distributional asymmetry
-   Allows batch-wise evaluation

This is statistically more reliable than applying a single global ±3×MAD cutoff across the entire dataset.

::: callout-tip
Always perform outlier detection **per technical batch (emulsion)** to control for batch-driven variability. Global thresholds can remove biologically valid cells from lower-depth or distinct batches !!!
:::

### Global QC Strategies

#### Stringent QC

Remove cells identified as outliers (by ≥1 method) for **at least one QC metric**.

#### Non-stringent QC

Remove cells identified as outliers by **both methods across all selected metrics**, or across a predefined subset of *critical metrics*.

::: callout-caution
QC stringency should be driven by biologically and technically critical metrics (e.g., library size, number of detected genes, mitochondrial proportion). Applying strict filtering uniformly across all computed metrics may result in excessive cell loss.
:::

### Recommended Workflow Structure

1.  Compute metrics from `counts`
2.  Annotate metadata with batch-specific outlier labels
3.  Define per-metric stringency
4.  Define global QC logic
5.  Apply filtering in a single, transparent subsetting step

This approach ensures interpretability, reproducibility, and precise control over QC stringency across batches.

## Pre-filtering QC {.tabset}

```{r Copy pre-QC object}
# Create a copy of the pre-QC object to work on and save QC updates on this object
qsave(
  seu,
  file = file.path(
    DIR_DATA,
    "Processed_data",
    "03_hd_mm_scrnaseq_tcell_qc_seurat.qs")
)
```

### Cells count {.tabset}

```{r QC Cell count}
log_dir <- file.path(DIR_LOG, "qc", "cell_count")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "qc", "cell_count", "qc_cell_count.R"),
  cmdargs = c(
    "-i", file.path(DIR_DATA, "Processed_data", "03_hd_mm_scrnaseq_tcell_qc_seurat.qs"),
    "-d", "All_Tcells",
    "-o", DIR_RESULT,
    "-q", "Before_QC"
  ),
  stdout = file.path(DIR_LOG, "qc", "cell_count", paste0("output_qc_cell_count_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "qc", "cell_count", paste0("error_qc_cell_count_", log_time, ".log"))
)
```

```{r display_plots Cell count Viz, echo=FALSE, out.width = "800px"}
img_dir <- file.path(DIR_RESULT, "All_Tcells", "QC", "Cell_Counts", "Before_QC")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(include_graphics(img, auto_pdf = TRUE, dpi = 150))

}
```

Apparently, there is also an emulsion T23491m_NK with no cells but it is not a Tcell but an NK cell. We will remove it from the dataset to avoid any bias in our analysis.

```{r Remove patients id "T23491m_NK"}
seu <- 
  qread(file.path(DIR_DATA, "Processed_data", "03_hd_mm_scrnaseq_tcell_qc_seurat.qs")) %>% 
  subset(subset = emulsion != "T23491m_NK") # Remove the patient id "T23491m_NK" which is not a CD8 Tcell but an NK cell (based on the annotation and the expression of CD8a and CD8b)
  
qsave(
  seu,
  file = file.path(
    DIR_DATA,
    "Processed_data",
    "03_hd_mm_scrnaseq_tcell_qc_seurat.qs")
)
```

### Mitochondria {.tabset}

We will now look at the distribution of the percentage of mitochondrial genes expression to see if there are some cells with high mitochondrial content that could be removed.

```{r QC_mito, echo=FALSE}

log_dir <- file.path(DIR_LOG, "qc", "mitochondria")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "qc", "mitochondria", "qc_mitochondria.R"),
  cmdargs = c(
    "-i", file.path(DIR_DATA, "Processed_data", "03_hd_mm_scrnaseq_tcell_qc_seurat.qs"),
    "-b", "emulsion",
    "-d", "All_Tcells",
    "-o", DIR_RESULT,
    "-q", "Before_QC",
    "-s", "TRUE"
  ),
  stdout = file.path(DIR_LOG, "qc", "mitochondria", paste0("output_qc_mitochondria_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "qc", "mitochondria", paste0("error_qc_mitochondria_", log_time, ".log"))
)

```

```{r display_plots 2, echo=FALSE, out.width = "800px"}

# Affichage automatique des plots

library(knitr)

img_dir <- file.path(DIR_RESULT, "All_Tcells", "QC", "Mitochondria", "Before_QC")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(knitr::include_graphics(img))

}

```

### Ribosomes {.tabset}

We will now look at the distribution of the percentage of ribosomal genes expression to see if there are some cells with high or low ribosomal content that should be removed.

```{r QC_ribo, echo=FALSE}

log_dir <- file.path(DIR_LOG, "qc", "ribosome")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "qc", "ribosome", "qc_ribosome.R"),
  cmdargs = c(
    "-i", file.path(DIR_DATA, "Processed_data", "03_hd_mm_scrnaseq_tcell_qc_seurat.qs"),
    "-b", "emulsion",
    "-d", "All_Tcells",
    "-o", DIR_RESULT,
    "-q", "Before_QC",
    "-s", "TRUE"
  ),
  stdout = file.path(DIR_LOG, "qc", "ribosome", paste0("output_qc_ribosome_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "qc", "ribosome", paste0("error_qc_ribosome_", log_time, ".log"))
)

```

```{r display_plots 3, echo=FALSE, out.width = "800px"}

# Affichage automatique des plots

img_dir <- file.path(DIR_RESULT, "All_Tcells", "QC", "Ribosomal", "Before_QC")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(knitr::include_graphics(img))

}

```

### Library size (nCountRNA) {.tabset}

```{r QC_lib_size, echo=FALSE}

log_dir <- file.path(DIR_LOG, "qc", "library_size")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "qc", "library_size", "qc_library_size.R"),
  cmdargs = c(
    "-i", file.path(DIR_DATA, "Processed_data", "03_hd_mm_scrnaseq_tcell_qc_seurat.qs"),
    "-b", "emulsion",
    "-d", "All_Tcells",
    "-o", DIR_RESULT,
    "-q", "Before_QC",
    "-s", "TRUE"
  ),
  stdout = file.path(DIR_LOG, "qc", "library_size", paste0("output_qc_library_size_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "qc", "library_size", paste0("error_qc_library_size_", log_time, ".log"))
)

```

```{r display_plots Lib_size, echo=FALSE, out.width = "800px"}

# Affichage automatique des plots

img_dir <- file.path(DIR_RESULT, "All_Tcells", "QC", "Library_Size", "Before_QC")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(knitr::include_graphics(img))

}

```

### Library diversity/heterogeneity (nFeatureRNA) {.tabset}

```{r QC_lib_div, echo=FALSE}

log_dir <- file.path(DIR_LOG, "qc", "library_diversity")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "qc", "library_diversity", "qc_library_diversity.R"),
  cmdargs = c(
    "-i", file.path(DIR_DATA, "Processed_data", "03_hd_mm_scrnaseq_tcell_qc_seurat.qs"),
    "-b", "emulsion",
    "-d", "All_Tcells",
    "-o", DIR_RESULT,
    "-q", "Before_QC",
    "-s", "TRUE"
  ),
  stdout = file.path(DIR_LOG, "qc", "library_diversity", paste0("output_qc_library_diversity_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "qc", "library_diversity", paste0("error_qc_library_diversity_", log_time, ".log"))
)

```

```{r display_plots Lib_div, echo=FALSE, out.width = "800px"}

# Affichage automatique des plots

img_dir <- file.path(DIR_RESULT, "All_Tcells", "QC", "Library_Diversity", "Before_QC")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(knitr::include_graphics(img))

}

```

### Librairy complexity {.tabset}

```{r QC_lib_complex, echo=FALSE}

log_dir <- file.path(DIR_LOG, "qc", "library_complexity")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "qc", "library_complexity", "qc_library_complexity.R"),
  cmdargs = c(
    "-i", file.path(DIR_DATA, "Processed_data", "03_hd_mm_scrnaseq_tcell_qc_seurat.qs"),
    "-b", "emulsion",
    "-d", "All_Tcells",
    "-o", DIR_RESULT,
    "-q", "Before_QC",
    "-s", "TRUE"
  ),
  stdout = file.path(DIR_LOG, "qc", "library_complexity", paste0("output_qc_library_complexity_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "qc", "library_complexity", paste0("error_qc_library_complexity_", log_time, ".log"))
)

```

```{r display_plots Lib_complex, echo=FALSE, out.width = "800px"}

# Affichage automatique des plots

img_dir <- file.path(DIR_RESULT, "All_Tcells", "QC", "Library_Complexity", "Before_QC")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(knitr::include_graphics(img))

}

```

### Cell Cycle Scoring {.tabset}

```{r QC_cell_cycle, echo=FALSE}

log_dir <- file.path(DIR_LOG, "qc", "cell_cycle")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "qc", "cell_cycle", "qc_cell_cycle.R"),
  cmdargs = c(
    "-i", file.path(DIR_DATA, "Processed_data", "03_hd_mm_scrnaseq_tcell_qc_seurat.qs"),
    "-b", "emulsion",
    "-d", "All_Tcells",
    "-o", DIR_RESULT,
    "-q", "Before_QC",
    "-s", "TRUE"
  ),
  stdout = file.path(DIR_LOG, "qc", "cell_cycle", paste0("output_qc_cell_cycle_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "qc", "cell_cycle", paste0("error_qc_cell_cycle_", log_time, ".log"))
)

```

```{r display_plots Cell cycle, echo=FALSE, out.width = "800px"}

# Affichage automatique des plots

img_dir <- file.path(DIR_RESULT, "All_Tcells", "QC", "Cell_Cycle", "Before_QC")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(knitr::include_graphics(img))

}

```

Later, see if clusters are enriched in specific cell cycle phases.

## Filtering {.tabset}

```{r Load pre-QC object for filtering}
# Load qs object
seu <- qread(file.path(DIR_DATA, "Processed_data", "03_hd_mm_scrnaseq_tcell_qc_seurat.qs"))
```


We have look at several QC metrics we will now set a cutoff for those to eliminate bad quality cells from our dataset. We have flag each cells for each QC metric based on the distribution of the data and the visualizations we have done. We have then create a new column in the metadata to indicate whether each cell passes or fails the QC criteria based on these cutoffs. Finally, we will subset the Seurat object to keep only the cells that pass QC for downstream analysis.

::: callout-tip
There are several way to set cutoff... we can do this the easy way… or the hard way ! It’s your choice.

-   Stringent QC will remove all cells that fail at least one QC metric (mito, ribo, libsize, libdiv, libcomplex)

-   Non stringent QC will remove only cells that fail all QC metrics (mito, ribo, libsize, libdiv, libcomplex)

    -   Often all cells pass QC so this is not really a good option

-   Moderate QC will remove cells that fail at least 2 (or 3) QC metrics (mito, ribo, libsize, libdiv, libcomplex) according what we thing it's the best compromise between stringency and cell retention.
:::

For Stringent QC :

```{r QC Cutoff Stringent QC}
seu@meta.data$Stringent_QC_status <- dplyr::case_when(
  # seu$DblFinder_class == "doublet" ~ "fail_doublet",
  (seu$mito_status == "high" | seu$libsize_status == "low" | seu$libdiv_status == "low" | seu$libcomplex_status == "low") ~ "fail",
  TRUE ~ "pass"
)

# Number of cells that fail QC
table(seu@meta.data$Stringent_QC_status)

# Add percentage of cells that fail QC for each emulsion and globally

```

On 87548 Tcells, 1737 cells fail the stringent QC (2% of cells).

For Non Stringent QC :

```{r QC Cutoff Non Stringent QC}
seu@meta.data$Non_stringent_QC_status <- dplyr::case_when(
  # seu$DblFinder_class == "doublet" ~ "fail_doublet",
  (seu$mito_status == "high" & seu$libsize_status == "low" & seu$libdiv_status == "low" & seu$libcomplex_status == "low") ~ "fail",
  TRUE ~ "pass"
)

# Number of cells that fail QC
table(seu@meta.data$Non_stringent_QC_status)
```

On 87548 Tcells, 87548 cells fail the non stringent QC (0% of cells).

::: callout-warning
This QC is not really good because it does not remove any cell from the dataset, even if some cells fail one or more QC metrics. It can be useful to keep all cells for some analysis, but it can also introduce noise and bias in the downstream analysis if we have a lot of low-quality cells in the dataset.
:::

I have chose the moderate QC for the downstream analysis.

For Moderate QC :

```{r QC Cutoff Moderate QC}
seu@meta.data$Moderate_QC_status <- dplyr::case_when(
  # seu$DblFinder_class == "doublet" ~ "fail_doublet",
  (seu$mito_status == "high" | (seu$ribo_status == 'high' & seu$libsize_status == "low" & seu$libdiv_status == "low" & seu$libcomplex_status == "low")) ~ "fail",
  TRUE ~ "pass"
)

# Number of cells that fail QC
table(seu@meta.data$Moderate_QC_status)
```

With Moderate QC we have 1570 cells that failed QC and the rest pass. Hence, we retain 85978 cells for downstream analysis, which represent ~98% of the original dataset. This QC is a good compromise between stringency and cell retention, as it removes cells that fail at least 2 QC metrics (mito high and libcomplex low) while keeping the majority of cells in the dataset.

We subset dataset to keep only cells that pass QC for downstream analysis.

```{r QC_cutoff, echo=FALSE}

log_dir <- file.path(DIR_LOG, "qc", "filter")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "qc", "qc_filter.R"),
  cmdargs = c(
    "-i", file.path(DIR_DATA, "Processed_data", "03_hd_mm_scrnaseq_tcell_qc_seurat.qs"),
    "-d", "CD4_Tcells",
    "-n", "Moderate_QC",
    "-r", "mito_status == 'high' | (ribo_status == 'high' & libsize_status == 'low' & libdiv_status == 'low' & libcomplex_status == 'low')",
    "-b", TRUE,
    "-s", TRUE
  ),
  stdout = file.path(DIR_LOG, "qc", "filter", paste0("output_qc_filter_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "qc", "filter", paste0("error_qc_filter_", log_time, ".log"))
)

```

------------------------------------------------------------------------

## Post-filtering QC {.tabset}

### Cells count {.tabset}

```{r QC Cell count after filtering}
log_dir <- file.path(DIR_LOG, "qc", "cell_count")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "qc", "cell_count", "qc_cell_count.R"),
  cmdargs = c(
    "-i", file.path(DIR_DATA, "Processed_data", "03_hd_mm_scrnaseq_tcell_qc_seurat.qs"),
    "-d", "All_Tcells",
    "-o", DIR_RESULT,
    "-q", "After_QC"
  ),
  stdout = file.path(DIR_LOG, "qc", "cell_count", paste0("output_qc_cell_count_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "qc", "cell_count", paste0("error_qc_cell_count_", log_time, ".log"))
)
```

```{r display_plots Cell count Viz after filtering, echo=FALSE, out.width = "800px"}
img_dir <- file.path(DIR_RESULT, "All_Tcells", "QC", "Cell_Counts", "After_QC")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(include_graphics(img, auto_pdf = TRUE, dpi = 150))

}
```

### Mitochondria {.tabset}

We will now look at the distribution of the percentage of mitochondrial genes expression to see if there are some cells with high mitochondrial content that could be removed.

```{r QC_mito after filtering, echo=FALSE}

log_dir <- file.path(DIR_LOG, "qc", "mitochondria")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "qc", "mitochondria", "qc_mitochondria.R"),
  cmdargs = c(
    "-i", file.path(DIR_DATA, "Processed_data", "03_hd_mm_scrnaseq_tcell_qc_seurat.qs"),
    "-b", "emulsion",
    "-d", "All_Tcells",
    "-o", DIR_RESULT,
    "-q", "After_QC",
    "-s", "TRUE"
  ),
  stdout = file.path(DIR_LOG, "qc", "mitochondria", paste0("output_qc_mitochondria_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "qc", "mitochondria", paste0("error_qc_mitochondria_", log_time, ".log"))
)

```

```{r display_plots 2 after filtering, echo=FALSE, out.width = "800px"}

# Affichage automatique des plots

library(knitr)

img_dir <- file.path(DIR_RESULT, "All_Tcells", "QC", "Mitochondria", "After_QC")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(knitr::include_graphics(img))

}

```

### Ribosomes {.tabset}

```{r QC_ribo after filtering, echo=FALSE}

log_dir <- file.path(DIR_LOG, "qc", "ribosome")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "qc", "ribosome", "qc_ribosome.R"),
  cmdargs = c(
    "-i", file.path(DIR_DATA, "Processed_data", "03_hd_mm_scrnaseq_tcell_qc_seurat.qs"),
    "-b", "emulsion",
    "-d", "All_Tcells",
    "-o", DIR_RESULT,
    "-q", "After_QC",
    "-s", "TRUE"
  ),
  stdout = file.path(DIR_LOG, "qc", "ribosome", paste0("output_qc_ribosome_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "qc", "ribosome", paste0("error_qc_ribosome_", log_time, ".log"))
)

```

```{r display_plots 3 after filtering, echo=FALSE, out.width = "800px"}

# Affichage automatique des plots

img_dir <- file.path(DIR_RESULT, "All_Tcells", "QC", "Ribosomal", "After_QC")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(knitr::include_graphics(img))

}

```

### Library size (nCountRNA) {.tabset}

```{r QC_lib_size after filtering, echo=FALSE}

log_dir <- file.path(DIR_LOG, "qc", "library_size")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "qc", "library_size", "qc_library_size.R"),
  cmdargs = c(
    "-i", file.path(DIR_DATA, "Processed_data", "03_hd_mm_scrnaseq_tcell_qc_seurat.qs"),
    "-b", "emulsion",
    "-d", "All_Tcells",
    "-o", DIR_RESULT,
    "-q", "After_QC",
    "-s", "TRUE"
  ),
  stdout = file.path(DIR_LOG, "qc", "library_size", paste0("output_qc_library_size_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "qc", "library_size", paste0("error_qc_library_size_", log_time, ".log"))
)

```

```{r display_plots Lib_size after filtering, echo=FALSE, out.width = "800px"}

# Affichage automatique des plots

img_dir <- file.path(DIR_RESULT, "All_Tcells", "QC", "Library_Size", "After_QC")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(knitr::include_graphics(img))

}

```

### Library diversity/heterogeneity (nFeatureRNA) {.tabset}

```{r QC_lib_div after filtering, echo=FALSE}

log_dir <- file.path(DIR_LOG, "qc", "library_diversity")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "qc", "library_diversity", "qc_library_diversity.R"),
  cmdargs = c(
    "-i", file.path(DIR_DATA, "Processed_data", "03_hd_mm_scrnaseq_tcell_qc_seurat.qs"),
    "-b", "emulsion",
    "-d", "All_Tcells",
    "-o", DIR_RESULT,
    "-q", "After_QC",
    "-s", "TRUE"
  ),
  stdout = file.path(DIR_LOG, "qc", "library_diversity", paste0("output_qc_library_diversity_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "qc", "library_diversity", paste0("error_qc_library_diversity_", log_time, ".log"))
)

```

```{r display_plots Lib_div after filtering, echo=FALSE, out.width = "800px"}

# Affichage automatique des plots

img_dir <- file.path(DIR_RESULT, "All_Tcells", "QC", "Library_Diversity", "After_QC")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(knitr::include_graphics(img))

}

```

### Librairy complexity {.tabset}

```{r QC_lib_complex after filtering, echo=FALSE}

log_dir <- file.path(DIR_LOG, "qc", "library_complexity")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "qc", "library_complexity", "qc_library_complexity.R"),
  cmdargs = c(
    "-i", file.path(DIR_DATA, "Processed_data", "03_hd_mm_scrnaseq_tcell_qc_seurat.qs"),
    "-b", "emulsion",
    "-d", "All_Tcells",
    "-o", DIR_RESULT,
    "-q", "After_QC",
    "-s", "TRUE"
  ),
  stdout = file.path(DIR_LOG, "qc", "library_complexity", paste0("output_qc_library_complexity_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "qc", "library_complexity", paste0("error_qc_library_complexity_", log_time, ".log"))
)

```

```{r display_plots Lib_complex after filtering, echo=FALSE, out.width = "800px"}

# Affichage automatique des plots

img_dir <- file.path(DIR_RESULT, "All_Tcells", "QC", "Library_Complexity", "After_QC")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(knitr::include_graphics(img))

}

```

### Cell Cycle Scoring {.tabset}

```{r QC_cell_cycle after filtering, echo=FALSE}

log_dir <- file.path(DIR_LOG, "qc", "cell_cycle")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "qc", "cell_cycle", "qc_cell_cycle.R"),
  cmdargs = c(
    "-i", file.path(DIR_DATA, "Processed_data", "03_hd_mm_scrnaseq_tcell_qc_seurat.qs"),
    "-b", "emulsion",
    "-d", "All_Tcells",
    "-o", DIR_RESULT,
    "-q", "After_QC",
    "-s", "TRUE"
  ),
  stdout = file.path(DIR_LOG, "qc", "cell_cycle", paste0("output_qc_cell_cycle_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "qc", "cell_cycle", paste0("error_qc_cell_cycle_", log_time, ".log"))
)

```

```{r display_plots Cell cycle after filtering, echo=FALSE, out.width = "800px"}

# Affichage automatique des plots

img_dir <- file.path(DIR_RESULT, "All_Tcells", "QC", "Cell_Cycle", "After_QC")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(knitr::include_graphics(img))

}

```


### Doublet detection {.tabset}

```{r QC_cell_cycle after filtering, echo=FALSE}

log_dir <- file.path(DIR_LOG, "qc", "doublet")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "qc", "doublet", "qc_doublet.R"),
  cmdargs = c(
    "-i", file.path(DIR_DATA, "Processed_data", "03_hd_mm_scrnaseq_tcell_qc_seurat.qs"),
    "-b", "emulsion",
    "-d", "All_Tcells",
    "-o", DIR_RESULT,
    "-s", "TRUE"
  ),
  stdout = file.path(DIR_LOG, "qc", "doublet", paste0("output_qc_cell_cycle_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "qc", "doublet", paste0("error_qc_cell_cycle_", log_time, ".log"))
)

```

```{r display_plots Cell cycle before filtering, echo=FALSE, out.width = "800px"}

# Affichage automatique des plots

img_dir <- file.path(DIR_RESULT, "All_Tcells", "QC", "Doublet", "Before_QC")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(knitr::include_graphics(img))

}

```

```{r display_plots Cell cycle after filtering, echo=FALSE, out.width = "800px"}

# Affichage automatique des plots
img_dir <- file.path(DIR_RESULT, "All_Tcells", "QC", "Doublet", "After_QC")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(knitr::include_graphics(img))

}

```

No doublets left in the dataset !


<br>

------------------------------------------------------------------------

------------------------------------------------------------------------

<br>


# Normalization {.tabset}

Now that we have a clean dataset of T cells, we can proceed to the normalization step.

::: callout-note
Normalization is a crucial step in single-cell RNA-seq data analysis, as it helps to correct for technical variations and allows for meaningful comparisons of gene expression levels across cells. We will perform normalization separately for the RNA and ADT matrices, as they represent different types of data (gene expression vs protein expression) and may require different normalization methods.
:::

## RNA matrix {.tabset}

For the RNA matrix, we will use the SCTransform method for normalization.

::: callout-note
SCTransform is a variance-stabilizing transformation that models gene expression using a regularized negative binomial regression. It accounts for technical noise and allows for more accurate identification of variable features, which can improve downstream analyses such as PCA, UMAP, and clustering. + We can regress some vars, like here mito percent per cells, it means that we remove the effect of this variable for calculation of normalisation factors and variance stabilization, which can help to reduce the influence of mitochondrial gene expression on downstream analyses, as it is often a source of technical variation in scRNA-seq data. By using SCTransform, we can ensure that our normalization is robust and that we are focusing on the most biologically relevant features for our analysis.

We also set the number of variable features to 3000, which is a common choice for scRNA-seq data and can be optimized based on the specific dataset and downstream analyses. By tuning this parameter, we can find the optimal number of variable features that capture the most informative genes for our analysis while avoiding noise from less informative genes.
:::

::: callout-warning
Optimize and tune according PCA, UMAP, clustering results the variable.feature arguments. Know how to justify 3000 variable features...
:::

```{r Normalize RNA}
seu_QC <- qread(file.path(DIR_DATA, "Processed_data", "03_hd_mm_scrnaseq_tcell_qc_seurat.qs"))
DefaultAssay(seu_QC) <- "RNA"

options(future.globals.maxSize = Inf)
future::plan("sequential")

seu_QC <- SCTransform(
  seu_QC,
  # vars.to.regress = c("subsets_mito_percent"),  # optionnal 
  variable.features.n = 3000, # To optimize and tune according PCA, UMAP, clustering results (see later)
  vst.flavor = "v2",
  return.only.var.genes = TRUE, # Limite la taille de l'objet Seurat en limitant la matrice scale.data uniquement aux gènes les plus variables (variable features) au lieu de tous les gènes, ce qui peut être très volumineux et consommer beaucoup de mémoire
  verbose = TRUE)

gc()
```

::: callout-tip
By regressing out the mitochondrial percentage, we can reduce the influence of mitochondrial gene expression on downstream analyses, which is often a source of technical variation in scRNA-seq data.
:::

::: callout-tip
Test ***different parameters*** (e.g., ***variable.features.n***) and evaluate their impact on downstream analyses (PCA, UMAP, clustering) to find the optimal settings for dataset.
:::

## ADT matrix {.tabset}

For the ADT matrix, we will use the CLR (Centered Log-Ratio) normalization method.

::: callout-note
It is commonly used for CITE-seq data. CLR normalization divides each count by the geometric mean of the values in that cell, then log-transforms the result. This method helps to account for the multiplicative nature of technical variations in ADT data, such as differences in antibody capture efficiency and background noise.
:::

```{r Normalize ADT}
DefaultAssay(seu_QC) <- "ADT"

seu_QC <- NormalizeData(
  seu_QC, 
  normalization.method = "CLR", # CLR = Centered Log-Ratio : standard for ADT (CITE-seq) data (Divide/Center each count by the geometric mean of the values in that cell, then log-transform)
  margin = 2, # Normalize across cells (so cols)
  assay = "ADT")

# Les counts ADT sont dominés par des rapports :
# efficacité de l’anticorps
# accessibilité de l’épitope
# capture / lavage
# bruit de fond
# 
# 👉 Si une cellule capte “2× plus d’anticorps”, toutes les protéines montent proportionnellement.
# 
# C’est un effet multiplicatif, pas une addition constante.
```

<br>

------------------------------------------------------------------------

------------------------------------------------------------------------

<br>


# Most Variable Features {.tabset}

::: callout-note
We don't need it here because SCTransform already identifies variable features during the normalization step and stores them in the "SCT" assay. However, if we were to use a different normalization method (e.g., NormalizeData), we would need to run FindVariableFeatures on the "RNA" assay to identify the most variable genes for downstream analyses.
:::

## RNA matrix {.tabset}

```{r MVP RNA}
# DefaultAssay(seu_QC) <- "RNA"
# seu_QC <- FindVariableFeatures(seu_QC, 
#                             assay = "RNA",
#                             selection.method = "vst",
#                             nfeatures = 2000)
```

::: callout-note
Not nécessary as we used SCTransform for normalization, which already identifies variable features during the process and stores them in the "SCT" assay. Otherwise, run NormalizeData and then FindVariableFeatures (with specific argument/options) on the "RNA" assay.
:::

::: callout-note
SCTTransform is based on Binomial Neg and perform Pearson residuals to identify variable features, which is more robust and accurate than traditional methods (e.g., vst) for scRNA-seq data. It accounts for technical noise and allows for better identification of biologically relevant features, leading to improved downstream analyses (PCA, UMAP, clustering). By using the variable features identified by SCTransform, we can ensure that we are focusing on the most informative genes for our analysis.
:::

## ADT matrix {.tabset}

```{r MVP ADT}
DefaultAssay(seu_QC) <- "ADT"
VariableFeatures(seu_QC) <- rownames(seu_QC[["ADT"]])
```

We take all the ADT features as variable features because we have only 13 ADT features, and we want to keep all of them for downstream analysis (clustering, annotation, etc.). All are biologically relevant and we don't want to exclude any of them.

<br>

------------------------------------------------------------------------

------------------------------------------------------------------------

<br>

# Scaling {.tabset}

## RNA matrix {.tabset}

```{r Scale RNA}
# DefaultAssay(seu_QC) <- "RNA"
# seu_QC <- ScaleData(
#   seu_QC,
#   assay = "RNA",
#   vars.to.regress = c("nCount_RNA", "subsets_mito_percent", "nFeature_RNA", "emulsion", "subsets_ribo_percent", "Phase"),
#   features = VariableFeatures(seu_QC))
```

::: callout-tip
We don't need to Scale data if we use SCTransform because it's already performed during the normalization step. It's still debate but run Scaling or not depends mostly on what we will see (OR NOT) later on PCA, UMAP, clustering results. If we see that the technical factors (e.g., nCount_RNA, subsets_mito_percent) are driving the separation of cells in PCA/UMAP, then we can try to regress them out during Scaling to reduce their influence on downstream analyses. Otherwise, if we see that the biological signal is strong and not confounded by technical factors, we can skip Scaling and directly run PCA/UMAP on the SCT assay.
:::

## ADT matrix {.tabset}

```{r Scale ADT}
DefaultAssay(seu_QC) <- "ADT"
seu_QC <- ScaleData(
  seu_QC,
  assay = "ADT")
```

<br>

------------------------------------------------------------------------

------------------------------------------------------------------------

<br>

# PCA {.tabset}

## RNA matrix {.tabset}

### Run

::: callout-tip
Run PCA on the "SCT" assay, which contains the normalized and scaled data from SCTransform.
:::

We will use the variable features identified by SCTransform for PCA, which are stored in the "SCT" assay. We can specify the number of principal components (npcs) to compute based on our dataset and downstream analysis needs. In this case, we will compute 50 PCs, but this can be adjusted based on the elbow plot and the amount of variance explained by the PCs.

```{r PCA RNA Run}
DefaultAssay(seu_QC) <- "SCT"
seu_QC <- RunPCA(
  seu_QC, 
  reduction.name = "pca",
  npcs = 50,
  weight.by.var = TRUE,
  verbose = TRUE)
```

### Visualize

```{r PCA RNA ElbowPlot}
p_pca_rna <- ElbowPlot(
  seu_QC, 
  ndims = 50, 
  reduction = "pca") +
  ggtitle("PCA Elbow Plot - RNA assay")
p_pca_rna

dir.create(
  file.path(DIR_RESULT,
            "All_Tcells",
            "PCA",
            "RNA"),
  recursive = TRUE,
  showWarnings = FALSE)

ggsave(
  plot = p_pca_rna,
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "PCA",
    "RNA",
    "RNA_PCA_elbowplot.png"),
  width = 12,
  height = 6)
```

11 PCs seems to be a good choice based on the elbow plot, but we can also look at the percentage of variance explained by each PC to make a more informed decision.

::: callout-tip
We can also test different numbers of PCs (e.g., 10, 15, 20) and evaluate their impact on downstream analyses (UMAP, clustering) to find the optimal number of PCs for our dataset.
:::

```{r PCA RNA Variance Explained}
# Percentage of variance explained by each PC
pct_var <- seu_QC[["pca"]]@stdev^2
pct_var <- pct_var / sum(pct_var) * 100 

cum_var <- cumsum(pct_var)

df_var <- data.frame(
  PC = 1:length(pct_var),
  Variance = pct_var,
  CumulativeVariance = cum_var
)

p_cumvar_rna <- ggplot(df_var, aes(x = PC)) +
  geom_bar(aes(y = Variance), stat = "identity", fill = "skyblue") +
  geom_line(aes(y = CumulativeVariance), color = "red", size = 1) +
  geom_point(aes(y = CumulativeVariance), color = "red", size = 2) +
  ylab("Variance Explained (%)") +
  xlab("Principal Components") +
  ggtitle("Explained variance per PC with cumulative variance") +
  theme_minimal() +
  scale_x_continuous(breaks = 1:length(pct_var)) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )
p_cumvar_rna

ggsave(
  plot = p_cumvar_rna,
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "PCA",
    "RNA",
    "RNA_Variance_explained_by_PCs.png"),
  width = 12,
  height = 6)
```

11 PC allow us to explained almost 75% of the variance in RNA matrix

```{r PCA RNA PC Heatmap}
#| fig-width: 10
#| fig-height: 12
#| fig-cap: "PCA Heatmap - RNA assay"
#| fig-note: "Heatmap of the top genes contributing to the first 16 PCs, which can help us understand the biological processes driving the variation in our dataset. We can look at the top genes for each PC and see if they are associated with specific pathways or cell types, which can provide insights into the underlying biology of our T cells."

DefaultAssay(seu_QC) <- "SCT"

p_pca_rna_heatmap <- DimHeatmap(
  seu_QC,
  reduction = "pca",
  dims = 1:11,
  cells = 500,
  balanced = TRUE,
  fast = TRUE
) + ggtitle("PCA Heatmap - RNA assay")
p_pca_rna_heatmap

ggsave(
  plot = p_pca_rna_heatmap,
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "PCA",
    "RNA",
    "RNA_PCA_heatmap.png"),
)
```

Heatmap of the top genes contributing to the first 11 PCs, which can help us understand the biological processes driving the variation in our dataset. We can look at the top genes for each PC and see if they are associated with specific pathways or cell types, which can provide insights into the underlying biology of our T cells.

```{r PCA DimPlot RNA emulsion}
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "PCA DimPlot by emulsion - RNA assay"
#| fig-alt: "PCA DimPlot colored by emulsion"
#| fig-note: "PCA DimPlot colored by emulsion, which can help us identify potential batch effects or technical variations associated with the emulsion. If we see that cells from different emulsions cluster separately in PCA space, it may indicate that there are technical differences between the emulsions that are driving the separation of cells. This can be important to consider when interpreting downstream analyses (e.g., clustering, differential expression) and may require batch correction methods (e.g., Harmony) to mitigate the influence of this technical variation."

p_pca_rna_dimplot_emulsion <- DimPlot(
  seu_QC,
  reduction = "pca",
  group.by = "emulsion") +
  scale_fill_manual(values = pal, drop = FALSE) +
  ggtitle("PCA DimPlot - RNA assay")
p_pca_rna_dimplot_emulsion

ggsave(
  plot = p_pca_rna_dimplot_emulsion,
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "PCA",
    "RNA",
    "RNA_PCA_DimPlot_by_emulsion.png"),
  width = 12,
  height = 6)
```
```{r CD4 PCA DimPlot RNA emulsion wrap}
#| fig-width: 7
#| fig-height: 10
#| fig-cap: "PCA DimPlot by emulsion - RNA assay"
#| fig-alt: "PCA DimPlot colored by emulsion"
#| fig-note: "PCA DimPlot colored by emulsion, which can help us identify potential batch effects or technical variations associated with the emulsion. If we see that cells from different emulsions cluster separately in PCA space, it may indicate that there are technical differences between the emulsions that are driving the separation of cells. This can be important to consider when interpreting downstream analyses (e.g., clustering, differential expression) and may require batch correction methods (e.g., Harmony) to mitigate the influence of this technical variation."

p_pca_rna_dimplot_emulsion_wrap <- DimPlot(
  seu_QC,
  reduction = "pca",
  group.by = "emulsion") +
  ggtitle("PCA DimPlot - RNA assay") + 
  scale_fill_manual(values = pal, drop = FALSE) +
  facet_wrap(~emulsion, ncol = 6) +
  NoLegend()
p_pca_rna_dimplot_emulsion_wrap

ggsave(
  plot = p_pca_rna_dimplot_emulsion_wrap,
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "PCA",
    "RNA",
    "RNA_PCA_DimPlot_by_emulsion_wrap.png"),
  width = 15,
  height = 10)
```

There is some separation of cells based on emulsion in PCA space, which suggests that there may be some batch effect or technical variation associated with the emulsion. This could be due to differences in sample processing, library preparation, or sequencing between emulsions. It's important to keep this in mind when interpreting downstream analyses (e.g., clustering, differential expression) and consider applying batch correction methods (e.g., Harmony) if necessary to mitigate the influence of this technical variation.

```{r PCA DimPlot RNA Phase}
p_pca_rna_dimplot_phase <- DimPlot(
  seu_QC,
  reduction = "pca",
  group.by = "Phase") +
  ggtitle("PCA DimPlot - RNA assay")
p_pca_rna_dimplot_phase

ggsave(
  plot = p_pca_rna_dimplot_phase,
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "PCA",
    "RNA",
    "RNA_PCA_DimPlot_by_Phase.png"),
  width = 12,
  height = 6)

p_pca_rna_dimplot_phase_wrap <- DimPlot(
  seu_QC,
  reduction = "pca",
  group.by = "Phase") +
  ggtitle("PCA DimPlot - RNA assay") + 
  facet_wrap(~Phase)
p_pca_rna_dimplot_phase_wrap

ggsave(
  plot = p_pca_rna_dimplot_phase_wrap,
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "PCA",
    "RNA",
    "RNA_PCA_DimPlot_by_Phase_wrap.png"),
  width = 12,
  height = 6)
```

No clear separation of cells based on cell cycle phase in PCA space, which is good because it means that the biological factors (e.g., cell cycle phase) are not driving the separation of cells in PCA space. This suggests that the normalization and scaling steps have successfully mitigated the influence of these factors on downstream analyses.

## ADT matrix {.tabset}

```{r PCA ADT Run}
DefaultAssay(seu_QC) <- "ADT"
seu_QC <- RunPCA(
  seu_QC, 
  reduction.name = "apca",
  npcs = 14,
  weight.by.var = TRUE,
  verbose = TRUE)
```

```{r PCA ADT ElbowPlot}
p_pca_adt <- ElbowPlot(
  seu_QC, 
  ndims = 50, 
  reduction = "apca") +
  ggtitle("PCA Elbow Plot - RNA assay")
p_pca_adt

dir.create(
  file.path(DIR_RESULT,
            "All_Tcells",
            "PCA",
            "ADT"),
  recursive = TRUE,
  showWarnings = FALSE)

ggsave(
  plot = p_pca_adt,
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "PCA",
    "ADT",
    "ADT_PCA_elbowplot.png"),
  width = 12,
  height = 6)
```

```{r PCA ADT Variance Explained}
# Percentage of variance explained by each PC
pct_var <- seu_QC[["apca"]]@stdev^2
pct_var <- pct_var / sum(pct_var) * 100 

cum_var <- cumsum(pct_var)


df_var <- data.frame(
  PC = 1:length(pct_var),
  Variance = pct_var,
  CumulativeVariance = cum_var
)

p_cumvar_adt <- ggplot(df_var, aes(x = PC)) +
  geom_bar(aes(y = Variance), stat = "identity", fill = "skyblue") +
  geom_line(aes(y = CumulativeVariance), color = "red", size = 1) +
  geom_point(aes(y = CumulativeVariance), color = "red", size = 2) +
  ylab("Variance Explained (%)") +
  xlab("Principal Components") +
  ggtitle("Explained variance per PC with cumulative variance") +
  theme_minimal() +
  scale_x_continuous(breaks = 1:length(pct_var)) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )
p_cumvar_adt

ggsave(
  plot = p_cumvar_adt,
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "PCA",
    "ADT",
    "ADT_Variance_explained_by_PCs.png"),
  width = 12,
  height = 6)
```

We take ALL !!! The 13 PCs for downstream analyses (UMAP, clustering) because we have only 13 ADT features, and we want to keep all of them for downstream analysis (clustering, annotation, etc.). All are biologically relevant and we don't want to exclude any of them.

```{r }
qsave(
  seu_QC,
  file.path(DIR_DATA, "Processed_data", "04_hd_mm_scrnaseq_tcell_normalized_seurat.qs")
)

seu_QC <- qread(file.path(DIR_DATA, "Processed_data", "04_hd_mm_scrnaseq_tcell_normalized_seurat.qs"))
```



<br>

------------------------------------------------------------------------

------------------------------------------------------------------------

<br>


# Batch correction {.tabset}

## Tuning {.tabset}

### Gridesearch {.tabset}

#### Harmony {.tabset}

```{r batch correction harmony, echo=FALSE}

log_dir <- file.path(DIR_LOG, "batch_correction", "harmony")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "batch_correction", "tool", "harmony_gridsearch.R"),
  cmdargs = c(
    "-i", file.path(DIR_DATA, "Processed_data", "04_hd_mm_scrnaseq_tcell_normalized_seurat.qs"),
    "-b", "emulsion",
    "-d", "All_Tcells",
    "-o", DIR_RESULT,
    "--theta", "1,2,3,4",
    "--sigma", "0.05,0.1,0.2",
    "--dims", 11
  ),
  stdout = file.path(DIR_LOG, "batch_correction", "harmony", paste0("output_batch_correction_harmony_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "batch_correction", "harmony", paste0("error_batch_correction_harmony_", log_time, ".log"))
)

```

### Compute metrics {.tabset}

```{r compute lisi, echo=FALSE}

log_dir <- file.path(DIR_LOG, "batch_correction", "lisi")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "batch_correction", "metric", "compute_lisi.R"),
  cmdargs = c(
    "-i", file.path(DIR_RESULT, "All_Tcells", "Integration", "Harmony_gridsearch"),
    "-b", "emulsion",
    "-d", "All_Tcells",
    "-o", DIR_RESULT,
    "-c", "condition"
  ),
  stdout = file.path(DIR_LOG, "batch_correction", "lisi", paste0("output_batch_correction_lisi_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "batch_correction", "lisi", paste0("error_batch_correction_lisi_", log_time, ".log"))
)

```

```{r display_plots compute lisi, echo=FALSE, out.width = "800px"}

# Affichage automatique des plots
img_dir <- file.path(DIR_RESULT, "All_Tcells", "Integration", "LISI")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(knitr::include_graphics(img))

}

```

```{r display_csv compute lisi, echo=FALSE}

csv_dir <- file.path(DIR_RESULT, "All_Tcells", "Integration", "LISI")

csv_files <- list.files(csv_dir, pattern = "\\.csv$", full.names = TRUE)

for (f in csv_files) {
  
  cat("##", basename(f), "\n\n")
  
  df <- read.csv(f)
  
  print(datatable(df))
  
  cat("\n\n")
}
```

```{r compute knn, echo=FALSE}

log_dir <- file.path(DIR_LOG, "batch_correction", "knn")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "batch_correction", "metric", "compute_knn.R"),
  cmdargs = c(
    "-i", file.path(DIR_RESULT, "All_Tcells", "Integration", "Harmony_gridsearch"),
    "-b", "emulsion",
    "-d", "All_Tcells",
    "-o", DIR_RESULT,
    "-l", "condition"
  ),
  stdout = file.path(DIR_LOG, "batch_correction", "knn", paste0("output_batch_correction_knn_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "batch_correction", "knn", paste0("error_batch_correction_knn_", log_time, ".log"))
)

```

```{r display_plots compute knn, echo=FALSE, out.width = "800px"}

# Affichage automatique des plots
img_dir <- file.path(DIR_RESULT, "All_Tcells", "Integration", "KNN_purity")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(knitr::include_graphics(img))

}

```

```{r display_csv compute knn, echo=FALSE}

csv_dir <- file.path(DIR_RESULT, "All_Tcells", "Integration", "KNN_purity")

csv_files <- list.files(csv_dir, pattern = "\\.csv$", full.names = TRUE)

for (f in csv_files) {
  
  cat("##", basename(f), "\n\n")
  
  df <- read.csv(f)
  
  print(datatable(df))
  
  cat("\n\n")
}
```

### Compare integration {.tabset}

```{r compare integration, echo=FALSE}

log_dir <- file.path(DIR_LOG, "batch_correction", "comparaison")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "batch_correction", "comparaison", "compare_integration.R"),
  cmdargs = c(
    "-i", file.path(DIR_RESULT, "All_Tcells", "Integration"),
    "-d", "All_Tcells",
    "-o", DIR_RESULT
  ),
  stdout = file.path(DIR_LOG, "batch_correction", "comparaison", paste0("output_batch_correction_comparaison_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "batch_correction", "comparaison", paste0("error_batch_correction_comparaison_", log_time, ".log"))
)

```

```{r display_plots compare integration, echo=FALSE, out.width = "800px"}

# Affichage automatique des plots
img_dir <- file.path(DIR_RESULT, "All_Tcells", "Integration", "Comparaison")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(knitr::include_graphics(img))

}

```

```{r display_csv compare integration, echo=FALSE}

csv_dir <- file.path(DIR_RESULT, "All_Tcells", "Integration", "Comparaison")

csv_files <- list.files(csv_dir, pattern = "\\.csv$", full.names = TRUE)

for (f in csv_files) {
  
  cat("##", basename(f), "\n\n")
  
  df <- read.csv(f)
  
  print(datatable(df))
  
  cat("\n\n")
}
```

::: callout-tip
Ideally, test different batch correction methods (e.g., Harmony, Seurat's integration, MNN) and evaluate their impact on downstream analyses (PCA, UMAP, clustering) to find the optimal method for dataset.
:::

If I want to test different batch correction methods (e.g., Harmony, Seurat's integration, MNN), I can run them in parallel and then compare their performance using metrics like LISI score and KNN purity. This will allow me to evaluate the effectiveness of each method in integrating the data while preserving biological variation, and ultimately choose the optimal method for my dataset.
For that I only have to rerun the first script here (not armony_gridsearch.R but the one with the method I want to test) and then run the scripts for computing LISI and KNN metrics, and finally the comparison script to visualize the results.

## Final correction {.tabset}

```{r final correction, echo=FALSE}

log_dir <- file.path(DIR_LOG, "batch_correction", "final")
dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)

callr::rscript(
  script = file.path(DIR_SCRIPT, "function", "batch_correction", "tool", "harmony_simple.R"),
  cmdargs = c(
    "-i", file.path(DIR_DATA, "Processed_data", "04_hd_mm_scrnaseq_tcell_normalized_seurat.qs"),
    "-d", "All_Tcells",
    "-b", "emulsion",
    "-D", 11,
    "-t", 1,
    "-S", 0.05,
    "-o", DIR_RESULT,
    "-s", file.path(DIR_DATA, "Processed_data", "05_hd_mm_scrnaseq_tcell_batch_corrected_seurat.qs")
  ),
  stdout = file.path(DIR_LOG, "batch_correction", "final", paste0("output_batch_correction_final_", log_time, ".log")),
  stderr = file.path(DIR_LOG, "batch_correction", "final", paste0("error_batch_correction_final_", log_time, ".log"))
)

```



```{r display_plots final batch correction, echo=FALSE, out.width = "800px"}

# Affichage automatique des plots
img_dir <- file.path(DIR_RESULT, "All_Tcells", "Integration", "Final", "Harmony")

img_files <- list.files(img_dir, pattern = "\\.png$", full.names = TRUE)

for (img in img_files) {

  print(knitr::include_graphics(img))

}

```


::: callout-tip
Test LISI score also BEFORE and AFTER batch correction to quantitatively evaluate the effectiveness of the batch correction method in integrating the data while preserving biological variation.
:::

To assess the batch effect presence, we can compute the LISI (Local Inverse Simpson's Index) score, which quantifies the degree of mixing between batches (emulsions) and conditions in the embedding space.

::: callout-note
The LISI score ranges from 1 to K, where K is the number of categories in the variable being evaluated. Higher values indicate better local mixing across categories, while values close to 1 indicate strong local separation.
:::


> emulsion quantifies batch mixing across the embedding: higher values indicate better batch correction.

> source & condition quantifies condition mixing: lower values indicate better preservation of biological separation between conditions.

::: callout-tip
A low batch effect is characterized by high emulsion_LISI and low condition_LISI & source_LISI, indicating that technical batch effects are not present while biological differences between HH and MM are present.
:::


This plot shows the convergence of the Harmony algorithm during batch correction. The y-axis represents the objective function value (or a similar metric) that the algorithm is trying to minimize, while the x-axis represents the iteration number. A decreasing trend in the objective function value indicates that the algorithm is successfully correcting for batch effects and converging towards a solution. If the plot shows a plateau or an increase in the objective function value, it may suggest that the algorithm is not converging properly, and you may need to adjust parameters (e.g., theta, lambda, sigma) or increase the number of iterations to achieve better batch correction.

> Theta is the diversity clustering penalty parameter, which controls the strength of batch correction. Higher values will result in stronger correction, while lower values will allow for more batch-specific variation to remain in the data. The optimal value of theta can depend on the specific dataset and the degree of batch effect present, so it may require some tuning to find the best balance between correcting for batch effects and preserving biological variation.

> Ridge regression penalty : Default lambda=1. Bigger values protect against over correction. If several covariates are specified, then lambda can also be a vector which needs to be equal length with the number of variables to be corrected. In this scenario, each covariate level group will be assigned the scalars specified by the user. If set to NULL, harmony will start lambda estimation mode to determine lambdas automatically and try to minimize overcorrection (Use with caution still in beta testing).

> Width of soft kmeans clusters : Default sigma=0.1. Sigma scales the distance from a cell to cluster centroids. Larger values of sigma result in cells assigned to more clusters. Smaller values of sigma make soft kmeans cluster approach hard clustering.


Convergence plot of the Harmony algorithm, which shows how the batch correction process is progressing over iterations. The plot typically displays the objective function value (or a similar metric) on the y-axis and the iteration number on the x-axis. A decreasing trend in the objective function value indicates that the algorithm is successfully correcting for batch effects and converging towards a solution. If the plot shows a plateau or an increase in the objective function value, it may suggest that the algorithm is not converging properly, and you may need to adjust parameters (e.g., theta, lambda, sigma) or increase the number of iterations to achieve better batch correction. Here we are good.

::: callout-note
We have the almost same number of PCs to take into account after Harmony correction, which is good because it means that the batch correction process has not significantly altered the structure of the data in PCA space. This suggests that the Harmony algorithm has successfully corrected for batch effects while preserving the underlying biological variation in our T cell dataset.
:::

We will take 11 PC post-harmony

We cannot see any separation of cells based on emulsion in PCA space after Harmony correction, which is good because it means that the batch effect associated with the emulsion has been successfully corrected. This suggests that the Harmony algorithm has effectively mitigated the influence of the emulsion on the separation of cells in PCA space, allowing us to focus on the underlying biological variation in our T cell dataset without confounding technical effects.

To assess the quality of our batch correction, we can compute the LISI (Local Inverse Simpson's Index) score, which quantifies the degree of mixing between batches (emulsions) and conditions in the embedding space.

::: callout-note
The LISI score ranges from 1 to K, where K is the number of categories in the variable being evaluated. Higher values indicate better local mixing across categories, while values close to 1 indicate strong local separation.
:::

> emulsion quantifies batch mixing across the embedding: higher values indicate better batch correction.

> source & condition quantifies condition mixing: lower values indicate better preservation of biological separation between conditions.

::: callout-tip
An optimal integration is therefore characterized by high emulsion_LISI and low condition_LISI & source_LISI, indicating that technical batch effects are removed while biological differences between HH and MM are preserved.
:::

We achieved effective correction of technical batch effects, as indicated by high emulsion LISI scores. However, we also observed increased mixing across biological variables such as condition and source. This suggests that Harmony may have partially overcorrected the data, potentially reducing biologically meaningful structure. This illustrates the inherent trade-off in batch correction between removing unwanted technical variation and preserving true biological heterogeneity, emphasizing the need for careful parameter tuning.


We can see that the emulsion_LISI score has significantly increased after Harmony correction, indicating improved batch mixing and successful correction of batch effects. However, the condition_LISI score has also increased, suggesting that some biological separation between conditions may have been lost during the batch correction process.


<br>

------------------------------------------------------------------------

------------------------------------------------------------------------

<br>


# Communities detection {.tabset}

Now that we have our PCA embeddings (RNA and ADT) and our batch-corrected embeddings (Harmony), we can proceed to the community detection step, which involves clustering the cells based on their similarity in the embedding space. This step is crucial for identifying distinct cell populations and understanding the heterogeneity within our T cell dataset. We will use the FindMultiModalNeighbors function to construct a shared nearest neighbor graph that integrates both RNA and ADT information, and then apply the FindClusters function to identify clusters of cells. We will also optimize the clustering parameters (e.g., k, resolution) to find the best clustering solution for our dataset.

## Tuning {.tabset}

Workflow and loop for k and res optimisation to find the best clustering solution for our T cell dataset. We will run this in parallel to speed up the process, and we will save the results (e.g., silhouette scores, cluster assignments) for each combination of parameters to evaluate their performance later on. This will allow us to systematically explore the parameter space and identify the optimal settings for our clustering analysis.

::: callout-note
IT IS VERY LONG !!! (15-20mins) TO RUN ALL THE COMBINATIONS OF PARAMETERS, SO IT'S COMMENTED
:::

```{r}
seu <- qread(file.path(DIR_DATA, "Processed_data", "05_hd_mm_scrnaseq_tcell_batch_corrected_seurat.qs"))
seu_path <- file.path(DIR_DATA, "Processed_data", "05_hd_mm_scrnaseq_tcell_batch_corrected_seurat.qs")

# --- Plan parallèle : multisession (workers isolés, Seurat-safe)
plan(multisession, workers = min(parallel::detectCores() - 1, 6))
plan(multisession, workers = 2)  # For testing with fewer workers


# ============================================================
# Main function : run clustering + silhouette + ARI bootstrap
# ============================================================

run_combo <- function(params, seu_path, dims_list, dims_use, base_dir, n_bootstrap = 3) {
  
  seu <- qread(seu_path)
  
  k          <- params$k
  resolution <- params$resolution
  
  message(sprintf("[k=%d | res=%.2f] Starting...", k, resolution))
  
  # --- 1. WNN + clustering ---
  seu_tmp <- 
    seu %>% 
    FindMultiModalNeighbors(
      reduction.list = list("harmony", "apca"),
      dims.list      = dims_list,
      k.nn           = k,
      verbose        = FALSE
    ) %>% 
    FindClusters(
      resolution = resolution,
      graph.name = "wsnn",
      algorithm  = 4,   # Leiden (More stable than Louvain)
      random.seed = 42, # For reproducibility
      verbose    = FALSE
    ) %>% 
    RunUMAP(
      nn.name   = "weighted.nn",
      reduction = "harmony",
      # dims      = dims_use,
      seed.use = 42,
      verbose   = FALSE
    )
  
  clust    <- as.integer(Idents(seu_tmp))
  n_clust  <- length(unique(clust))
  emb      <- Embeddings(seu_tmp, "harmony")[, dims_use]
  
  # --- 2. Approximate silhouette (sampling to avoid dist() on full dataset (very long)) ---
  n_cells  <- nrow(emb)
  n_sample <- min(n_cells, 3000)
  idx      <- sample(n_cells, n_sample)
  
  sil      <- silhouette(clust[idx], dist(emb[idx, ]))
  sil_df   <- data.frame(
    sil     = sil[, "sil_width"],
    cluster = factor(clust[idx])
  )
  
  mean_sil_global <- mean(sil_df$sil)
  prop_neg        <- mean(sil_df$sil < 0)
  
  mean_sil_per_cluster <- sil_df %>%
    group_by(cluster) %>%
    summarise(mean_sil = mean(sil), .groups = "drop")
  
  # --- 3. ARI stability by bootstrap ---
  ari_scores <- vapply(seq_len(n_bootstrap), function(i) {
    idx_boot  <- sample(n_cells, n_cells, replace = TRUE)
    seu_boot  <- seu[, idx_boot]
    seu_boot  <- FindMultiModalNeighbors(seu_boot,
                                         reduction.list = list("harmony", "apca"),
                                         dims.list = dims_list, k.nn = k, verbose = FALSE)
    seu_boot  <- FindClusters(seu_boot, resolution = resolution,
                              graph.name = "wsnn", algorithm = 4, verbose = FALSE)
    mclust::adjustedRandIndex(clust[idx_boot], as.integer(Idents(seu_boot)))
  }, numeric(1))
  
  mean_ari <- mean(ari_scores)
  
  # --- 4. Outputs ---
  n_colors   <- max(n_clust, 1)
  fill_colors <- if (n_clust <= length(pal)) pal[1:n_clust] else colors_55[1:n_clust]
  
  param_name <- sprintf("k%d_res%.2f_dims%d-%d", k, resolution, min(dims_use), max(dims_use))
  param_dir  <- file.path(base_dir, param_name)
  dir.create(param_dir, recursive = TRUE, showWarnings = FALSE)
  
  # Common title for all plots
  title_str <- sprintf("k=%d | res=%.2f | n_clust=%d\nmean_sil=%.3f | prop_neg=%.3f | ARI=%.3f",
                       k, resolution, n_clust, mean_sil_global, prop_neg, mean_ari)
  
  # 4a. Violin silhouette
  p_violin <- ggplot(sil_df, aes(x = cluster, y = sil, fill = cluster)) +
    geom_violin(trim = FALSE, alpha = 0.8) +
    geom_boxplot(width = 0.1, fill = "white", outlier.size = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
    scale_fill_manual(values = fill_colors) +
    theme_minimal(base_size = 11) +
    labs(title = title_str, x = "Cluster", y = "Silhouette width") +
    theme(legend.position = "none")
  
  # 4b. Histogram
  p_hist <- ggplot(sil_df, aes(x = sil, fill = cluster)) +
    geom_histogram(bins = 80, alpha = 0.7, position = "identity") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
    scale_fill_manual(values = fill_colors) +
    theme_minimal(base_size = 11) +
    labs(x = "Silhouette width", y = "Cell count", fill = "Cluster") +
    theme(legend.position = "none")
  
  # 4c. Mean silhouette per cluster
  p_mean <- ggplot(mean_sil_per_cluster, aes(x = cluster, y = mean_sil, fill = cluster)) +
    geom_col(alpha = 0.85) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5) +
    scale_fill_manual(values = fill_colors) +
    theme_minimal(base_size = 11) +
    labs(x = "Cluster", y = "Mean silhouette width") +
    theme(legend.position = "none")
  
  # 4d. UMAP silhouette (sur sous-échantillon)
  seu_tmp$silhouette <- NA_real_
  seu_tmp$silhouette[idx] <- sil[, "sil_width"]
  
  p_umap_sil <- FeaturePlot(
    seu_tmp, features = "silhouette",
    cols = c("blue", "white", "red"), order = TRUE
  ) + ggtitle(title_str)
  
  # 4e. UMAP density per cluster
  df_umap <- as.data.frame(Embeddings(seu_tmp, "umap"))
  colnames(df_umap) <- c("UMAP_1", "UMAP_2")
  df_umap$cluster   <- factor(clust)
  df_umap$condition <- seu_tmp$condition
  
  p_umap_density <- ggplot(df_umap, aes(UMAP_1, UMAP_2)) +
    geom_point(aes(color = cluster), size = 0.4, alpha = 0.5) +
    stat_density_2d(bins = 12, contour_var = "density",
                    color = "black", linewidth = 0.2, h = c(0.8, 0.8)) +
    scale_color_manual(values = fill_colors) +
    coord_equal() +
    theme_classic(base_size = 11) +
    labs(title = title_str, x = "UMAP 1", y = "UMAP 2", color = "Cluster") +
    guides(color = guide_legend(override.aes = list(size = 3)))
  
  # 4f. Clustree (all res tested for this object)
  p_clustree <- clustree(seu_tmp, prefix = "wsnn_res.")
  
  # 4g. Save combined (patchwork)
  p_combo <- (p_violin | p_hist) / (p_mean | p_umap_density)
  ggsave(file.path(param_dir, "summary_combo.png"),
         p_combo, width = 14, height = 10, dpi = 300)
  
  ggsave(file.path(param_dir, "UMAP_silhouette.png"),
         p_umap_sil, width = 8, height = 6, dpi = 300)
  
  # Silhouette classique
  png(file.path(param_dir, "silhouette_classic.png"),
      width = 2000, height = 1500, res = 300)
  plot(sil, border = NA)
  dev.off()
  
  message(sprintf("[k=%d | res=%.2f] Done. mean_sil=%.3f | ARI=%.3f", k, resolution, mean_sil_global, mean_ari))
  
  return(data.frame(
    k              = k,
    resolution     = resolution,
    n_clusters     = n_clust,
    mean_sil       = mean_sil_global,
    prop_negative  = prop_neg,
    mean_ari       = mean_ari,
    sd_ari         = sd(ari_scores)
  ))
}

# ============================================================
# Grid + lancement
# ============================================================

dims_list <- list(1:11, 1:13)
dims_use  <- 1:11

k_values   <- c(20, 30, 40)
res_values <- seq(0.2, 1.0, by = 0.2)
grid       <- expand.grid(k = k_values, resolution = res_values)

base_dir <- file.path(DIR_RESULT, "All_Tcells", "Clustering", "Tuning_k_res")
dir.create(base_dir, recursive = TRUE, showWarnings = FALSE)

all_results <- future_map_dfr(
  split(grid, seq(nrow(grid))),
  run_combo,
  seu      = seu_path,
  dims_list = dims_list,
  dims_use = dims_use,
  base_dir = base_dir,
  .options = furrr_options(seed = TRUE)
)

# ============================================================
# Summary global — heatmap k × resolution
# ============================================================

# Heatmap mean silhouette
p_heatmap_sil <- ggplot(all_results,
                        aes(x = factor(resolution), y = factor(k), fill = mean_sil)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(mean_sil, 3)), size = 3) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal(base_size = 11) +
  labs(title = "Mean silhouette — k × resolution",
       x = "Resolution", y = "k", fill = "Mean sil.")

# Heatmap ARI
p_heatmap_ari <- ggplot(all_results,
                        aes(x = factor(resolution), y = factor(k), fill = mean_ari)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(mean_ari, 3)), size = 3) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  theme_minimal(base_size = 11) +
  labs(title = "Mean ARI (bootstrap) — k × resolution",
       x = "Resolution", y = "k", fill = "ARI")

# Heatmap n_clusters
p_heatmap_n <- ggplot(all_results,
                      aes(x = factor(resolution), y = factor(k), fill = n_clusters)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n_clusters), size = 3) +
  scale_fill_viridis_c(option = "plasma") +
  theme_minimal(base_size = 11) +
  labs(title = "N clusters — k × resolution",
       x = "Resolution", y = "k", fill = "N clust.")

p_global <- p_heatmap_sil / p_heatmap_ari / p_heatmap_n

ggsave(file.path(base_dir, "GLOBAL_heatmap_summary.png"),
       p_global, width = 10, height = 14, dpi = 300)

write.csv(all_results,
          file.path(base_dir, "all_results_metrics.csv"),
          row.names = FALSE)


```








```{r Clusters stability optimization workflow}
plan(multicore, workers = parallel::detectCores() - 1)

run_and_save_parallel <- function(params, seu, dims_use, base_dir) {

  k <- params$k
  resolution <- params$resolution

  message("Running k=", k, " res=", resolution)

  seu_tmp <- seu

  # seu_tmp <- FindNeighbors(
  #   seu_tmp,
  #   reduction = "harmony",
  #   dims = dims_use,
  #   k.param = k,
  #   verbose = FALSE
  # )

  seu_tmp <- FindMultiModalNeighbors(seu_tmp,
                               reduction.list = list("harmony", "apca"),
                               dims.list = dims_list,
                               k.nn = k,
                               verbose = TRUE)

  seu_tmp <- FindClusters(
    seu_tmp,
    resolution = resolution,
    graph.name = "wsnn",
    verbose = FALSE
  )

  seu_tmp <- RunUMAP(
    seu_tmp,
    reduction = "harmony",
    dims = dims_use,
    verbose = FALSE
  )

  clust <- as.integer(Idents(seu_tmp))
  emb <- Embeddings(seu_tmp, "harmony")[, dims_use]

  sil <- silhouette(clust, dist(emb))

  sil_df <- data.frame(
    sil = sil[, "sil_width"],
    cluster = factor(clust)
  )

  mean_sil <- mean(sil_df$sil)
  prop_neg <- mean(sil_df$sil < 0)
  n_clusters <- length(unique(clust))

  # Choisir palette adaptée
  fill_colors <- if (n_clusters <= length(pal)) {
    pal[1:n_clusters]
  } else {
    colors_55[1:n_clusters]
  }

  param_name <- paste0("k", k,
                       "_res", resolution,
                       "_dims", min(dims_use), "-", max(dims_use))

  param_dir <- file.path(base_dir, param_name)
  dir.create(param_dir, showWarnings = FALSE)

  # # --- Clust tree ---
  # clustree(seu_tmp, prefix = "wsnn_res.")
  #
  # dir.create(
  #   file.path(DIR_RESULT,
  #             "CD4_Tcells",
  #             "Clustering"),
  #   recursive = TRUE,
  #   showWarnings = FALSE)
  #
  # ggsave(
  #   filename = file.path(
  #     param_dir,
  #     "Clustree_WNN.png"),
  #   width = 12,
  #   height = 12)

  # ---- Plot violin ----
  p1 <- ggplot(sil_df,
               aes(x = cluster, y = sil, fill = cluster)) +
    geom_violin(trim = FALSE) +
    geom_boxplot(width = 0.1, fill = "white") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    theme_minimal() +
    labs(title = paste0("k=", k,
                        " res=", resolution,
                        "\nmean_sil=", round(mean_sil, 3),
                        " neg=", round(prop_neg, 3)),
         x = "Cluster",
         y = "Silhouette width") +
  scale_fill_manual(values = fill_colors)

  ggsave(file.path(param_dir, "silhouette_violin.png"),
         p1, width = 8, height = 6, dpi = 300)

  # ---- Histogram ----
  all_silouhette <-
    ggplot(sil_df, aes(x = sil, fill = cluster)) +
      geom_histogram(bins = 100, alpha = 0.7, position = "identity") +
      geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
      theme_minimal() +
      labs(x = "Silhouette width", y = "Cell count", fill = "Cluster") +
      scale_fill_manual(values = fill_colors)
  all_silouhette

  ggsave(
    plot = all_silouhette,
    file.path(
      param_dir,
      "silhouette_hist.png"),
    width = 8,
    height = 6, dpi = 300)

  # --- Plot mean silhouette per cluster ----
  mean_sil <- sil_df %>%
  group_by(cluster) %>%
  summarise(mean_sil = mean(sil))

  mean_silouhette <- ggplot(mean_sil, aes(x = cluster, y = mean_sil, fill = cluster)) +
    geom_col() +
    geom_hline(yintercept = 0, linetype = "dashed") +
    theme_minimal() +
    labs(x = "Cluster", y = "Mean silhouette width") +
    scale_fill_manual(values = fill_colors)
  mean_silouhette

  ggsave(
    plot = mean_silouhette,
    filename =
      file.path(
        param_dir,
        "Silhouette_mean_width_per_cluster.png"),
    width = 12,
    height = 6)

  # --- Plot silhouette classic ----
  png(filename = file.path(
      param_dir,
      "Silhouette_classic_plot.png"),
      width = 2000,
      height = 1500,
      res = 300)

  classic_silouhette <- plot(sil, border = NA)
  classic_silouhette

  dev.off()

  # --- UMAP colored by silhouette width ----
  seu_tmp$silhouette <- sil[, "sil_width"]
  p_umap_silhouette <- FeaturePlot(
    seu_tmp,
    features = "silhouette",
    cols = c("blue", "white", "red"),
    order = TRUE
  ) + ggtitle(paste0("k=", k, " res=", resolution))

  ggsave(
    plot = p_umap_silhouette,
    filename = file.path(
      param_dir,
      "UMAP_silhouette.png"),
    width = 12,
    height = 6)

  # --- UMAP Density colored by cluster ----
  df <- as.data.frame(
    Embeddings(seu_tmp, "umap")
  )

  colnames(df) <- c("UMAP_1", "UMAP_2")

  # Ajouter metadata utile
  df$condition    <- seu_tmp$condition
  # df$BCL6_status  <- seu_tmp$BCL6_status
  df$seurat_cluster <- seu_tmp$seurat_clusters


  p_umap_density_clusters <- ggplot(df, aes(UMAP_1, UMAP_2)) +
    # Points colorés par cluster
    geom_point(
      aes(color = seurat_cluster),
      size = 0.5,
      alpha = 0.6
    ) +
    scale_color_manual(values = fill_colors) +

    # Contours en arrière-plan (structure globale)
    stat_density_2d(
      bins = 14,
      contour_var = "density",
      color = "black",
      linewidth = 0.2,
      h = c(0.8, 0.8)
    ) +

    coord_equal() +
    labs(title = "UMAP density plot of CD4 T cells",
         x = "UMAP 1",
         y = "UMAP 2",
         color = "Cluster") +

    theme_classic(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.text.y = element_text(angle = 45, hjust = 1),
      legend.title = element_text(size = 14, face = "bold"),    # taille + gras du titre
      legend.text  = element_text(size = 12)
    ) +

    # Agrandir les points dans la légende
    guides(color = guide_legend(
      override.aes = list(size = 4)  # taille des symboles dans la légende
    ))
  p_umap_density_clusters

  ggsave(
    plot = p_umap_density_clusters,
    filename = file.path(
      param_dir,
      "UMAP_density_clusters.png"),
    width = 12,
    height = 6)




  # --- Return results ----

  return(data.frame(
    k = k,
    resolution = resolution,
    n_clusters = n_clusters,
    mean_sil = mean_sil,
    prop_negative = prop_neg
  ))
}

dims_list <- list(1:13, 1:13)
dims_use <- 1:13

k_values <- c(20, 30, 40)
res_values <- seq(0.2, 1.0, by = 0.2)

grid <- expand.grid(k = k_values,
                    resolution = res_values)

base_dir <- file.path(DIR_RESULT, "CD4_Tcells", "Clustering", "Tuning_k_res")

dir.create(
  file.path(base_dir),
  recursive = TRUE,
  showWarnings = FALSE)

all_results <- future_map_dfr(
  split(grid, seq(nrow(grid))),
  run_and_save_parallel,
  seu = seu_CD4_QC_Moderate,
  dims_use = dims_use,
  base_dir = base_dir,
  .options = furrr_options(seed = TRUE)
)

```






# Weighted Nearest Neighbor (WNN) graph {.tabset}

```{r WNN}
seu <- FindMultiModalNeighbors(seu,
                               reduction.list = list("harmony", "apca"),
                               dims.list = list(1:9, 1:13))
```

# UMAP {.tabset}

## Tuning {.tabset}

```{r Tuning UMAP}
# Dir  to save
dir.create(
  file.path(DIR_RESULT,
            "All_Tcells",
            "UMAP",
            "Tuning"),
  recursive = TRUE,
  showWarnings = FALSE)

params <- expand.grid(
  n.neighbors = c(40),
  min.dist    = c(0.04, 0.05, 0.06, 0.07, 0.08, 0.09),
  spread      = c(1.5, 2)
)

for (i in seq_len(nrow(params))) {

  md <- params$min.dist[i]
  sp <- params$spread[i]
  nn <- params$n.neighbors[i]

  red_name <- paste0("umap_md", md, "_sp", sp, "_nn", nn)

  seu <- RunUMAP(
    seu,
    nn.name     = "weighted.nn",
    min.dist    = md,
    spread      = sp,
    n.neighbors = nn,
    seed.use    = 42,
    reduction.name = red_name
  )

  p <- DimPlot(
    seu,
    reduction = red_name,
    group.by  = "predicted.celltype.l2") +
    ggtitle(paste("min.dist =", md,
                  "spread =", sp,
                  "n.neighbors =", nn)) +
    scale_color_manual(values = pal)

  ggsave(
    plot = p,
    filename = file.path(
      DIR_RESULT,
      "All_Tcells",
      "UMAP",
      "Tuning",
      paste0(red_name, ".png")
    ),
    width = 12,
    height = 6
  )
}

```

## Run {.tabset}

```{r Run UMAP}
seu <- RunUMAP(seu,
               nn.name = "weighted.nn",
               min.dist = 0.06,
               spread = 1.0,
               n.neighbors = 50,
               seed.use = 42)
```

## Visualization {.tabset}

```{r Viz UMAP}
# Visualization
metadata <- c("condition", "source", "id", "predicted.celltype.l1", "predicted.celltype.l2", "predicted.celltype.l3")


dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "UMAP",
    "Metadata"),
  recursive = TRUE,
  showWarnings = FALSE)

for (meta in metadata) {
  p <- DimPlot(seu, reduction = "umap", group.by = meta) + ggtitle(meta)
  
  print(p)
  
  # Nom de fichier
  file_name <- paste0(output_dir, "/UMAP_metadata_", meta, ".png")

  ggsave(file_name, plot = p, width = 6, height = 5)
  
  ggsave(
    filename = file.path(
      DIR_RESULT,
      "All_Tcells",
      "UMAP",
      "Metadata",
      paste0("UMAP_metadata_", meta, ".pdf")
    ),
    plot = p,
    width = 6,
    height = 5
  )
}
```

# Clustering {.tabset}

## Test several clustering resolutions {.tabset}

```{r FindClusters}
# # Clustering (commun RNA + ADT on WNN graph)
# Explore different resolutions
resolutions <- seq(0.01, 0.1, by = 0.01)
for (res in resolutions) {
  seu <- FindClusters(seu, graph.name = "wsnn", resolution = res, verbose = FALSE)
}
```

## Visualization {.tabset}

```{r Viz clusters (Clustree)}
clustree(seu, prefix = "wsnn_res.")

dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering"),
  recursive = TRUE,
  showWarnings = FALSE)

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering",
    "Clustree_WNN.pdf"),
  width = 10,
  height = 8
)
```

## Optimal resolution {.tabset}

```{r Choose opti res}
# Choose resolution 0.06
seu$seurat_clusters <- seu$wsnn_res.0.1

# Reorganize clusters names
seu$seurat_clusters <- factor(seu$seurat_clusters, 
                              levels = names(sort(table(seu$seurat_clusters), 
                                                  decreasing = TRUE)))
```

## Clusters on UMAP {.tabset}

```{r Clusters on UMAP}
DimPlot(seu, reduction = "umap", group.by = "seurat_clusters") + ggtitle("seurat_clusters")

# Save the plot
ggsave(
  plot = last_plot(),
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering",
    "UMAP_seurat_clusters.pdf"),
  width = 12,
  height = 6)

ggsave(
  plot = last_plot(),
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "UMAP",
    "Metadata",
    "UMAP_seurat_clusters.pdf"),
  width = 12,
  height = 6)

DimPlot(seu, reduction = "umap", group.by = "seurat_clusters") + ggtitle("seurat clusters") +
  facet_wrap(~seurat_clusters)

ggsave(
     plot = last_plot(),
     filename = file.path(
       DIR_RESULT,
       "All_Tcells",
       "Clustering",
       "UMAP_seurat_clusters_facet.pdf"),
     width = 12,
     height = 6)

ggsave(
  plot = last_plot(),
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "UMAP",
    "Metadata",
    "UMAP_seurat_clusters_facet.pdf"),
  width = 12,
  height = 6)
```

#### UMAP density per cluster {.tabset}

```{r UMAP density per cluster}
### UMAP density

df <- as.data.frame(
  Embeddings(seu_CD8_QC_Moderate, "umap")
)

colnames(df) <- c("UMAP_1", "UMAP_2")

# Ajouter metadata utile
df$condition    <- seu_CD8_QC_Moderate$condition
df$BCL6_status  <- seu_CD8_QC_Moderate$BCL6_status
df$seurat_cluster <- seu_CD8_QC_Moderate$seurat_clusters


p_umap_density_clusters <- ggplot(df, aes(UMAP_1, UMAP_2)) +
  # Points colorés par cluster
  geom_point(
    aes(color = seurat_cluster),
    size = 0.5,
    alpha = 0.6
  ) +
  scale_colour_manual(values = pal) +
  
  # Contours en arrière-plan (structure globale)
  stat_density_2d(
    bins = 14,
    contour_var = "density",
    color = "black",
    linewidth = 0.2,
    h = c(0.8, 0.8)
  ) +
  
  coord_equal() +
  labs(title = "UMAP density plot of CD8 T cells",
       x = "UMAP 1",
       y = "UMAP 2",
       color = "Cluster") +
  
  theme_classic(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 14, face = "bold"),    # taille + gras du titre
    legend.text  = element_text(size = 12)
  ) +
  
  # Agrandir les points dans la légende
  guides(color = guide_legend(
    override.aes = list(size = 4)  # taille des symboles dans la légende
  ))
p_umap_density_clusters

ggsave(
  plot = p_umap_density_clusters,
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering",
    "UMAP_seurat_clusters_density.png"),
  width = 12,
  height = 6)

ggsave(
  plot = p_umap_density_clusters,
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "UMAP",
    "Metadata",
    "UMAP_seurat_clusters_density.pdf"),
  width = 12,
  height = 6)
```

## Clusters quality {.tabset}

### QC metrics per cluster {.tabset}

```{r QC metrics per clusters}
VlnPlot(
  seu,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  group.by = "seurat_clusters",
  pt.size = 0,
  ncol = 3
)

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering",
    "Clusters_QC_metrics.pdf"),
  width = 12,
  height = 6)
```

nCount_RNA: Library size

nFeature_RNA: Library complexity

percent.mt: Mitochondrial content

### Libraries size per cluster {.tabset}

```{r Libsize per clusters}
# Plot number of cells per cluster with percentage labels of total cells
cell_counts <- seu@meta.data %>%
  count(seurat_clusters, name = "Cell_Count") %>%
  mutate(
    Percentage = round(100 * Cell_Count / sum(Cell_Count), 6),
    label = paste0(Percentage, "%")
  )


ggplot(cell_counts, aes(y = seurat_clusters, x = Cell_Count, fill = seurat_clusters)) +
  geom_col() +
  geom_text(aes(label = label), hjust = -0.1, size = 3) +
  theme_minimal() +
  labs(
    title = "Number (and %) of cells per cluster",
    x = "Number of cells",
    y = "Cluster"
  ) +
  theme(legend.position = "none")

# Save the plot
ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering",
    "Clusters_cell_counts.pdf"),
  width = 8,
  height = 6
)
```

## Cluster repartition per patients {.tabset}

```{r Clusters repartition per patients}
# Proportion of cells per cluster for each patient
cluster_patient <- seu@meta.data %>%
  group_by(id, seurat_clusters) %>%
  summarise(Cell_Count = n()) %>%
  group_by(id) %>%
  mutate(Percentage = Cell_Count / sum(Cell_Count) * 100)

ggplot(cluster_patient, aes(x = id, y = Percentage, fill = seurat_clusters)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Proportion of cells per cluster for each patient",
    x = "Patient ID",
    y = "Proportion of cells",
    fill = "Cluster"
  ) + 
  theme(
      axis.text.x = element_text(angle = 45, 
                                 hjust = 1, 
                                 vjust = 1))

# Save the plot
ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering",
    "Clusters_repartition_per_patient.pdf"),
  width = 12,
  height = 6
)
```

### Mitochondrial content per cluster {.tabset}

```{r Mito per clusters}
stats_mt <- seu@meta.data %>%
  group_by(seurat_clusters) %>%
  summarise(
    mean_pct_mt   = mean(percent.mt),
    median_pct_mt = median(percent.mt),
    sd_pct_mt     = sd(percent.mt)
  )

kable(stats_mt, format = "html") %>%
  kable_styling()

# Save the table
write.csv(stats_mt, 
          file = file.path(
            DIR_RESULT,
            "All_Tcells",
            "Clustering",
            "cluster_mitochondrial_content_stats.csv"),
          row.names = FALSE)
```

```{r Viz Mito per clusters}
ggplot(seu@meta.data,
       aes(
         x = percent.mt,
         y = seurat_clusters,
         fill = seurat_clusters
       )) +
  geom_density_ridges(
    alpha = 0.7,
    scale = 1.2,
    rel_min_height = 0.01,
    color = "white") +
  # theme_ridges() +
  # theme(
  #   legend.position = "none"
  # ) +
  # geom_vline(xintercept = 20, linetype = "dashed", color = "red") +
  labs(
    x = "% mitochondrial genes",
    y = "Clusters")

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering",
    "Clusters_mitochondrial_content_ridges.pdf"),
  width = 12,
  height = 6
)
```

Too few cells in clusters 10 to be seen

<!-- Mitcochondrial content on UMAP -->

<!-- ```{r} -->

<!-- FeaturePlot( -->

<!--   seu, -->

<!--   features = "percent.mt", -->

<!--   cols = c("lightgrey", "red"), -->

<!--   reduction = "umap") -->

<!-- ``` -->

## Remove bad clusters {.tabset}

```{r Remove bad clusters}
# Remove clusters 
seu <- subset(seu, 
              subset = !(seurat_clusters %in% c("8", "9","10")))

seu$seurat_clusters <- factor(seu$seurat_clusters)

```

```{r UMAP without bad clusters}
DimPlot(seu, reduction = "umap", group.by = "seurat_clusters") + ggtitle("seurat clusters after removing bad clusters")

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "UMAP",
    "Metadata",
    "UMAP_seurat_clusters_after_removing_bad_clusters.pdf"),
  width = 12,
  height = 6
) 

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering",
    "UMAP_seurat_clusters_after_removing_bad_clusters.pdf"),
  width = 12,
  height = 6
)

DimPlot(seu, reduction = "umap", group.by = "seurat_clusters") + ggtitle("seurat clusters after removing bad clusters") +
  facet_wrap(~seurat_clusters, ncol = (length(unique(seu$seurat_clusters))/2) )

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "UMAP",
    "Metadata",
    "UMAP_seurat_clusters_after_removing_bad_clusters_facet.pdf"),
  width = 12,
  height = 6
)

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering",
    "UMAP_seurat_clusters_after_removing_bad_clusters_facet.pdf"),
  width = 12,
  height = 6)
```

# Annotate clusters {.tabset}

## FeaturePlot {.tabset}

```{r FeaturePlot ADT}
dict <- list(
  "CD16-TotalSeqB" = "FCGR3A",
  "CD335-NKp46-TotalSeqB" = "NCR1",
  "CD197-CCR7-TotalSeqB" = "CCR7",
  "CD4-TotalSeqB" = "CD4",
  "CD8a-TotalSeqB" = "CD8A",
  "CD45RA-TotalSeqB" = "PTPRC",
  "CD62L-TotalSeqB" = "SELL",
  "CD127-IL-7Ra-TotalSeqB" = "IL7R",
  "CD25-TotalSeqB" = "IL2RA",
  "TIGIT-VSTM3-TotalSeqB" = "TIGIT",
  "CD226-DNAM-1-TotalSeqB" = "CD226",
  "CD38-TotalSeqB" = "CD38"
)

dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "FeaturePlot",
    "Prot_RNA_Clusters_FeaturePlot"),
  recursive = TRUE,
  showWarnings = FALSE
)

for (prot in names(dict)) {
  gene <- dict[[prot]]
  
  # ADT plot
  DefaultAssay(seu) <- "ADT"
  p1 <- FeaturePlot(seu, features = prot, reduction = "umap") + ggtitle(paste0(prot, " (ADT)"))
  
  # RNA plot
  DefaultAssay(seu) <- "RNA"
  # Vérifie que le gène existe dans RNA
  if (gene %in% rownames(seu[["RNA"]])) {
    p2 <- FeaturePlot(seu, features = gene, reduction = "umap") + ggtitle(paste0(gene, " (RNA)"))
  } else {
    p2 <- ggplot() + 
      ggtitle(paste0(gene, " (RNA) - Not found")) + 
      theme_void()
  }
  
  p3 <- DimPlot(seu, reduction = "umap", group.by = "seurat_clusters") + ggtitle("seurat_clusters")
  
  # Combine plots
  p <- (p1 | p2) / p3
  p


  
  # Sauvegarde
  ggsave(
    filename = file.path(
      DIR_RESULT,
      "All_Tcells",
      "FeaturePlot",
      "Prot_RNA_Clusters_FeaturePlot",
      paste0("Prot_RNA_Clusters_FeaturePlot_", prot, "_", gene, "_featureplot.png")),
    plot = p,
    width = 12,
    height = 12)
}
```

# Clusters markers {.tabset}

## Find all markers {.tabset}

```{r FindAllMarkers}
markers <- FindAllMarkers(seu, only.pos = TRUE, min.pct = 0.25)

markers_filt <- markers %>%
  filter(
    p_val_adj < 0.05,
    avg_log2FC > 0.5,
    !grepl("^MT-|^RPL|^RPS", gene)
  )
```

## Top 10 markers per cluster {.tabset}

```{r Top10 markers}
genes.scaled <- rownames(seu[["RNA"]]@scale.data)

top10 <- markers_filt %>%
  filter(gene %in% genes.scaled) %>%
  group_by(cluster) %>%
  slice_max(avg_log2FC, n = 10)

top10 %>% arrange(cluster, desc(avg_log2FC))
```

##### FeaturePlot of top10 markers per cluster {.tabset}

```{r All FeaturePlot Top10}
dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "FeaturePlot",
    "Top10_Markers_Per_Cluster"),
  recursive = TRUE,
  showWarnings = FALSE)

for (i in unique(top10$cluster)) {
  
  dir.create(
    file.path(
      DIR_RESULT,
      "All_Tcells",
      "FeaturePlot",
      "Top10_Markers_Per_Cluster",
      paste0("Cluster_", i)),
    recursive = TRUE,
    showWarnings = FALSE
  )

  genes_cluster <- top10 %>%
    filter(cluster == i) %>%
    pull(gene)

  p_list <- lapply(genes_cluster, function(gene) {
    FeaturePlot(
      seu,
      features = gene, 
      reduction = "umap") +
      # scale_color_gradient(low = "lightgrey", high = "red") +
      ggtitle(paste0("Cluster ", i, ": ", gene))
  })

  p_combined <- wrap_plots(p_list, ncol = 5)
  p_combined
  
  ggsave(
    filename = file.path(
      DIR_RESULT,
      "All_Tcells",
      "FeaturePlot",
      "Top10_Markers_Per_Cluster",
      paste0("Cluster_", i),
      paste0("Cluster_", i, "_Top10_Markers_featureplot.png")),
    plot = p_combined,
    width = 20,
    height = 12)
  
  lapply(genes_cluster, function(gene) {
    
    dir.create(
      file.path(
        DIR_RESULT,
        "All_Tcells",
        "FeaturePlot",
        "Top10_Markers_Per_Cluster",
        paste0("Cluster_", i),
        paste0("Gene_", gene)),
      recursive = TRUE,
      showWarnings = FALSE)
      
    violin <- VlnPlot(
      seu, 
      cols = pal,
      features = gene, 
      group.by = "seurat_clusters", 
      pt.size = 0) +
      ggtitle(paste0("Cluster ", i, ": ", gene))
    violin
      
    ggsave(
      plot = violin,
      filename = file.path(
        DIR_RESULT,
        "All_Tcells",
        "FeaturePlot",
        "Top10_Markers_Per_Cluster",
        paste0("Cluster_", i),
        paste0("Gene_", gene),
        paste0("Cluster_", i, "_", gene, "_VlnPlot.png")
      ),
      width = 6,
      height = 4)
      
    feature <- FeaturePlot(
      seu, 
      features = gene, 
      reduction = "umap") +
      ggtitle(paste0("Cluster ", i, ": ", gene))
    feature
    
    ggsave(
      filename = file.path(
        DIR_RESULT,
        "All_Tcells",
        "FeaturePlot",
        "Top10_Markers_Per_Cluster",
        paste0("Cluster_", i),
        paste0("Gene_", gene),
        paste0("Cluster_", i, "_", gene, "_FeaturePlot.png")
      ),
      plot = feature,
      width = 6,
      height = 5)
      
    p_cluster <- DimPlot(
      seu, 
      reduction = "umap", 
      group.by = "seurat_clusters") +
      scale_color_manual(values = pal)
    p_cluster
      
    p_violin_feature_cluster <- (violin | feature) / p_cluster
    p_violin_feature_cluster
      
    ggsave(
      filename = file.path(
        DIR_RESULT,
        "All_Tcells",
        "FeaturePlot",
        "Top10_Markers_Per_Cluster",
        paste0("Cluster_", i),
        paste0("Gene_", gene),
        paste0("Cluster_", i, "_", gene, "_Vln_Feature_Cluster.png")
      ),
      plot = p_violin_feature_cluster,
      width = 12,
      height = 10)
  })
}
```

## Heatmap of markers per cluster {.tabset}

```{r Heatmap Top10}
genes <- unique(top10$gene)

# Extract scaled data for the selected genes
mat <- GetAssayData(
  seu,
  assay = "RNA",
  layer = "scale.data"
)[genes, ]

# Order cells by cluster
clusters <- seu$seurat_clusters
ord <- order(clusters)

mat <- mat[, ord]
clusters <- clusters[ord]

# Create annotation for clusters
cluster_ha <- HeatmapAnnotation(
  Cluster = clusters,
  col = list(
    Cluster = structure(
      Seurat::DiscretePalette(length(unique(clusters)), "polychrome"),
      names = levels(clusters)
    )
  )
)
# Create heatmap
heatmap <- Heatmap(
  mat,
  name = "Expression",
  top_annotation = cluster_ha,
  show_column_names = FALSE,
  show_row_names = TRUE,
  cluster_columns = FALSE,
  cluster_rows = TRUE,
  row_names_gp = grid::gpar(fontsize = 8),
  column_title = "Top 10 markers per cluster"
  )


# Save heatmap
dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "Heatmap"),
  recursive = TRUE,
  showWarnings = FALSE)
  
draw(heatmap)
png(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Heatmap",
    "Top10_Markers_Per_Cluster_Heatmap.png"),
  width = 1200,
  height = 800
)



```

# Automatic annotation (SingleR) {.tabset}

## Make automatic predictions {.tabset}

```{r Run auto annot (SingleR)}
# Prepare SingleR input
sce <- as.SingleCellExperiment(seu, assay = "RNA")

# Reference dataset
ref <- fetchReference("monaco_immune", "2024-02-26")

# Run SingleR
predictions <- SingleR(
  test = sce,
  ref = ref,
  labels = ref$label.fine,
  assay.type.test = "logcounts")

# Add predictions to Seurat object
seu$predicted.celltype.l3_jordan <- predictions$pruned.labels
```

### Visualization of automatic annotations {.tabset}

```{r Viz auto Annot}

dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "Annotation",
    "Automatic"),
  recursive = TRUE,
  showWarnings = FALSE)

p1 <- DimPlot(seu, reduction = "umap", group.by = "predicted.celltype.l3_jordan") + ggtitle("predicted.celltype.l1")
p1
ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Annotation",
    "Automatic",
    "UMAP_predicted_celltype_JD.png"),
  plot = p1,
  width = 12,
  height = 6
)

p2 <- DimPlot(seu, reduction = "umap", group.by = "predicted.celltype.l3_jordan") + ggtitle("predicted.celltype.l3_jordan") + 
  facet_wrap(~predicted.celltype.l3_jordan, ncol = 6)
p2
ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Annotation",
    "Automatic",
    "UMAP_predicted_celltype_JD_facet.png"),
  plot = p2,
  width = 12,
  height = 12
)

p3 <- p1 | p2
p3
ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Annotation",
    "Automatic",
    "UMAP_predicted_celltype_JD_combined.pdf"),
  plot = p3,
  width = 12,
  height = 6
)
```

# Manual annotation {.tabset}

## Create my celltype signatures {.tabset}

### Run presto to find markers {.tabset}

```{r Run presto}
expr_matrix <- as.matrix(seu@assays$RNA@data)
labels <- seu$seurat_clusters

presto_markers <- wilcoxauc(expr_matrix, labels)
```

### Construction of the signature table {.tabset}

```{r Construct signature tibble}
signature_tbl <- tribble(
  ~feature, ~cell_type, ~class,
  
  # Activated
  "CD69", "Activated", "State",
  "TNFRSF9", "Activated", "State",
  "IL2RA", "Activated", "State",

  # Apoptotic / stress
  "MALAT1", "Apoptotic", "State",
  "NEAT1", "Apoptotic", "State",

  # CD4 lineage
  "CD3D", "CD4", "Identity",
  "CD3E", "CD4", "Identity",
  "CD4",  "CD4", "Identity",
  "TRBC1","CD4", "Identity",

  # CD8 lineage
  "FCGR3A","CD8", "Identity",
  "CD3D", "CD8", "Identity",
  "CD3E", "CD8", "Identity",
  "CD8A", "CD8", "Identity",
  "CD8B", "CD8", "Identity",
  "TRBC1","CD8", "Identity",
  "NCR1", "CD8", "Identity",

  # Cytotoxic
  "PRF1","Cytotoxic", "Function",
  "GZMB","Cytotoxic", "Function",
  "GNLY","Cytotoxic", "Function",
  "GZMH","Cytotoxic", "Function",
  "CCL5","Cytotoxic", "Function",
  "NKG7","Cytotoxic", "Function",
  "KLRG1","Cytotoxic", "Function",
  "GZMA","Cytotoxic", "Function",

  # Exhausted
  "PDCD1","Exhausted", "State",
  "LAG3","Exhausted", "State",
  "TOX","Exhausted", "State",
  "TIGIT","Exhausted", "State",
  "CD226","Exhausted", "State",
  "GZMK","Exhausted", "State",
  "HAVCR2","Exhausted", "State",
  "XCL1","Exhausted", "State",
  "XCL2","Exhausted", "State",
  "CD160","Exhausted", "State",

  # HLA
  "HLA-DRA","HLA", "Function",
  "HLA-DRB1","HLA", "Function",

  # IFN / ISG
  "IFNG","IFN", "Function",
  "IFIT2","IFN", "Function",
  "IFIT3","IFN", "Function",
  "ISG15","IFN", "Function",
  "STAT1","IFN", "Function",
  "IFI6","IFN", "Function",
  "IRF7","IFN", "Function",
  "MX1","IFN", "Function",
  "IFITM3","IFN", "Function",

  # Inflammatory / NFkB
  "TNFAIP3","Inflammatory", "Function",
  "NFKB1","Inflammatory", "Function",
  "NFKBIA","Inflammatory", "Function",
  "NFKBIZ","Inflammatory", "Function",
  "STAT4","Inflammatory", "Function",

  # Memory
  "CD44","Memory", "State",
  "PTPRC","Memory", "State",
  "CCR7","Memory", "State",
  "CD27","Memory", "State",
  "CD28","Memory", "State",

  # Naive
  "CCR7","Naif", "State",
  "SELL","Naif", "State",
  "CD27","Naif", "State",
  "CD28","Naif", "State",
  "TCF7","Naif", "State",
  "IL7R","Naif", "State",
  "LEF1","Naif", "State",

  # Proliferating
  "CENPV","Prolif", "State",
  "MKI67","Prolif", "State",
  "NCAPG","Prolif", "State",
  "KIFC1","Prolif", "State",
  "CENPE","Prolif", "State",
  "ORC6","Prolif", "State",

  # Resident (Trm)
  "CD69","Resident", "State",
  "CXCR6","Resident", "State",
  "NR4A2","Resident", "State",
  "RUNX3","Resident", "State",

  # Senescent
  "KLRG1","Senescent", "State",
  "B3GAT1","Senescent", "State",
  "CD27","Senescent", "State",
  "CD28","Senescent", "State",
  "LAMP1","Senescent", "State",
  "IL1A","Senescent", "State",
  "FAS","Senescent", "State",
  "TP53","Senescent", "State",
  "TP53BP1","Senescent", "State",
  "CDKN1A","Senescent", "State",
  "CDKN2A","Senescent", "State",
  "MKI67","Senescent", "State",

  # Tfh
  "IL21","Tfh", "Identity",
  # "FR4","Tfh", "Identity", # Only in mouse
  "CXCR5","Tfh", "Identity",
  "BCL6","Tfh", "Identity",
  "PDCD1","Tfh", "Identity",
  "ICOS","Tfh", "Identity",

  # Th1
  "CX3CR1","Th1", "Identity",
  "IFNG","Th1", "Identity",
  "TBX21","Th1", "Identity",
  "CXCR3","Th1", "Identity",
  "STAT4","Th1", "Identity",
  "IL12RB2","Th1", "Identity",

  # Th2
  "GATA3","Th2", "Identity",
  "CCR4","Th2", "Identity",
  "IL4","Th2", "Identity",
  "IL5","Th2", "Identity",
  "IL13","Th2", "Identity",

  # Th17
  "RORC","Th17", "Identity",
  "IL17A","Th17", "Identity",
  "CCR6","Th17", "Identity",
  "IL17F","Th17", "Identity",
  "IL23R","Th17", "Identity",

  # Treg
  "FOXP3","Treg", "Identity",
  "IL2RA","Treg", "Identity",
  "CTLA4","Treg", "Identity",
  "TIGIT","Treg", "Identity",
  "IKZF2","Treg", "Identity",
  "CD27","Treg", "Identity",
  "IL7R","Treg", "Identity"
)
```

### Visualization of all signatures together {.tabset}

```{r}
Tcell_signature <- "
^CD3D$|^CD3E$|^CD4$|^TRBC1$|
^FCGR3A$|^CD8A$|^CD8B$|^NCR1$|
^CX3CR1$|^IFNG$|^TBX21$|^CXCR3$|^STAT4$|^IL12RB2$|
^GATA3$|^CCR4$|^IL4$|^IL5$|^IL13$|
^RORC$|^IL17A$|^CCR6$|^IL17F$|^IL23R$|
^IL21$|^CXCR5$|^BCL6$|^PDCD1$|^ICOS$|
^FOXP3$|^IL2RA$|^CTLA4$|^TIGIT$|^IKZF2$|^CD27$|^IL7R$|
^PRF1$|^GZMB$|^GNLY$|^GZMH$|^CCL5$|^NKG7$|^KLRG1$|^GZMA$|
^LAG3$|^TOX$|^CD226$|^GZMK$|^HAVCR2$|^XCL1$|^XCL2$|^CD160$|
^B3GAT1$|^CD28$|^LAMP1$|^IL1A$|^FAS$|^TP53$|^TP53BP1$|^CDKN1A$|^CDKN2A$|^MKI67$|
^CD69$|^CXCR6$|^NR4A2$|^RUNX3$|^TNFRSF9$|
^CENPV$|^NCAPG$|^KIFC1$|^CENPE$|^ORC6$|
^CCR7$|^SELL$|^TCF7$|^LEF1$|^Tcells4$|^PTPRC$|
^MALAT1$|^NEAT1$|
^IFIT2$|^IFIT3$|^ISG15$|^STAT1$|^IFI6$|^IRF7$|^MX1$|^IFITM3$|
^TNFAIP3$|^NFKB1$|^NFKBIA$|^NFKBIZ$|
^HLA-DRA$|^HLA-DRB1$
"


selected_features <- grep(Tcell_signature, rownames(sce), value=TRUE)


m2 <- presto_markers %>%
  filter(feature %in% selected_features) %>%
  select(feature, group, avgExpr, pct_in) %>%
  left_join(signature_tbl, by = "feature") %>%
  mutate(signature_class = case_when(
  cell_type %in% c("CD4","CD8","Th1","Th2","Th17","Tfh","Treg") ~ "Identity",
  cell_type %in% c("Cytotoxic","IFN","Inflammatory","HLA") ~ "Function",
  cell_type %in% c("Naif","Memory","Activated","Exhausted",
                    "Resident","Prolif","Senescent","Apoptotic") ~ "State"
)) %>%
  filter(!is.na(cell_type)) %>%
  group_by(feature) %>%
  mutate(
    avgExpr = avgExpr - mean(avgExpr),
    avgExpr = scales::rescale(avgExpr, to = c(-2,2))
  ) %>%
  ungroup()

ggplot(m2, 
       aes(x = feature,
               y = fct_reorder(as.character(group), group),
               color = avgExpr,
               size = pct_in)) +
  geom_point() +
  scale_color_gradient2(low="blue", mid="white", high="red") +
  facet_wrap(~cell_type, scales="free", ncol=10) +
  theme(
    axis.text.x = element_text(angle=45, hjust=1),
    axis.text.y = element_text(angle=45, hjust=1)
  ) +
  labs(
    title = "LT Compartment - All Signatures",
    x = "Genes markers",
    y = "Clusters"
  )

dir.create(
   file.path(
     DIR_RESULT,
     "All_Tcells",
     "Annotation",
     "Manual"),
   recursive = TRUE,
   showWarnings = FALSE)

ggsave(
       filename = file.path(
         DIR_RESULT,
         "All_Tcells",
         "Annotation",
         "Manual",
         "Marker_Signatures_LT_Compartment_All_Signatures.png"),
       width = 20, height = 10)
)
```

### View of specific signatures {.tabset}

```{r Viz specific signature}
m3 <- presto_markers %>%
  filter(feature %in% selected_features) %>%
  select(feature, group, avgExpr, pct_in) %>%
  left_join(signature_tbl, by = "feature") %>%
  filter(!is.na(cell_type)) %>%
  group_by(feature) %>%
  mutate(
    avgExpr = avgExpr - mean(avgExpr),
    avgExpr = scales::rescale(avgExpr, to = c(-2,2))
  ) %>%
  ungroup()

for (i in unique(m3$class)) {
  p <- ggplot(m3 %>% filter(class == i), 
         aes(x=feature, y=fct_reorder(as.character(group), group), color = avgExpr, size = pct_in)) +
    geom_point()+
    scale_color_gradient2(low = "blue", mid="white",high = "red")+
    # theme_minimal() +
    facet_wrap(~ cell_type, scales = "free", ncol = 4) +
    theme(axis.text.y = element_text(angle=45, hjust=1), 
          axis.text.x = element_text(angle=45, hjust=1)) +
    labs(
    title = paste0("LT Compartment - ", i),
    x = "Genes markers",
    y = "Clusters"
  )
  
  print(p)
  
  dir.create(
    file.path(
      DIR_RESULT,
      "All_Tcells",
      "Annotation",
      "Manual"),
    recursive = TRUE,
    showWarnings = FALSE
  )
  
  ggsave(
    filename = file.path(
      DIR_RESULT,
      "All_Tcells",
      "Annotation",
      "Manual",
      paste0("Marker_Signatures_LT_Compartment_", i, ".png")),
    plot = p,
    width = 20, height = 10
  )
}
```

For now I think that we don't need to be to specific on pop annotation because we have all Tcells. For now we have to identified major populations (CD4, CD8) by combining annotation from me + automatic + from PPA to be sure that we analyse the right populations later.

And then we extract count for specific pop cd8 and cd4 cells and perform a new clustering/annotation on each of them separately to have more precise subpopulations.

```{r Combine annot}
cluster_annotation_list <- list(
  "0" = "CD4_naive_and_memory",
  "1" = "CD8_Cytotox_exhausted",
  "2" = "CD4_inflammed",
  "3" = "CD8_cytotox_low_resident_inflammed",
  "4" = "CD8_naive",
  "5" = "???",
  "6" = "Dying_Tcells",
  "7" = "CD8_IFN"
)

# Add annotations to Seurat object
seu$manual_annotation <- plyr::mapvalues(
  seu$seurat_clusters,
  from = names(cluster_annotation_list),
  to = unlist(cluster_annotation_list)
)

# Names of cells with matching annotations between automatic (mine and PPA SingleR predictions) and manual (mine) annotation
## But first let's harmonize the names between predicted.celltype.l3_jordan and predicted.celltype.l3
seu$predicted.celltype.l3_jordan <- plyr::mapvalues(
  seu$predicted.celltype.l3_jordan,
  from = c("Non-Vd2 gd T cells", "Naive CD4 T cells", "Follicular helper T cells", "Terminal effector CD8 T cells", "Vd2 gd T cells", "MAIT cells", "Terminal effector CD4 T cells", "Th1/Th17 cells", "Th17 cells", "Th1 cells", "Th2 cells", "Effector memory CD8 T cells", "T regulatory cells", "Central memory CD8 T cells", "Naive CD8 T cells", "Natural killer cells", "Myeloid dendritic cells", "Progenitor cells"),
  to = c("Non-Vd2 gd T cells", "Naive CD4 T cells", "Follicular helper T cells CD4", "Terminal effector CD8 T cells", "Vd2 gd T cells", "MAIT cells", "Terminal effector CD4 T cells", "Th1/Th17 cells CD4", "Th17 cells CD4", "Th1 cells CD4", "Th2 cells CD4", "Effector memory CD8 T cells", "T regulatory cells CD4", "Central memory CD8 T cells", "Naive CD8 T cells", "Natural killer cells", "Myeloid dendritic cells", "Progenitor cells")
)



matched_cells <- WhichCells(
  seu,
  expression =
    (
      grepl("CD4", predicted.celltype.l3_jordan) &
      grepl("CD4", predicted.celltype.l1) &
      grepl("CD4", manual_annotation)
    ) |
    (
      grepl("CD8", predicted.celltype.l3_jordan) &
      grepl("CD8", predicted.celltype.l1) &
      grepl("CD8", manual_annotation)
    )
)
length(matched_cells)
length(seu@meta.data$manual_annotation)
# Le nombre de cellules avec annotation "Natural killer cells" dans predicted.celltype.l3_jordan
length(
  predicted_cells <- WhichCells(
    seu,
    expression = (predicted.celltype.l1 == "other T")
))

```

There is a lot of cells annotate gdTcells but I will not focus on them for now as I want to focus on CD4 and CD8 Tcells, and there are too different from cd4+ and cd8+.

# Session Info {.tabset .unnumbered}

```{r Session Info}
sessionInfo()
```

::: callout-note
This is a note.
:::

::: callout-tip
This is a tip.
:::

::: callout-warning
This is a warning.
:::

::: callout-important
Important message.
:::

::: callout-caution
Caution!
:::

::: {.callout-tip title="Pro Tip" icon="lightbulb" collapse="true"}
Use `SCTransform` instead of log-normalization for variance stabilization.
:::

::: {.panel-tabset .panel-tabset-pills}
## SCTransform

Plot A

## Log-normalization

Plot B
:::

::: my-highlight
Some highlighted text.
:::
