---
title: "HD vs MM – scRNA-seq All T Cell Analysis"
subtitle: "Integrated multi-sample reanalysis"
author: 
  - name: Jordan Dutel
    affiliation: 
      - CRCT – CHU Toulouse
      - GENIM Team
    email: jordan.dutel@inserm.fr
date: today
date-format: "DD MMMM YYYY"

lang: en
title-block-banner: true
title-block-style: default

format:
  html:
    theme:
      light: cosmo
      dark: darkly
    # theme: superhero         # Bootstrap
    css: styles.css
    toc: true
    toc-title: Table of Contents
    toc-depth: 4
    toc-location: left
    toc-expand: 2
    number-sections: true
    number-depth: 3
    page-layout: full
    smooth-scroll: true
    anchor-sections: true
    section-divs: true
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    code-copy: true
    code-overflow: wrap
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    highlight-style: github
    embed-resources: true
    fig-responsive: true
    fig-format: retina
    fig-align: center
    fig-dpi: 300
    df-print: paged
    other-links:
      - text: Thomas's PPT
        icon: filetype-ppt
        href: https://spam.inserm.fr/fmlurlsvc/?fewReq=:B:JV07MDQzOyd3PDMvMSdoZTwxMDsxMCdyaGZvYHV0c2Q8NGUxM2M4Nzk2OWAwMmUzNTU4OWM1NDI4MTIzN2Q4NzFkYjFlYjAxOCd1PDA2NjE4MjE1NDAncGhlPDcwQk02VFBEMTM3MTE5LDcwQk02VFBGMTM3MTE5J3NicXU8a25zZWBvL2V0dWRtQWhvcmRzbC9ncydiPDQzJ2llbTwx&url=https%3a%2f%2ficons.getbootstrap.com%2f%3fq%3dppt 
    
execute:
  echo: true
  warning: false
  message: false
  error: false
  cache: true
  freeze: auto
  eval: true

editor: visual
---

```{r Setup, echo=FALSE}
############
# PACKAGES #
############

# BiocManager::install("SingleCellExperiment")
# install.packages("devtools")
# BiocManager::install("DropletUtils")
# devtools::install_github("immunogenomics/presto")
# BiocManager::install("scDblFinder")

library(DropletUtils)
library(SingleCellExperiment)
library(scater)
library(scran)
library(qs)
library(Seurat)
library(ggplot2)
library(patchwork)
library(clustree)
library(harmony)
library(patchwork)
library(dplyr)
library(ggridges)
library(readr)
library(knitr)
library(kableExtra)
library(SingleR)
library(celldex)
library(ComplexHeatmap)
library(circlize)
library(presto)
library(tidyverse)
library(tibble)
library(matrixStats)
library(viridis)
library(scuttle)
library(scDblFinder)
library(ggbeeswarm)
library(here)
library(scales)
library(Polychrome)
library(future)
library(glmGamPoi)
library(lisi)
library(pheatmap)
library(cluster)
library(ggrepel)
library(rstatix)
library(ggpubr)
library(pals)

###########
# OPTIONS #
###########

options(
  stringsAsFactors = FALSE,
  scipen = 999
)

options(future.globals.maxSize = 4 * 1024^3)  # 4 Go


plan(sequential)

#########
# PATHS #
#########

here::i_am("04_report/HD_vs_MM_scRNAseq_All_Tcells_analysis.qmd")
# here::i_am("02_script/HD_vs_MM_scRNAseq_reanalysis.qmd")

# knitr::opts_knit$set(
#   root.dir = "~/Documents/Project/Thomas/20260116_scRNAseq_HD_MM_reanalysis")

seurat_path_Tcell_obj <- "/mnt/crct-share/crct13/PPA/PPA_work/project/01_HCvsMM/01A_HCvsMM_10x/results/Tcells_analysis/seurat_objects/QCed_seurat_Tcells_clean.qs"

# DIR_RAW    <- here("01_raw")
DIR_DATA   <- here("01_data")
# DIR_PROCESSED_DATA <- here("02_processed_data")
DIR_SCRIPT <- here("02_script")
DIR_RESULT <- here("03_result")
DIR_REPORT <- here("04_report")
DIR_ENV <- here("05_environment")
# DIR_QC     <- here("03_result", "QC")

dir.create(here(DIR_DATA, "Processed_data"),
           recursive = TRUE, 
           showWarnings = FALSE)



##########
# COLORS #
##########

colors_30 <- c(
  "#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00",
  "#A65628", "#F781BF", "#999999", "#66C2A5", "#FC8D62",
  "#8DA0CB", "#E78AC3", "#A6D854", "#FFD92F", "#E5C494",
  "#B3B3B3", "#1B9E77", "#D95F02", "#7570B3", "#E7298A",
  "#66A61E", "#E6AB02", "#A6761D", "#666666", "#8DD3C7",
  "#BEBADA", "#FB8072", "#80B1D3", "#FDB462", "#B3DE69"
)
colors_55 <- c(
  "#E41A1C","#377EB8","#4DAF4A","#984EA3","#FF7F00",
  "#A65628","#F781BF","#999999","#66C2A5","#FC8D62",
  "#8DA0CB","#E78AC3","#A6D854","#FFD92F","#E5C494",
  "#1B9E77","#D95F02","#7570B3","#E7298A","#66A61E",
  "#E6AB02","#A6761D","#666666","#8DD3C7","#BEBADA",
  "#FB8072","#80B1D3","#FDB462","#B3DE69","#FCCDE5",
  "#D9D9D9","#BC80BD","#CCEBC5","#FFED6F","#1F78B4",
  "#33A02C","#FB9A99","#E31A1C","#FDBF6F","#CAB2D6",
  "#6A3D9A","#FFFF99","#B15928","#8E0152","#DE77AE",
  "#7FBC41","#4EB265","#4393C3","#2166AC","#B2182B",
  "#F4A582","#92C5DE","#FDDBC7","#D1E5F0","#542788"
)

# 22 col palette
pal <-c('#e6194B', '#3cb44b', '#ffe119', '#4363d8', '#f58231', '#911eb4', '#42d4f4', '#f032e6', '#bfef45', '#fabed4', '#9A6324', '#469990', '#dcbeff', '#fffac8', '#800000', '#aaffc3', '#808000', '#ffd8b1', '#000075', '#a9a9a9', '#ffffff', '#000000')

Glasbey = glasbey.colors(32)
swatch(Glasbey)

# --- Sanity check ---
cat("Project root:", here(), "\n")
```

# Load data {.tabset}

```{r Load raw data}
seu <- qread(seurat_path_Tcell_obj)
```

# QC {.tabset}

### Calculations and vizualizations before cleaning {.tabset}

Maybe for each metrics, create a col in metatadata with status_metric (normal, low, high) based on 3\*MAD from median per emulsion AND another col based on isOutlier. Then perform cutting/cutoff based on these status cols. Non Stringent QC : remove cells that are outliers only in both of the methods ; Stringent QC : remove cells that are outliers in at least one of the methods.

And then Global Stringent QC : remove cells that are outliers in at least one of the methods for at least one of the metrics. Global Non Stringent QC : remove cells that are outliers only in both of the methods for all of the metrics (or critical metrics)

Maybe found critical metrics (not all metrics for QC stringence)

We use the function isOutlier from scuttle package to identify outliers for each metrics and for each emulsion. This function uses the median and the MAD to identify outliers, but it also takes into account the distribution of the data and can be more robust than a simple and global cutoff based on 3\*MAD.

Cut off/ OUTLIER DETECTION BY TECHNICAL BATCH (more qualitative QC than a global cut)

!!! Be careful to perform QC on counts slot (raw data) !!!

#### Cells count {.tabset}

##### Visualize {.tabset}

```{r QC Tcells Viz cell counts 1}
seu_Tcells@meta.data %>% 
  group_by(emulsion) %>%
  summarise(cell_count = n())
```

```{r QC Tcells Viz cell counts 2}
features <- c("condition", "id")

dir.create(
  file.path(DIR_RESULT, 
            "All_Tcells",
            "QC",
            "Cell_Counts"),
  recursive = TRUE,
  showWarnings = FALSE)


for (feature in features) {
  p <- seu_Tcells@meta.data %>%
    ggplot(aes(x = .data[[feature]], fill = .data[[feature]])) +
    geom_bar() +
    scale_fill_manual(values = pal, drop = FALSE) +
    theme_classic() +
    theme(
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
      plot.title = element_text(hjust = 0.5, face = "bold")
    ) +
    ggtitle(paste("NCells -", feature)) +
    NoLegend()

  print(p)
  
  ggsave(
  plot = p,
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Cell_Counts",
    "Before_QC",
    paste0("Cell_Counts_By_", feature, "_Before_QC_All_Tcells.png")),
  width = 6,
  height = 4)
}


features <- c("emulsion")

for (feature in features) {
  p <- seu_Tcells@meta.data %>%
    ggplot(aes(x = .data[[feature]], fill = .data[[feature]])) +
    geom_bar() +
    # scale_fill_manual(values = pal, drop = FALSE) +
    theme_classic() +
    theme(
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
      plot.title = element_text(hjust = 0.5, face = "bold")
    ) +
    ggtitle(paste("NCells -", feature)) +
    NoLegend()

  print(p)
  
  ggsave(
  plot = p,
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Cell_Counts",
    "Before_QC",
    paste0("Cell_Counts_By_", feature, "_Before_QC_All_Tcells.png")),
  width = 6,
  height = 4)
}
```

```{r QC Tcells Viz cell counts 3}
DimPlot(seu_Tcells, group.by = "emulsion") +
  facet_wrap(~emulsion, ncol = 6) +
  ggtitle("UMAP by emulsion before QC All Tcells") + 
  NoLegend()

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Cell_Counts",
    "Before_QC",
    "UMAP_Cell_Counts_by_emulsion_Before_QC_All_Tcells.png"),
  width = 23,
  height = 12)
```

#### Mitochondria {.tabset}

##### Run {.tabset}

```{r QC Tcells Run Mito}
# !!! Be careful to perform QC on counts slot (raw data) !!!

qc_mito <- perCellQCMetrics(
  seu_Tcells@assays$RNA@counts,
  subsets = list(mito = grep("^MT-", rownames(seu_Tcells@assays$RNA@counts), value = TRUE)))

seu_Tcells <- AddMetaData(seu_Tcells, as.data.frame(qc_mito))


qc_mito2 <- data.frame(
  emulsion = seu_Tcells@meta.data$emulsion,
  subsets_mito_percent = seu_Tcells@meta.data$subsets_mito_percent
)

qc_mito2$high_mito_outliers <- isOutlier(
  seu_Tcells@meta.data$subsets_mito_percent,
  nmads = 3,
  type = "higher"
)

qc_mito2 <- qc_mito2 %>%
  dplyr::group_by(emulsion) %>%
  dplyr::mutate(
    mito_status = dplyr::case_when(
      high_mito_outliers == TRUE ~ "high",
      TRUE ~ "normal"
    )
  ) %>%
  dplyr::ungroup() %>% 
  filter(!is.na(mito_status))

seu_Tcells <- AddMetaData(seu_Tcells, qc_mito2$mito_status, col.name = "mito_status")


```

##### Visualize {.tabset}

```{r QC Tcells Viz Mito 1}
ggplot(seu_Tcells@meta.data, aes(x = subsets_mito_percent, y = emulsion, fill = after_stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01
  ) +
  scale_fill_viridis(
    alpha = 0.9,
    name = "subsets_mito_percent",
    option = "C",
    guide = "none"
  ) +
  labs(
    title = "Mitochondrial content by emulsion",
    x = "Percent mitochondrial reads",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.spacing = unit(0.1, "lines"),
    strip.text.x = element_text(size = 8),
    plot.title = element_text(face = "bold")
  )

dir.create(
  file.path(DIR_RESULT, 
            "All_Tcells",
            "QC",
            "Mitochondria",
            "Before_QC"),
  recursive = TRUE,
  showWarnings = FALSE)

ggsave(
  filename = file.path(DIR_RESULT,
                       "All_Tcells",
                       "QC",
                       "Mitochondria",
                       "Before_QC",
                       "Ridgeplot_mitochondrial_content_by_emulsion_before_QC_All_Tcells.png"),
  width = 8,
  height = 5)
```

```{r QC Tcells Viz Mito 2}
FeaturePlot(
  seu_Tcells,
  features = "subsets_mito_percent",
  reduction = "umap",
  order = TRUE,
  cols = c("lightgrey", "red")
)

ggsave(
  filename = file.path(DIR_RESULT,
                       "All_Tcells",
                       "QC",
                       "Mitochondria",
                       "Before_QC",
                       "UMAP_mitochondrial_content_before_QC_All_Tcells.png"),
  width = 6,
  height = 5)
```

```{r QC Tcells Viz Mito 3}
ggplot(seu_Tcells@meta.data, aes(x = emulsion, y = subsets_mito_percent, color = mito_status)) +
  geom_violin(fill = "grey90", color = NA) +
  geom_quasirandom(alpha = 0.5, size = 0.8) +
  scale_y_log10() +
  scale_color_manual(
    values = c("normal" = "green3", "high" = "red3")
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))


ggsave(
  filename = file.path(DIR_RESULT,
                       "All_Tcells",
                       "QC",
                       "Mitochondria",
                       "Before_QC",
                       "ViolinPlot_mito_percent_by_emulsion.png"),
  width = 12,
  height = 6)

```

There are some cells with high mito content in some emulsions. We will remove these cells later

#### Ribosomes {.tabset}

##### Run {.tabset}

ADD AUTOMATIC MADS outliers by function isOutlier

```{r QC Tcells Run Ribo}
# !!! Be careful to perform QC on counts slot (raw data) !!!

qc_ribo <- perCellQCMetrics(
  seu_Tcells@assays$RNA@counts,
  subsets = list(
    ribo = grep("^RPL|^RPS", rownames(seu_Tcells@assays$RNA@counts), value = TRUE)))

seu_Tcells <- AddMetaData(seu_Tcells, as.data.frame(qc_ribo))

qc_ribo2 <- data.frame(
  emulsion = seu_Tcells@meta.data$emulsion,
  subsets_mito_percent = seu_Tcells@meta.data$subsets_ribo_percent
)

qc_ribo2$high_ribo_outliers <- isOutlier(
  seu_Tcells@meta.data$subsets_ribo_percent,
  nmads = 3,
  type = "higher"
)

qc_ribo2$lower_ribo_outliers <- isOutlier(
  seu_Tcells@meta.data$subsets_ribo_percent,
  nmads = 3,
  type = "lower"
)

qc_ribo2 <- qc_ribo2 %>%
  dplyr::group_by(emulsion) %>%
  dplyr::mutate(
    ribo_status = dplyr::case_when(
      high_ribo_outliers == TRUE ~ "high",
      lower_ribo_outliers == TRUE ~ "low",
      TRUE ~ "normal"
    )
  ) %>%
  dplyr::ungroup() %>% 
  filter(!is.na(ribo_status))

seu_Tcells <- AddMetaData(seu_Tcells, qc_ribo2$ribo_status, col.name = "ribo_status")

```

##### Visualize {.tabset}

```{r QC Tcells Viz Ribo 1}
ggplot(seu_Tcells@meta.data, aes(x = subsets_ribo_percent, y = emulsion, fill = after_stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01
  ) +
  scale_fill_viridis(
    alpha = 0.9,
    name = "subsets_ribo_percent",
    option = "C",
    guide = "none"
  ) +
  labs(
    title = "Ribosomal content by emulsion",
    x = "Percent ribosomal reads",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.spacing = unit(0.1, "lines"),
    strip.text.x = element_text(size = 8),
    plot.title = element_text(face = "bold")
  )

dir.create(
  file.path(DIR_RESULT,
            "All_Tcells",
            "QC", 
            "Ribosomal",
            "Before_QC"),
  recursive = TRUE,
  showWarnings = FALSE)

ggsave(
  filename = file.path(DIR_RESULT,
                       "All_Tcells",
                       "QC",
                       "Ribosomal",
                       "Before_QC",
                       "Ridgeplot_ribosomal_content_by_emulsion_before_QC_All_Tcells.png"),
  width = 8,
  height = 5)
```

```{r QC Tcells Viz Ribo 2}
FeaturePlot(
  seu_Tcells,
  features = "subsets_ribo_percent",
  reduction = "umap",
  order = TRUE,
  cols = c("lightgrey", "red"))

ggsave(
  filename = file.path(DIR_RESULT,
                       "All_Tcells",
                       "QC",
                       "Ribosomal",
                       "Before_QC",
                       "UMAP_ribosomal_content_before_QC_All_Tcells.png"),
  width = 6,
  height = 5)
```

```{r QC Tcells Viz Ribo 3}
ggplot(seu_Tcells@meta.data, aes(x = emulsion, y = subsets_ribo_percent, color = ribo_status)) +
  geom_violin(fill = "grey90", color = NA) +
  geom_quasirandom(alpha = 0.5, size = 0.8) +
  scale_y_log10() +
  scale_color_manual(
    values = c("normal" = "green3", "high" = "red3", "low" = "blue3")
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))


ggsave(
  filename = file.path(DIR_RESULT,
                       "All_Tcells",
                       "QC",
                       "Ribosomal",
                       "Before_QC",
                       "ViolinPlot_ribo_percent_by_emulsion.png"),
  width = 12,
  height = 6)
```

#### Library size (nCountRNA) {.tabset}

##### Run {.tabset}

Replace manual MADS by function isOutlier

```{r QC Tcells Run Libsize}
qc_lib_size <- seu_Tcells@meta.data %>%
  dplyr::group_by(emulsion) %>%
  dplyr::mutate(
    libsize_status = dplyr::case_when(
      nCount_RNA < median(nCount_RNA) - 3 * mad(nCount_RNA) ~ "low",
      nCount_RNA > median(nCount_RNA) + 3 * mad(nCount_RNA) ~ "high",
      TRUE ~ "normal"
    )
  ) %>%
  dplyr::ungroup() %>% 
  filter(nCount_RNA > 0)


# qc_lib_size <- seu_Tcells@meta.data %>%
#   group_by(emulsion) %>%
#   mutate(
#     libsize_status = dplyr::if_else(
#       isOutlier(nCount_RNA, nmads = 3, type = "both"),
#       "outlier",
#       "normal"
#     )
#   ) %>%
#   ungroup()


# Add libsize_status to metadata
seu_Tcells <- AddMetaData(seu_Tcells, qc_lib_size$libsize_status, col.name = "libsize_status")
```

##### Visualize {.tabset}

```{r QC Tcells Viz Libsize 1}
ggplot(qc_lib_size, aes(x = nCount_RNA)) +
  geom_density(fill = "grey80", alpha = 0.6) +
  geom_rug(
    aes(color = libsize_status),
    alpha = 0.6,
    sides = "b"
  ) +
  scale_x_log10() +
  scale_color_manual(
    values = c(
      low = "blue",
      normal = "green",
      high = "red"
    )
  ) +
  facet_wrap(~emulsion) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  ) +
  labs(
    color = "Library size QC",
    x = "nCount_RNA (log10)",
    y = "Density"
  )

dir.create(
  file.path(DIR_RESULT,
            "All_Tcells",
            "QC", 
            "Library_Size",
            "Before_QC"),
  recursive = TRUE,
  showWarnings = FALSE)

ggsave(
  filename = file.path(DIR_RESULT,
                       "All_Tcells",
                       "QC",
                       "Library_Size",
                       "Before_QC",
                       "DensityPlot_library_size_density_by_emulsion.png"),
  width = 12,
  height = 6)
```

```{r QC Tcells Viz Libsize 2}
ggplot(qc_lib_size, aes(x = emulsion, y = nCount_RNA, color = libsize_status)) +
  geom_violin(fill = "grey90", color = NA) +
  geom_quasirandom(alpha = 0.5, size = 0.8) +
  scale_y_log10() +
  scale_color_manual(
    values = c(low = "blue", normal = "green", high = "red")
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))


ggsave(
  filename = file.path(DIR_RESULT,
                       "All_Tcells",
                       "QC",
                       "Library_Size",
                       "Before_QC",
                       "ViolinPlot_library_size_by_emulsion.png"),
  width = 12,
  height = 6)
```

#### Library diversity/heterogeneity (nFeatureRNA) {.tabset}

##### Run {.tabset}

Replace manual MADS by function isOutlier

```{r QC Tcells Run Libdiv}
qc_lib_diversity <- seu_Tcells@meta.data %>%
  dplyr::group_by(emulsion) %>%
  dplyr::mutate(
    libdiv_status = dplyr::case_when(
      nFeature_RNA < median(nFeature_RNA) - 3 * mad(nFeature_RNA) ~ "low",
      nFeature_RNA > median(nFeature_RNA) + 3 * mad(nFeature_RNA) ~ "high",
      TRUE ~ "normal"
    )
  ) %>%
  dplyr::ungroup() %>% 
  filter(nFeature_RNA > 0)

# Add libdiv_status to metadata
seu_Tcells <- AddMetaData(seu_Tcells, qc_lib_diversity$libdiv_status, col.name = "libdiv_status")
```

##### Visualize {.tabset}

```{r QC Tcells Viz Libdiv 1}
ggplot(qc_lib_diversity, aes(x = nFeature_RNA)) +
  geom_density(fill = "grey80", alpha = 0.6) +
  geom_rug(
    aes(color = libdiv_status),
    alpha = 0.6,
    sides = "b"
  ) +
  scale_x_log10() +
  scale_color_manual(
    values = c(
      low = "blue",
      normal = "green",
      high = "red"
    )
  ) +
  facet_wrap(~emulsion) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  ) +
  labs(
    color = "Library diversity QC",
    x = "nFeatureRNA (log10)",
    y = "Density"
  )

dir.create(
  file.path(DIR_RESULT,
            "All_Tcells",
            "QC", 
            "Library_Diversity",
            "Before_QC"),
  recursive = TRUE,
  showWarnings = FALSE)

ggsave(
  filename = file.path(DIR_RESULT,
                       "All_Tcells",
                       "QC",
                       "Library_Diversity",
                       "Before_QC",
                       "DensityPlot_library_diversity_density_by_emulsion.png"),
  width = 12,
  height = 6)
```

```{r QC Tcells Viz Libdiv 2}
ggplot(qc_lib_diversity, aes(x = emulsion, y = nFeature_RNA, color = libdiv_status)) +
  geom_violin(fill = "grey90", color = NA) +
  geom_quasirandom(alpha = 0.5, size = 0.8) +
  scale_y_log10() +
  scale_color_manual(
    values = c(low = "blue", normal = "green", high = "red")
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

ggsave(
  filename = file.path(DIR_RESULT,
                       "All_Tcells",
                       "QC",
                       "Library_Diversity",
                       "Before_QC",
                       "ViolinPlot_library_diversity_by_emulsion.png"),
  width = 12,
  height = 6)
```

#### Librairy complexity {.tabset}

##### Run {.tabset}

```{r}
seu_Tcells$log10_Complexity <- log10(seu_Tcells$nFeature_RNA) / log10(seu_Tcells$nCount_RNA)

qc_lib_complexity <- data.frame(
  emulsion = seu_Tcells$emulsion,
  log10_Complexity = seu_Tcells$log10_Complexity
)

qc_lib_complexity$low_complexity_outliers <- isOutlier(
  seu_Tcells$log10_Complexity,
  nmads = 3,
  type = "lower"
)

qc_lib_complexity$high_complexity_outliers <- isOutlier(
  seu_Tcells$log10_Complexity,
  nmads = 3,
  type = "higher"
)

qc_lib_complexity <- qc_lib_complexity %>%
  dplyr::group_by(emulsion) %>%
  dplyr::mutate(
    libcomplex_status = dplyr::case_when(
      low_complexity_outliers == TRUE ~ "low",
      high_complexity_outliers == TRUE ~ "high",
      TRUE ~ "normal"
    )
  ) %>%
  dplyr::ungroup() %>% 
  filter(!is.na(log10_Complexity))
  

low_quantil_complexity <- quantile(seu_Tcells$log10_Complexity, 0.05, na.rm = TRUE)

# Add libcomplex_status to metadata
seu_Tcells <- AddMetaData(seu_Tcells, qc_lib_complexity$libcomplex_status, col.name = "libcomplex_status")
```

##### Visualize {.tabset}

```{r QC Tcells Viz Libcomplex 1}
ggplot(seu_Tcells@meta.data, aes(x = log10_Complexity)) +
  geom_density(fill = "lightgreen", alpha = 0.6) +
  geom_vline(xintercept = low_quantil_complexity, linetype = "dashed", color = "red") +
  labs(
    title = "Library complexity (log10GenesPerUMI)",
    subtitle = paste0("Red dashed line = 5% quantile (", round(low_quantil_complexity, 2), ")"),
    x = "log10GenesPerUMI"
  ) +
  theme_classic()

dir.create(
  file.path(DIR_RESULT,
            "All_Tcells",
            "QC", 
            "Library_Complexity",
            "Before_QC"),
  recursive = TRUE,
  showWarnings = FALSE)

ggsave(
  filename = file.path(DIR_RESULT,
                       "All_Tcells",
                       "QC",
                       "Library_Complexity",
                       "Before_QC",
                       "DensityPlot_library_complexity_before_QC_All_Tcells.png"),
  width = 12,
  height = 6)
```

```{r QC Tcells Viz Libcomplex 2}
# ggplot(seu_Tcells@meta.data, aes(x = nCount_RNA, y = nFeature_RNA)) +
#   geom_point(
#     data = subset(seu_Tcells@meta.data, !low_complexity_outliers),
#     color = "grey70",
#     alpha = 0.25,
#     size = 0.4
#   ) +
#   geom_point(
#     data = subset(seu_Tcells@meta.data, low_complexity_outliers),
#     color = "blue3",
#     alpha = 0.7,
#     size = 0.6
#   ) +
#   geom_smooth(formula = y ~ x, method = "lm", color = "black", linewidth = 0.6) +
#   scale_x_log10() +
#   scale_y_log10() +
#   labs(
#     title = "Library complexity",
#     subtitle = "Genes detected vs UMIs (log–log)",
#     x = "UMIs per cell",
#     y = "Genes per cell"
#   ) +
#   theme_classic(base_size = 11)

ggplot(seu_Tcells@meta.data, aes(x = nCount_RNA, y = nFeature_RNA, color = libcomplex_status)) +
  geom_point(alpha = 0.6, size = 0.6) +
  geom_smooth(formula = y ~ x, method = "lm", color = "black", linewidth = 0.6) +
  scale_color_manual(values = c("normal" = "grey70", "low" = "blue3", "high" = "red3")) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = "Library complexity",
    subtitle = "Genes(nFeature)/UMIs(nCount) per cell",
    x = "UMIs per cell",
    y = "Genes per cell",
    color = "Complexity status"
  ) +
  theme_classic(base_size = 11)


ggsave(
  filename = file.path(DIR_RESULT,
                       "All_Tcells",
                       "QC",
                       "Library_Complexity",
                       "Before_QC",
                       "ScatterPlot_library_complexity_before_QC_All_Tcells.png"),
  width = 12,
  height = 6)
```

```{r QC Tcells Viz Libcomplex 3}
ggplot(seu_Tcells@meta.data,
       aes(x = log10_Complexity, fill = emulsion)) +
  geom_density(alpha = 0.4) +
  geom_vline(
    seu_Tcells@meta.data,
    xintercept = median(seu_Tcells@meta.data$log10_Complexity) - 3 * mad(seu_Tcells@meta.data$log10_Complexity),
    linetype = "dashed",
    color = "blue3"
  ) +
  geom_vline(
    seu_Tcells@meta.data,
    xintercept = median(seu_Tcells@meta.data$log10_Complexity) + 3 * mad(seu_Tcells@meta.data$log10_Complexity),
    linetype = "dashed",
    color = "red3"
  ) +
  facet_wrap(~ emulsion, ncol = 4) +
  labs(
    title = "Library complexity by emulsion",
    x = "log10GenesPerUMI"
  ) +
  theme_classic() +
  theme(legend.position = "none")

ggsave(
  filename = file.path(DIR_RESULT,
                       "All_Tcells",
                       "QC",
                       "Library_Complexity",
                       "Before_QC",
                       "DensityPlot_library_complexity_by_emulsion_before_QC_All_Tcells.png"),
  width = 12,
  height = 6)
```

```{r QC Tcells Viz Libcomplex 4}
ggplot(seu_Tcells@meta.data,
       aes(x = emulsion, y = log10_Complexity, color = libcomplex_status)) +
  geom_violin(fill = "grey90", color = NA) +
  geom_quasirandom(alpha = 0.4, size = 0.6) +
  scale_color_manual(
    values = c(low = "blue", normal = "green", high = "red")
  ) +
  labs(
    title = "Library complexity per emulsion",
    y = "log10GenesPerUMI",
    x = NULL
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
  filename = file.path(DIR_RESULT,
                       "All_Tcells",
                       "QC",
                       "Library_Complexity",
                       "Before_QC",
                       "ViolinPlot_library_complexity_by_emulsion_before_QC_All_Tcells.png"),
  width = 12,
  height = 6)
```

```{r QC Tcells Viz Libcomplex 5}
ggplot(seu_Tcells@meta.data,
       aes(x = log10_Complexity, fill = libcomplex_status)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("high" = "red3", "normal" = "green3", "low" = "blue3")) +
  theme_classic()

ggsave(
  filename = file.path(DIR_RESULT,
                       "All_Tcells",
                       "QC",
                       "Library_Complexity",
                       "Before_QC",
                       "DensityPlot_library_complexity_outliers_before_QC_All_Tcells.png"),
  width = 12,
  height = 6)
```

#### Cell Cycle Scoring {.tabset}

##### Run {.tabset}

```{r QC Tcells Run cellcycle}
seu_Tcells <- CellCycleScoring(
  seu_Tcells,
  s.features = cc.genes$s.genes,
  g2m.features = cc.genes$g2m.genes,
  set.ident = FALSE
)
```

##### Visualize {.tabset}

```{r QC Tcells Viz cellcycle}
DimPlot(seu_Tcells, group.by = "Phase") +
  ggtitle("Cell cycle phase before QC All Tcells") +
  facet_wrap(~Phase)

dir.create(
  file.path(DIR_RESULT,
            "All_Tcells",
            "QC", 
            "Cell_Cycle",
            "Before_QC"),
  recursive = TRUE,
  showWarnings = FALSE)

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Cell_Cycle",
    "Before_QC",
    "UMAP_Cell_Cycle_Phase_before_QC_All_Tcells.png"),
  width = 12,
  height = 6)
```

Later, see if clusters are enriched in specific cell cycle phases.

### Cutoff {.tabset}

```{r QC Tcells Cutoff Stringent QC}
seu_Tcells@meta.data$Stringent_QC_status <- dplyr::case_when(
  # seu_Tcells$DblFinder_class == "doublet" ~ "fail_doublet",
  (seu_Tcells$mito_status == "high" | seu_Tcells$libsize_status == "low" | seu_Tcells$libdiv_status == "low" | seu_Tcells$libcomplex_status == "low") ~ "fail",
  TRUE ~ "pass"
)

# Number of cells that fail QC
table(seu_Tcells@meta.data$Stringent_QC_status)
```

```{r QC Tcells Cutoff Moderate QC}
seu_Tcells@meta.data$Moderate_QC_status <- dplyr::case_when(
  # seu_Tcells$DblFinder_class == "doublet" ~ "fail_doublet",
  (seu_Tcells$mito_status == "high" | seu_Tcells$libcomplex_status == "low") ~ "fail",
  TRUE ~ "pass"
)

# Number of cells that fail QC
table(seu_Tcells@meta.data$Moderate_QC_status)
```

```{r QC Tcells Cutoff Non Stringent QC}
seu_Tcells@meta.data$Non_stringent_QC_status <- dplyr::case_when(
  # seu_Tcells$DblFinder_class == "doublet" ~ "fail_doublet",
  (seu_Tcells$mito_status == "high" & seu_Tcells$libsize_status == "low" & seu_Tcells$libdiv_status == "low" & seu_Tcells$libcomplex_status == "low") ~ "fail",
  TRUE ~ "pass"
)

# Number of cells that fail QC
table(seu_Tcells@meta.data$Non_stringent_QC_status)
```

```{r QC Tcells Cutoff/Subset}
# Subset the Seurat object to keep only cells that pass QC
seu_Tcells_QC_Moderate <- subset(seu_Tcells, subset = Moderate_QC_status == "pass")
```

### Visualization AFTER cutoff {.tabset}

#### Cells count {.tabset}

```{r QC Tcells Viz cell counts After QC 1}
seu_Tcells_QC_Moderate@meta.data %>% 
  group_by(emulsion) %>%
  summarise(cell_count = n())
```

```{r QC Tcells Viz cell counts After QC 2}
features <- c("condition", "id")

dir.create(
  file.path(DIR_RESULT,
    "All_Tcells",
    "QC",
    "Cell_Counts", 
    "After_QC"),
  recursive = TRUE,
  showWarnings = FALSE)


for (feature in features) {
  p <- seu_Tcells_QC_Moderate@meta.data %>%
    ggplot(aes(x = .data[[feature]], fill = .data[[feature]])) +
    geom_bar() +
    scale_fill_manual(values = pal, drop = FALSE) +
    theme_classic() +
    theme(
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
      plot.title = element_text(hjust = 0.5, face = "bold")
    ) +
    ggtitle(paste("NCells -", feature)) +
    NoLegend()

  print(p)
  
  ggsave(
  plot = p,
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Cell_Counts",
    "After_QC",
    paste0("Cell_Counts_By_", feature, "_After_QC_All_Tcells.png")),
  width = 6,
  height = 4)
}


features <- c("emulsion")

for (feature in features) {
  p <- seu_Tcells_QC_Moderate@meta.data %>%
    ggplot(aes(x = .data[[feature]], fill = .data[[feature]])) +
    geom_bar() +
    # scale_fill_manual(values = pal2, drop = FALSE) +
    theme_classic() +
    theme(
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
      plot.title = element_text(hjust = 0.5, face = "bold")
    ) +
    ggtitle(paste("NCells -", feature)) +
    NoLegend()

  print(p)
  
  ggsave(
  plot = p,
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Cell_Counts",
    "After_QC",
    paste0("Cell_Counts_By_", feature, "_After_QC_All_Tcells.png")),
  width = 6,
  height = 4)
}
```

```{r QC Tcells Viz cell counts After QC 3}
DimPlot(seu_Tcells_QC_Moderate, group.by = "emulsion") +
  facet_wrap(~emulsion, ncol = 6) +
  ggtitle("UMAP by emulsion after QC All Tcells") + 
  NoLegend()

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Cell_Counts",
    "After_QC",
    "UMAP_Cell_Counts_by_emulsion_After_QC_All_Tcells.png"),
  width = 23,
  height = 12)
```

#### Mitochondria {.tabset}

##### Run {.tabset}

```{r QC Tcells Run Mito After QC}
# !!! Be careful to perform QC on counts slot (raw data) !!!

qc_mito <- perCellQCMetrics(
  seu_Tcells_QC_Moderate@assays$RNA@counts,
  subsets = list(mito = grep("^MT-", rownames(seu_Tcells_QC_Moderate@assays$RNA@counts), value = TRUE)))

seu_Tcells_QC_Moderate <- AddMetaData(seu_Tcells_QC_Moderate, as.data.frame(qc_mito))


qc_mito2 <- data.frame(
  emulsion = seu_Tcells_QC_Moderate@meta.data$emulsion,
  subsets_mito_percent = seu_Tcells_QC_Moderate@meta.data$subsets_mito_percent
)

qc_mito2$high_mito_outliers <- isOutlier(
  seu_Tcells_QC_Moderate@meta.data$subsets_mito_percent,
  nmads = 3,
  type = "higher"
)

qc_mito2 <- qc_mito2 %>%
  dplyr::group_by(emulsion) %>%
  dplyr::mutate(
    mito_status = dplyr::case_when(
      high_mito_outliers == TRUE ~ "high",
      TRUE ~ "normal"
    )
  ) %>%
  dplyr::ungroup() %>% 
  filter(!is.na(mito_status))

seu_Tcells_QC_Moderate <- AddMetaData(seu_Tcells_QC_Moderate, qc_mito2$mito_status, col.name = "mito_status")


```

##### Visualize {.tabset}

```{r QC Tcells Viz Mito After QC 1}
ggplot(seu_Tcells_QC_Moderate@meta.data, aes(x = subsets_mito_percent, y = emulsion, fill = after_stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01
  ) +
  scale_fill_viridis(
    alpha = 0.9,
    name = "subsets_mito_percent",
    option = "C",
    guide = "none"
  ) +
  labs(
    title = "Mitochondrial content by emulsion",
    x = "Percent mitochondrial reads",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.spacing = unit(0.1, "lines"),
    strip.text.x = element_text(size = 8),
    plot.title = element_text(face = "bold")
  )

dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC", 
    "Mitochondria",
    "After_QC"),
  recursive = TRUE,
  showWarnings = FALSE)

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Mitochondria",
    "After_QC", "Ridgeplot_mitochondrial_content_by_emulsion_after_QC_All_Tcells.png"),
  width = 8,
  height = 5)
```

```{r QC Tcells Viz Mito After QC 2}
FeaturePlot(
  seu_Tcells_QC_Moderate,
  features = "subsets_mito_percent",
  reduction = "umap",
  order = TRUE,
  cols = c("lightgrey", "red")
)

ggsave(
  filename = file.path(DIR_RESULT,
    "All_Tcells",
    "QC",
    "Mitochondria",
    "After_QC", "UMAP_mitochondrial_content_after_QC_All_Tcells.png"),
  width = 6,
  height = 5)
```

```{r QC Tcells Viz Mito After QC 3}
ggplot(seu_Tcells_QC_Moderate@meta.data, aes(x = emulsion, y = subsets_mito_percent, color = mito_status)) +
  geom_violin(fill = "grey90", color = NA) +
  geom_quasirandom(alpha = 0.5, size = 0.8) +
  scale_y_log10() +
  scale_color_manual(
    values = c("normal" = "green3", "high" = "red3")
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))


ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Mitochondria",
    "After_QC",
    "ViolinPlot_after_QC_mito_percent_by_emulsion.png"),
  width = 12,
  height = 6)

```

There are some cells with high mito content in some emulsions. We will remove these cells later

#### Ribosomes {.tabset}

##### Run {.tabset}

ADD AUTOMATIC MADS outliers by function isOutlier

```{r QC Tcells Run Ribo After QC}
# !!! Be careful to perform QC on counts slot (raw data) !!!

qc_ribo <- perCellQCMetrics(
  seu_Tcells_QC_Moderate@assays$RNA@counts,
  subsets = list(
    ribo = grep("^RPL|^RPS", rownames(seu_Tcells_QC_Moderate@assays$RNA@counts), value = TRUE)))

seu_Tcells_QC_Moderate <- AddMetaData(seu_Tcells_QC_Moderate, as.data.frame(qc_ribo))

qc_ribo2 <- data.frame(
  emulsion = seu_Tcells_QC_Moderate@meta.data$emulsion,
  subsets_mito_percent = seu_Tcells_QC_Moderate@meta.data$subsets_ribo_percent
)

qc_ribo2$high_ribo_outliers <- isOutlier(
  seu_Tcells_QC_Moderate@meta.data$subsets_ribo_percent,
  nmads = 3,
  type = "higher"
)

qc_ribo2$lower_ribo_outliers <- isOutlier(
  seu_Tcells_QC_Moderate@meta.data$subsets_ribo_percent,
  nmads = 3,
  type = "lower"
)

qc_ribo2 <- qc_ribo2 %>%
  dplyr::group_by(emulsion) %>%
  dplyr::mutate(
    ribo_status = dplyr::case_when(
      high_ribo_outliers == TRUE ~ "high",
      lower_ribo_outliers == TRUE ~ "low",
      TRUE ~ "normal"
    )
  ) %>%
  dplyr::ungroup() %>% 
  filter(!is.na(ribo_status))

seu_Tcells_QC_Moderate <- AddMetaData(seu_Tcells_QC_Moderate, qc_ribo2$ribo_status, col.name = "ribo_status")

```

##### Visualize {.tabset}

```{r QC Tcells Viz Ribo After QC 1}
ggplot(seu_Tcells_QC_Moderate@meta.data, aes(x = subsets_ribo_percent, y = emulsion, fill = after_stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01
  ) +
  scale_fill_viridis(
    alpha = 0.9,
    name = "subsets_ribo_percent",
    option = "C",
    guide = "none"
  ) +
  labs(
    title = "Ribosomal content by emulsion",
    x = "Percent ribosomal reads",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.spacing = unit(0.1, "lines"),
    strip.text.x = element_text(size = 8),
    plot.title = element_text(face = "bold")
  )

dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Ribosomal",
    "After_QC"),
  recursive = TRUE,
  showWarnings = FALSE)

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Ribosomal",
    "After_QC", 
    "Ridgeplot_ribosomal_content_by_emulsion_after_QC_All_Tcells.png"),
  width = 8,
  height = 5)
```

```{r QC Tcells Viz Ribo After QC 2}
FeaturePlot(
  seu_Tcells_QC_Moderate,
  features = "subsets_ribo_percent",
  reduction = "umap",
  order = TRUE,
  cols = c("lightgrey", "red"))

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Ribosomal",
    "After_QC",
    "UMAP_ribosomal_content_after_QC_All_Tcells.png"),
  width = 6,
  height = 5)
```

```{r QC Tcells Viz Ribo After QC 3}
ggplot(seu_Tcells_QC_Moderate@meta.data, aes(x = emulsion, y = subsets_ribo_percent, color = ribo_status)) +
  geom_violin(fill = "grey90", color = NA) +
  geom_quasirandom(alpha = 0.5, size = 0.8) +
  scale_y_log10() +
  scale_color_manual(
    values = c("normal" = "green3", "high" = "red3", "low" = "blue3")
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))


ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Ribosomal",
    "After_QC",
    "ViolinPlot_ribo_percent_by_emulsion_after_QC.png"),
  width = 12,
  height = 6)
```

#### Library size (nCountRNA) {.tabset}

##### Run {.tabset}

Replace manual MADS by function isOutlier

```{r QC Tcells Run Libsize After QC}
qc_lib_size <- seu_Tcells_QC_Moderate@meta.data %>%
  dplyr::group_by(emulsion) %>%
  dplyr::mutate(
    libsize_status = dplyr::case_when(
      nCount_RNA < median(nCount_RNA) - 3 * mad(nCount_RNA) ~ "low",
      nCount_RNA > median(nCount_RNA) + 3 * mad(nCount_RNA) ~ "high",
      TRUE ~ "normal"
    )
  ) %>%
  dplyr::ungroup() %>% 
  filter(nCount_RNA > 0)


# qc_lib_size <- seu_Tcells_QC_Moderate@meta.data %>%
#   group_by(emulsion) %>%
#   mutate(
#     libsize_status = dplyr::if_else(
#       isOutlier(nCount_RNA, nmads = 3, type = "both"),
#       "outlier",
#       "normal"
#     )
#   ) %>%
#   ungroup()


# Add libsize_status to metadata
seu_Tcells_QC_Moderate <- AddMetaData(seu_Tcells_QC_Moderate, qc_lib_size$libsize_status, col.name = "libsize_status")
```

##### Visualize {.tabset}

```{r QC Tcells Viz Libsize After QC1}
ggplot(qc_lib_size, aes(x = nCount_RNA)) +
  geom_density(fill = "grey80", alpha = 0.6) +
  geom_rug(
    aes(color = libsize_status),
    alpha = 0.6,
    sides = "b"
  ) +
  scale_x_log10() +
  scale_color_manual(
    values = c(
      low = "blue",
      normal = "green",
      high = "red"
    )
  ) +
  facet_wrap(~emulsion) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  ) +
  labs(
    color = "Library size QC",
    x = "nCount_RNA (log10)",
    y = "Density"
  )

dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC", 
    "Library_Size",
    "After_QC"),
  recursive = TRUE,
  showWarnings = FALSE)

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Library_Size",
    "After_QC",
    "DensityPlot_library_size_density_by_emulsion_after_QC.png"),
  width = 12,
  height = 6)
```

```{r QC Tcells Viz Libsize After QC 2}
ggplot(qc_lib_size, aes(x = emulsion, y = nCount_RNA, color = libsize_status)) +
  geom_violin(fill = "grey90", color = NA) +
  geom_quasirandom(alpha = 0.5, size = 0.8) +
  scale_y_log10() +
  scale_color_manual(
    values = c(low = "blue", normal = "green", high = "red")
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))


ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Library_Size",
    "After_QC",
    "ViolinPlot_library_size_by_emulsion_after_QC.png"),
  width = 12,
  height = 6)
```

#### Library diversity/heterogeneity (nFeatureRNA) {.tabset}

##### Run {.tabset}

Replace manual MADS by function isOutlier

```{r QC Tcells Run Libdiv After QC}
qc_lib_diversity <- seu_Tcells_QC_Moderate@meta.data %>%
  dplyr::group_by(emulsion) %>%
  dplyr::mutate(
    libdiv_status = dplyr::case_when(
      nFeature_RNA < median(nFeature_RNA) - 3 * mad(nFeature_RNA) ~ "low",
      nFeature_RNA > median(nFeature_RNA) + 3 * mad(nFeature_RNA) ~ "high",
      TRUE ~ "normal"
    )
  ) %>%
  dplyr::ungroup() %>% 
  filter(nFeature_RNA > 0)

# Add libdiv_status to metadata
seu_Tcells_QC_Moderate <- AddMetaData(seu_Tcells_QC_Moderate, qc_lib_diversity$libdiv_status, col.name = "libdiv_status")
```

##### Visualize {.tabset}

```{r QC Tcells Viz Libdiv After QC 1}
ggplot(qc_lib_diversity, aes(x = nFeature_RNA)) +
  geom_density(fill = "grey80", alpha = 0.6) +
  geom_rug(
    aes(color = libdiv_status),
    alpha = 0.6,
    sides = "b"
  ) +
  scale_x_log10() +
  scale_color_manual(
    values = c(
      low = "blue",
      normal = "green",
      high = "red"
    )
  ) +
  facet_wrap(~emulsion) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  ) +
  labs(
    color = "Library diversity QC",
    x = "nFeatureRNA (log10)",
    y = "Density"
  )

dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC", 
    "Library_Diversity",
    "After_QC"),
  recursive = TRUE,
  showWarnings = FALSE)

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Library_Diversity",
    "After_QC", 
    "DensityPlot_library_diversity_density_by_emulsion_after_QC.png"),
  width = 12,
  height = 6)
```

```{r QC Tcells Viz Libdiv After QC 2}
ggplot(qc_lib_diversity, aes(x = emulsion, y = nFeature_RNA, color = libdiv_status)) +
  geom_violin(fill = "grey90", color = NA) +
  geom_quasirandom(alpha = 0.5, size = 0.8) +
  scale_y_log10() +
  scale_color_manual(
    values = c(low = "blue", normal = "green", high = "red")
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Library_Diversity",
    "After_QC",
    "ViolinPlot_library_diversity_by_emulsion_after_QC.png"),
  width = 12,
  height = 6)
```

#### Librairy complexity {.tabset}

##### Run {.tabset}

```{r QC Tcells Run Libcomplex After QC}
seu_Tcells_QC_Moderate$log10_Complexity <- log10(seu_Tcells_QC_Moderate$nFeature_RNA) / log10(seu_Tcells_QC_Moderate$nCount_RNA)

qc_lib_complexity <- data.frame(
  emulsion = seu_Tcells_QC_Moderate$emulsion,
  log10_Complexity = seu_Tcells_QC_Moderate$log10_Complexity
)

qc_lib_complexity$low_complexity_outliers <- isOutlier(
  seu_Tcells_QC_Moderate$log10_Complexity,
  nmads = 3,
  type = "lower"
)

qc_lib_complexity$high_complexity_outliers <- isOutlier(
  seu_Tcells_QC_Moderate$log10_Complexity,
  nmads = 3,
  type = "higher"
)

qc_lib_complexity <- qc_lib_complexity %>%
  dplyr::group_by(emulsion) %>%
  dplyr::mutate(
    libcomplex_status = dplyr::case_when(
      low_complexity_outliers == TRUE ~ "low",
      high_complexity_outliers == TRUE ~ "high",
      TRUE ~ "normal"
    )
  ) %>%
  dplyr::ungroup() %>% 
  filter(!is.na(log10_Complexity))
  

low_quantil_complexity <- quantile(seu_Tcells_QC_Moderate$log10_Complexity, 0.05, na.rm = TRUE)

# Add libcomplex_status to metadata
seu_Tcells_QC_Moderate <- AddMetaData(seu_Tcells_QC_Moderate, qc_lib_complexity$libcomplex_status, col.name = "libcomplex_status")
```

##### Visualize {.tabset}

```{r QC Tcells Viz Libcomplex 1}
ggplot(seu_Tcells_QC_Moderate@meta.data, aes(x = log10_Complexity)) +
  geom_density(fill = "lightgreen", alpha = 0.6) +
  geom_vline(xintercept = low_quantil_complexity, linetype = "dashed", color = "red") +
  labs(
    title = "Library complexity (log10GenesPerUMI)",
    subtitle = paste0("Red dashed line = 5% quantile (", round(low_quantil_complexity, 2), ")"),
    x = "log10GenesPerUMI"
  ) +
  theme_classic()

dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC", 
    "Library_Complexity",
    "After_QC"),
  recursive = TRUE,
  showWarnings = FALSE)

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Library_Complexity",
    "After_QC", 
    "DensityPlot_library_complexity_after_QC_All_Tcells.png"),
  width = 12,
  height = 6)
```

```{r QC Tcells Viz Libcomplex After QC 2}
# ggplot(seu_Tcells@meta.data, aes(x = nCount_RNA, y = nFeature_RNA)) +
#   geom_point(
#     data = subset(seu_Tcells@meta.data, !low_complexity_outliers),
#     color = "grey70",
#     alpha = 0.25,
#     size = 0.4
#   ) +
#   geom_point(
#     data = subset(seu_Tcells@meta.data, low_complexity_outliers),
#     color = "blue3",
#     alpha = 0.7,
#     size = 0.6
#   ) +
#   geom_smooth(formula = y ~ x, method = "lm", color = "black", linewidth = 0.6) +
#   scale_x_log10() +
#   scale_y_log10() +
#   labs(
#     title = "Library complexity",
#     subtitle = "Genes detected vs UMIs (log–log)",
#     x = "UMIs per cell",
#     y = "Genes per cell"
#   ) +
#   theme_classic(base_size = 11)

ggplot(seu_Tcells_QC_Moderate@meta.data, aes(x = nCount_RNA, y = nFeature_RNA, color = libcomplex_status)) +
  geom_point(alpha = 0.6, size = 0.6) +
  geom_smooth(formula = y ~ x, method = "lm", color = "black", linewidth = 0.6) +
  scale_color_manual(values = c("normal" = "grey70", "low" = "blue3", "high" = "red3")) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = "Library complexity",
    subtitle = "Genes(nFeature)/UMIs(nCount) per cell",
    x = "UMIs per cell",
    y = "Genes per cell",
    color = "Complexity status"
  ) +
  theme_classic(base_size = 11)


ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Library_Complexity",
    "After_QC", 
    "ScatterPlot_library_complexity_after_QC_All_Tcells.png"),
  width = 12,
  height = 6)
```

```{r QC Tcells Viz Libcomplex After QC 3}
ggplot(seu_Tcells_QC_Moderate@meta.data,
       aes(x = log10_Complexity, fill = emulsion)) +
  geom_density(alpha = 0.4) +
  geom_vline(
    seu_Tcells_QC_Moderate@meta.data,
    xintercept = median(seu_Tcells_QC_Moderate@meta.data$log10_Complexity) - 3 * mad(seu_Tcells_QC_Moderate@meta.data$log10_Complexity),
    linetype = "dashed",
    color = "blue3"
  ) +
  geom_vline(
    seu_Tcells_QC_Moderate@meta.data,
    xintercept = median(seu_Tcells_QC_Moderate@meta.data$log10_Complexity) + 3 * mad(seu_Tcells_QC_Moderate@meta.data$log10_Complexity),
    linetype = "dashed",
    color = "red3"
  ) +
  facet_wrap(~ emulsion, ncol = 4) +
  labs(
    title = "Library complexity by emulsion",
    x = "log10GenesPerUMI"
  ) +
  theme_classic() +
  theme(legend.position = "none")

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Library_Complexity",
    "After_QC", 
    "DensityPlot_library_complexity_by_emulsion_after_QC_All_Tcells.png"),
  width = 12,
  height = 6)
```

```{r QC Tcells Viz Libcomplex After QC 4}
ggplot(seu_Tcells_QC_Moderate@meta.data,
       aes(x = emulsion, y = log10_Complexity, color = libcomplex_status)) +
  geom_violin(fill = "grey90", color = NA) +
  geom_quasirandom(alpha = 0.4, size = 0.6) +
  scale_color_manual(
    values = c(low = "blue", normal = "green", high = "red")
  ) +
  labs(
    title = "Library complexity per emulsion",
    y = "log10GenesPerUMI",
    x = NULL
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Library_Complexity",
    "After_QC", 
    "ViolinPlot_library_complexity_by_emulsion_after_QC_All_Tcells.png"),
  width = 12,
  height = 6)
```

```{r QC Tcells Viz Libcomplex After QC 5}
ggplot(seu_Tcells_QC_Moderate@meta.data,
       aes(x = log10_Complexity, fill = libcomplex_status)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("high" = "red3", "normal" = "green3", "low" = "blue3")) +
  theme_classic()

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Library_Complexity",
    "After_QC", 
    "DensityPlot_library_complexity_outliers_after_QC_All_Tcells.png"),
  width = 12,
  height = 6)
```

#### Cell Cycle Scoring {.tabset}

##### Run {.tabset}

```{r QC Tcells Run cellcycle After QC}
seu_Tcells_QC_Moderate <- CellCycleScoring(
  seu_Tcells_QC_Moderate,
  s.features = cc.genes$s.genes,
  g2m.features = cc.genes$g2m.genes,
  set.ident = FALSE
)
```

##### Visualize {.tabset}

```{r QC Tcells Viz cellcycle After QC}
DimPlot(seu_Tcells_QC_Moderate, group.by = "Phase") +
  ggtitle("Cell cycle phase before QC All Tcells") +
  facet_wrap(~Phase)

dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC", 
    "Cell_Cycle",
    "After_QC"),
  recursive = TRUE,
  showWarnings = FALSE)

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Cell_Cycle",
    "After_QC",
    "UMAP_Cell_Cycle_Phase_before_QC_All_Tcells.png"),
  width = 12,
  height = 6)
```

### Doublet detection {.tabset}

#### Run {.tabset}

```{r QC Tcells Run DblFinder}
sce <- as.SingleCellExperiment(seu_Tcells_QC_Moderate, assay = "RNA")
sce <- scDblFinder(sce, samples = colData(sce)$emulsion)

seu_Tcells_QC_Moderate$doublet_score <- sce$scDblFinder.score
seu_Tcells_QC_Moderate$doublet_class <- sce$scDblFinder.class
```

#### Visualize before remove {.tabset}

```{r QC Tcells Viz DblFinder Violin}
VlnPlot(
  seu_Tcells_QC_Moderate,
  features = "doublet_score",
  group.by = "emulsion",
  pt.size = 0
)

dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC", 
    "Doublet"),
  recursive = TRUE,
  showWarnings = FALSE)

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Doublet",
    "ViolinPlot_Doublet_Score_by_emulsion_All_Tcells.png"),
  width = 12,
  height = 6)

```

```{r QC Tcells Viz DblFinder Dimplot}
DimPlot(
  seu_Tcells_QC_Moderate,
  group.by = "doublet_class",
  cols = c("singlet" = "grey70", "doublet" = "red3")
)

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Doublet",
    "UMAP_Doublet_class_All_Tcells.png"),
  width = 6,
  height = 5)
```

#### Remove doublets {.tabset}

```{r}
seu_Tcells_QC_Moderate <- subset(seu_Tcells_QC_Moderate, subset = doublet_class == "singlet")
```

#### Visualize after remove {.tabset}

```{r QC Tcells Viz DblFinder Violin after remove}
VlnPlot(
  seu_Tcells_QC_Moderate,
  features = "doublet_score",
  group.by = "emulsion",
  pt.size = 0
)

dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC", 
    "Doublet",
    "After_QC"),
  recursive = TRUE,
  showWarnings = FALSE)

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Doublet",
    "After_QC",
    "ViolinPlot_Doublet_Score_by_emulsion_All_Tcells.png"),
  width = 12,
  height = 6)

```

```{r QC Tcells Viz DblFinder Dimplot after remove}
DimPlot(
  seu_Tcells_QC_Moderate,
  group.by = "doublet_class",
  cols = c("singlet" = "grey70", "doublet" = "red3")
)

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "QC",
    "Doublet",
    "After_QC",
    "UMAP_Doublet_class_All_Tcells.png"),
  width = 6,
  height = 5)
```

# Normalization {.tabset}

## RNA matrix {.tabset}

```{r Normalize RNA}
DefaultAssay(seu) <- "RNA"
seu <- NormalizeData(seu, assay = "RNA")
```

## ADT matrix {.tabset}

```{r Normalize ADT}
DefaultAssay(seu) <- "ADT"
seu <- NormalizeData(seu, normalization.method = "CLR", margin = 2, assay = "ADT")
```

# Most Variable Features {.tabset}

## RNA matrix {.tabset}

```{r MVP RNA}
DefaultAssay(seu) <- "RNA"
seu <- FindVariableFeatures(seu, assay = "RNA")
```

## ADT matrix {.tabset}

```{r MVP ADT}
DefaultAssay(seu) <- "ADT"
VariableFeatures(seu) <- rownames(seu[["ADT"]])
```

# Scaling {.tabset}

## RNA matrix {.tabset}

```{r Scale RNA}
DefaultAssay(seu) <- "RNA"
seu <- ScaleData(seu, 
                 assay = "RNA",
                 vars.to.regress = c("nCount_RNA", "percent.mt"),
                 features = VariableFeatures(seu))
```

## ADT matrix {.tabset}

```{r Scale ADT}
DefaultAssay(seu) <- "ADT"
seu <- ScaleData(seu, assay = "ADT")
```

# PCA {.tabset}

## RNA matrix {.tabset}

```{r PCA RNA}
DefaultAssay(seu) <- "RNA"
seu <- RunPCA(seu, reduction.name = "pca")
p_elbow_PCA_RNA <- ElbowPlot(seu, ndims = 50, reduction = "pca")


dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "PCA"),
  recursive = TRUE,
  showWarnings = FALSE
)

ggsave(
  plot = p_elbow_PCA_RNA,
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "PCA",
    "RNA_PCA_elbowplot.png"),
  width = 12,
  height = 6,
  dpi = 300)
```

## ADT matrix {.tabset}

```{r PCA ADT}
DefaultAssay(seu) <- "ADT"
seu <- RunPCA(seu, reduction.name = "apca")
p_Elbow_PCA_ADT <- ElbowPlot(seu, ndims = 13, reduction = "apca")

ggsave(
  plot = p_Elbow_PCA_ADT,
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "PCA",
    "ADT_PCA_elbowplot.png"),
  width = 12,
  height = 6,
  dpi = 300)
```

# Batch correction (Harmony) {.tabset}

## RNA matrix {.tabset}

```{r Harmony RNA}
DefaultAssay(seu) <- "RNA"
#batch correction with Harmony (RNA)
seu <- RunHarmony(seu, 
                  group.by.vars = "id", 
                  plot_convergence = TRUE)

# Save convergence plot
dir.create(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Harmony",
    "Convergence_Plot"),
  showWarnings = FALSE
)

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Harmony",
    "Convergence_Plot",
    "RNA_Harmony_convergence_plot.png"),
  width = 8,
  height = 6,
  dpi = 300
)
```

# Weighted Nearest Neighbor (WNN) graph {.tabset}

```{r WNN}
seu <- FindMultiModalNeighbors(seu,
                               reduction.list = list("harmony", "apca"),
                               dims.list = list(1:9, 1:13))
```

# UMAP {.tabset}

## Tuning {.tabset}

```{r Tuning UMAP}
# Dir  to save
dir.create(
  file.path(DIR_RESULT,
            "All_Tcells",
            "UMAP",
            "Tuning"),
  recursive = TRUE,
  showWarnings = FALSE)

params <- expand.grid(
  n.neighbors = c(40),
  min.dist    = c(0.04, 0.05, 0.06, 0.07, 0.08, 0.09),
  spread      = c(1.5, 2)
)

for (i in seq_len(nrow(params))) {

  md <- params$min.dist[i]
  sp <- params$spread[i]
  nn <- params$n.neighbors[i]

  red_name <- paste0("umap_md", md, "_sp", sp, "_nn", nn)

  seu <- RunUMAP(
    seu,
    nn.name     = "weighted.nn",
    min.dist    = md,
    spread      = sp,
    n.neighbors = nn,
    seed.use    = 42,
    reduction.name = red_name
  )

  p <- DimPlot(
    seu,
    reduction = red_name,
    group.by  = "predicted.celltype.l2") +
    ggtitle(paste("min.dist =", md,
                  "spread =", sp,
                  "n.neighbors =", nn)) +
    scale_color_manual(values = pal)

  ggsave(
    plot = p,
    filename = file.path(
      DIR_RESULT,
      "All_Tcells",
      "UMAP",
      "Tuning",
      paste0(red_name, ".png")
    ),
    width = 12,
    height = 6
  )
}

```

## Run {.tabset}

```{r Run UMAP}
seu <- RunUMAP(seu,
               nn.name = "weighted.nn",
               min.dist = 0.06,
               spread = 1.0,
               n.neighbors = 50,
               seed.use = 42)
```

## Visualization {.tabset}

```{r Viz UMAP}
# Visualization
metadata <- c("condition", "source", "id", "predicted.celltype.l1", "predicted.celltype.l2", "predicted.celltype.l3")


dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "UMAP",
    "Metadata"),
  recursive = TRUE,
  showWarnings = FALSE)

for (meta in metadata) {
  p <- DimPlot(seu, reduction = "umap", group.by = meta) + ggtitle(meta)
  
  print(p)
  
  # Nom de fichier
  file_name <- paste0(output_dir, "/UMAP_metadata_", meta, ".png")

  ggsave(file_name, plot = p, width = 6, height = 5)
  
  ggsave(
    filename = file.path(
      DIR_RESULT,
      "All_Tcells",
      "UMAP",
      "Metadata",
      paste0("UMAP_metadata_", meta, ".pdf")
    ),
    plot = p,
    width = 6,
    height = 5
  )
}
```

# Clustering {.tabset}

## Test several clustering resolutions {.tabset}

```{r FindClusters}
# # Clustering (commun RNA + ADT on WNN graph)
# Explore different resolutions
resolutions <- seq(0.01, 0.1, by = 0.01)
for (res in resolutions) {
  seu <- FindClusters(seu, graph.name = "wsnn", resolution = res, verbose = FALSE)
}
```

## Visualization {.tabset}

```{r Viz clusters (Clustree)}
clustree(seu, prefix = "wsnn_res.")

dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering"),
  recursive = TRUE,
  showWarnings = FALSE)

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering",
    "Clustree_WNN.pdf"),
  width = 10,
  height = 8
)
```

## Optimal resolution {.tabset}

```{r Choose opti res}
# Choose resolution 0.06
seu$seurat_clusters <- seu$wsnn_res.0.1

# Reorganize clusters names
seu$seurat_clusters <- factor(seu$seurat_clusters, 
                              levels = names(sort(table(seu$seurat_clusters), 
                                                  decreasing = TRUE)))
```

## Clusters on UMAP {.tabset}

```{r Clusters on UMAP}
DimPlot(seu, reduction = "umap", group.by = "seurat_clusters") + ggtitle("seurat_clusters")

# Save the plot
ggsave(
  plot = last_plot(),
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering",
    "UMAP_seurat_clusters.pdf"),
  width = 12,
  height = 6)

ggsave(
  plot = last_plot(),
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "UMAP",
    "Metadata",
    "UMAP_seurat_clusters.pdf"),
  width = 12,
  height = 6)

DimPlot(seu, reduction = "umap", group.by = "seurat_clusters") + ggtitle("seurat clusters") +
  facet_wrap(~seurat_clusters)

ggsave(
     plot = last_plot(),
     filename = file.path(
       DIR_RESULT,
       "All_Tcells",
       "Clustering",
       "UMAP_seurat_clusters_facet.pdf"),
     width = 12,
     height = 6)

ggsave(
  plot = last_plot(),
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "UMAP",
    "Metadata",
    "UMAP_seurat_clusters_facet.pdf"),
  width = 12,
  height = 6)
```

#### UMAP density per cluster {.tabset}

```{r UMAP density per cluster}
### UMAP density

df <- as.data.frame(
  Embeddings(seu_CD8_QC_Moderate, "umap")
)

colnames(df) <- c("UMAP_1", "UMAP_2")

# Ajouter metadata utile
df$condition    <- seu_CD8_QC_Moderate$condition
df$BCL6_status  <- seu_CD8_QC_Moderate$BCL6_status
df$seurat_cluster <- seu_CD8_QC_Moderate$seurat_clusters


p_umap_density_clusters <- ggplot(df, aes(UMAP_1, UMAP_2)) +
  # Points colorés par cluster
  geom_point(
    aes(color = seurat_cluster),
    size = 0.5,
    alpha = 0.6
  ) +
  scale_colour_manual(values = pal) +
  
  # Contours en arrière-plan (structure globale)
  stat_density_2d(
    bins = 14,
    contour_var = "density",
    color = "black",
    linewidth = 0.2,
    h = c(0.8, 0.8)
  ) +
  
  coord_equal() +
  labs(title = "UMAP density plot of CD8 T cells",
       x = "UMAP 1",
       y = "UMAP 2",
       color = "Cluster") +
  
  theme_classic(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 14, face = "bold"),    # taille + gras du titre
    legend.text  = element_text(size = 12)
  ) +
  
  # Agrandir les points dans la légende
  guides(color = guide_legend(
    override.aes = list(size = 4)  # taille des symboles dans la légende
  ))
p_umap_density_clusters

ggsave(
  plot = p_umap_density_clusters,
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering",
    "UMAP_seurat_clusters_density.png"),
  width = 12,
  height = 6)

ggsave(
  plot = p_umap_density_clusters,
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "UMAP",
    "Metadata",
    "UMAP_seurat_clusters_density.pdf"),
  width = 12,
  height = 6)
```

## Clusters quality {.tabset}

### QC metrics per cluster {.tabset}

```{r QC metrics per clusters}
VlnPlot(
  seu,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  group.by = "seurat_clusters",
  pt.size = 0,
  ncol = 3
)

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering",
    "Clusters_QC_metrics.pdf"),
  width = 12,
  height = 6)
```

nCount_RNA: Library size

nFeature_RNA: Library complexity

percent.mt: Mitochondrial content

### Libraries size per cluster {.tabset}

```{r Libsize per clusters}
# Plot number of cells per cluster with percentage labels of total cells
cell_counts <- seu@meta.data %>%
  count(seurat_clusters, name = "Cell_Count") %>%
  mutate(
    Percentage = round(100 * Cell_Count / sum(Cell_Count), 6),
    label = paste0(Percentage, "%")
  )


ggplot(cell_counts, aes(y = seurat_clusters, x = Cell_Count, fill = seurat_clusters)) +
  geom_col() +
  geom_text(aes(label = label), hjust = -0.1, size = 3) +
  theme_minimal() +
  labs(
    title = "Number (and %) of cells per cluster",
    x = "Number of cells",
    y = "Cluster"
  ) +
  theme(legend.position = "none")

# Save the plot
ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering",
    "Clusters_cell_counts.pdf"),
  width = 8,
  height = 6
)
```

## Cluster repartition per patients {.tabset}

```{r Clusters repartition per patients}
# Proportion of cells per cluster for each patient
cluster_patient <- seu@meta.data %>%
  group_by(id, seurat_clusters) %>%
  summarise(Cell_Count = n()) %>%
  group_by(id) %>%
  mutate(Percentage = Cell_Count / sum(Cell_Count) * 100)

ggplot(cluster_patient, aes(x = id, y = Percentage, fill = seurat_clusters)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Proportion of cells per cluster for each patient",
    x = "Patient ID",
    y = "Proportion of cells",
    fill = "Cluster"
  ) + 
  theme(
      axis.text.x = element_text(angle = 45, 
                                 hjust = 1, 
                                 vjust = 1))

# Save the plot
ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering",
    "Clusters_repartition_per_patient.pdf"),
  width = 12,
  height = 6
)
```

### Mitochondrial content per cluster {.tabset}

```{r Mito per clusters}
stats_mt <- seu@meta.data %>%
  group_by(seurat_clusters) %>%
  summarise(
    mean_pct_mt   = mean(percent.mt),
    median_pct_mt = median(percent.mt),
    sd_pct_mt     = sd(percent.mt)
  )

kable(stats_mt, format = "html") %>%
  kable_styling()

# Save the table
write.csv(stats_mt, 
          file = file.path(
            DIR_RESULT,
            "All_Tcells",
            "Clustering",
            "cluster_mitochondrial_content_stats.csv"),
          row.names = FALSE)
```

```{r Viz Mito per clusters}
ggplot(seu@meta.data,
       aes(
         x = percent.mt,
         y = seurat_clusters,
         fill = seurat_clusters
       )) +
  geom_density_ridges(
    alpha = 0.7,
    scale = 1.2,
    rel_min_height = 0.01,
    color = "white") +
  # theme_ridges() +
  # theme(
  #   legend.position = "none"
  # ) +
  # geom_vline(xintercept = 20, linetype = "dashed", color = "red") +
  labs(
    x = "% mitochondrial genes",
    y = "Clusters")

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering",
    "Clusters_mitochondrial_content_ridges.pdf"),
  width = 12,
  height = 6
)
```

Too few cells in clusters 10 to be seen

<!-- Mitcochondrial content on UMAP -->

<!-- ```{r} -->

<!-- FeaturePlot( -->

<!--   seu, -->

<!--   features = "percent.mt", -->

<!--   cols = c("lightgrey", "red"), -->

<!--   reduction = "umap") -->

<!-- ``` -->

## Remove bad clusters {.tabset}

```{r Remove bad clusters}
# Remove clusters 
seu <- subset(seu, 
              subset = !(seurat_clusters %in% c("8", "9","10")))

seu$seurat_clusters <- factor(seu$seurat_clusters)

```

```{r UMAP without bad clusters}
DimPlot(seu, reduction = "umap", group.by = "seurat_clusters") + ggtitle("seurat clusters after removing bad clusters")

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "UMAP",
    "Metadata",
    "UMAP_seurat_clusters_after_removing_bad_clusters.pdf"),
  width = 12,
  height = 6
) 

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering",
    "UMAP_seurat_clusters_after_removing_bad_clusters.pdf"),
  width = 12,
  height = 6
)

DimPlot(seu, reduction = "umap", group.by = "seurat_clusters") + ggtitle("seurat clusters after removing bad clusters") +
  facet_wrap(~seurat_clusters, ncol = (length(unique(seu$seurat_clusters))/2) )

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "UMAP",
    "Metadata",
    "UMAP_seurat_clusters_after_removing_bad_clusters_facet.pdf"),
  width = 12,
  height = 6
)

ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Clustering",
    "UMAP_seurat_clusters_after_removing_bad_clusters_facet.pdf"),
  width = 12,
  height = 6)
```

# Annotate clusters {.tabset}

## FeaturePlot {.tabset}

```{r FeaturePlot ADT}
dict <- list(
  "CD16-TotalSeqB" = "FCGR3A",
  "CD335-NKp46-TotalSeqB" = "NCR1",
  "CD197-CCR7-TotalSeqB" = "CCR7",
  "CD4-TotalSeqB" = "CD4",
  "CD8a-TotalSeqB" = "CD8A",
  "CD45RA-TotalSeqB" = "PTPRC",
  "CD62L-TotalSeqB" = "SELL",
  "CD127-IL-7Ra-TotalSeqB" = "IL7R",
  "CD25-TotalSeqB" = "IL2RA",
  "TIGIT-VSTM3-TotalSeqB" = "TIGIT",
  "CD226-DNAM-1-TotalSeqB" = "CD226",
  "CD38-TotalSeqB" = "CD38"
)

dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "FeaturePlot",
    "Prot_RNA_Clusters_FeaturePlot"),
  recursive = TRUE,
  showWarnings = FALSE
)

for (prot in names(dict)) {
  gene <- dict[[prot]]
  
  # ADT plot
  DefaultAssay(seu) <- "ADT"
  p1 <- FeaturePlot(seu, features = prot, reduction = "umap") + ggtitle(paste0(prot, " (ADT)"))
  
  # RNA plot
  DefaultAssay(seu) <- "RNA"
  # Vérifie que le gène existe dans RNA
  if (gene %in% rownames(seu[["RNA"]])) {
    p2 <- FeaturePlot(seu, features = gene, reduction = "umap") + ggtitle(paste0(gene, " (RNA)"))
  } else {
    p2 <- ggplot() + 
      ggtitle(paste0(gene, " (RNA) - Not found")) + 
      theme_void()
  }
  
  p3 <- DimPlot(seu, reduction = "umap", group.by = "seurat_clusters") + ggtitle("seurat_clusters")
  
  # Combine plots
  p <- (p1 | p2) / p3
  p


  
  # Sauvegarde
  ggsave(
    filename = file.path(
      DIR_RESULT,
      "All_Tcells",
      "FeaturePlot",
      "Prot_RNA_Clusters_FeaturePlot",
      paste0("Prot_RNA_Clusters_FeaturePlot_", prot, "_", gene, "_featureplot.png")),
    plot = p,
    width = 12,
    height = 12)
}
```

# Clusters markers {.tabset}

## Find all markers {.tabset}

```{r FindAllMarkers}
markers <- FindAllMarkers(seu, only.pos = TRUE, min.pct = 0.25)

markers_filt <- markers %>%
  filter(
    p_val_adj < 0.05,
    avg_log2FC > 0.5,
    !grepl("^MT-|^RPL|^RPS", gene)
  )
```

## Top 10 markers per cluster {.tabset}

```{r Top10 markers}
genes.scaled <- rownames(seu[["RNA"]]@scale.data)

top10 <- markers_filt %>%
  filter(gene %in% genes.scaled) %>%
  group_by(cluster) %>%
  slice_max(avg_log2FC, n = 10)

top10 %>% arrange(cluster, desc(avg_log2FC))
```

##### FeaturePlot of top10 markers per cluster {.tabset}

```{r All FeaturePlot Top10}
dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "FeaturePlot",
    "Top10_Markers_Per_Cluster"),
  recursive = TRUE,
  showWarnings = FALSE)

for (i in unique(top10$cluster)) {
  
  dir.create(
    file.path(
      DIR_RESULT,
      "All_Tcells",
      "FeaturePlot",
      "Top10_Markers_Per_Cluster",
      paste0("Cluster_", i)),
    recursive = TRUE,
    showWarnings = FALSE
  )

  genes_cluster <- top10 %>%
    filter(cluster == i) %>%
    pull(gene)

  p_list <- lapply(genes_cluster, function(gene) {
    FeaturePlot(
      seu,
      features = gene, 
      reduction = "umap") +
      # scale_color_gradient(low = "lightgrey", high = "red") +
      ggtitle(paste0("Cluster ", i, ": ", gene))
  })

  p_combined <- wrap_plots(p_list, ncol = 5)
  p_combined
  
  ggsave(
    filename = file.path(
      DIR_RESULT,
      "All_Tcells",
      "FeaturePlot",
      "Top10_Markers_Per_Cluster",
      paste0("Cluster_", i),
      paste0("Cluster_", i, "_Top10_Markers_featureplot.png")),
    plot = p_combined,
    width = 20,
    height = 12)
  
  lapply(genes_cluster, function(gene) {
    
    dir.create(
      file.path(
        DIR_RESULT,
        "All_Tcells",
        "FeaturePlot",
        "Top10_Markers_Per_Cluster",
        paste0("Cluster_", i),
        paste0("Gene_", gene)),
      recursive = TRUE,
      showWarnings = FALSE)
      
    violin <- VlnPlot(
      seu, 
      cols = pal,
      features = gene, 
      group.by = "seurat_clusters", 
      pt.size = 0) +
      ggtitle(paste0("Cluster ", i, ": ", gene))
    violin
      
    ggsave(
      plot = violin,
      filename = file.path(
        DIR_RESULT,
        "All_Tcells",
        "FeaturePlot",
        "Top10_Markers_Per_Cluster",
        paste0("Cluster_", i),
        paste0("Gene_", gene),
        paste0("Cluster_", i, "_", gene, "_VlnPlot.png")
      ),
      width = 6,
      height = 4)
      
    feature <- FeaturePlot(
      seu, 
      features = gene, 
      reduction = "umap") +
      ggtitle(paste0("Cluster ", i, ": ", gene))
    feature
    
    ggsave(
      filename = file.path(
        DIR_RESULT,
        "All_Tcells",
        "FeaturePlot",
        "Top10_Markers_Per_Cluster",
        paste0("Cluster_", i),
        paste0("Gene_", gene),
        paste0("Cluster_", i, "_", gene, "_FeaturePlot.png")
      ),
      plot = feature,
      width = 6,
      height = 5)
      
    p_cluster <- DimPlot(
      seu, 
      reduction = "umap", 
      group.by = "seurat_clusters") +
      scale_color_manual(values = pal)
    p_cluster
      
    p_violin_feature_cluster <- (violin | feature) / p_cluster
    p_violin_feature_cluster
      
    ggsave(
      filename = file.path(
        DIR_RESULT,
        "All_Tcells",
        "FeaturePlot",
        "Top10_Markers_Per_Cluster",
        paste0("Cluster_", i),
        paste0("Gene_", gene),
        paste0("Cluster_", i, "_", gene, "_Vln_Feature_Cluster.png")
      ),
      plot = p_violin_feature_cluster,
      width = 12,
      height = 10)
  })
}
```

## Heatmap of markers per cluster {.tabset}

```{r Heatmap Top10}
genes <- unique(top10$gene)

# Extract scaled data for the selected genes
mat <- GetAssayData(
  seu,
  assay = "RNA",
  layer = "scale.data"
)[genes, ]

# Order cells by cluster
clusters <- seu$seurat_clusters
ord <- order(clusters)

mat <- mat[, ord]
clusters <- clusters[ord]

# Create annotation for clusters
cluster_ha <- HeatmapAnnotation(
  Cluster = clusters,
  col = list(
    Cluster = structure(
      Seurat::DiscretePalette(length(unique(clusters)), "polychrome"),
      names = levels(clusters)
    )
  )
)
# Create heatmap
heatmap <- Heatmap(
  mat,
  name = "Expression",
  top_annotation = cluster_ha,
  show_column_names = FALSE,
  show_row_names = TRUE,
  cluster_columns = FALSE,
  cluster_rows = TRUE,
  row_names_gp = grid::gpar(fontsize = 8),
  column_title = "Top 10 markers per cluster"
  )


# Save heatmap
dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "Heatmap"),
  recursive = TRUE,
  showWarnings = FALSE)
  
draw(heatmap)
png(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Heatmap",
    "Top10_Markers_Per_Cluster_Heatmap.png"),
  width = 1200,
  height = 800
)



```

# Automatic annotation (SingleR) {.tabset}

## Make automatic predictions {.tabset}

```{r Run auto annot (SingleR)}
# Prepare SingleR input
sce <- as.SingleCellExperiment(seu, assay = "RNA")

# Reference dataset
ref <- fetchReference("monaco_immune", "2024-02-26")

# Run SingleR
predictions <- SingleR(
  test = sce,
  ref = ref,
  labels = ref$label.fine,
  assay.type.test = "logcounts")

# Add predictions to Seurat object
seu$predicted.celltype.l3_jordan <- predictions$pruned.labels
```

### Visualization of automatic annotations {.tabset}

```{r Viz auto Annot}

dir.create(
  file.path(
    DIR_RESULT,
    "All_Tcells",
    "Annotation",
    "Automatic"),
  recursive = TRUE,
  showWarnings = FALSE)

p1 <- DimPlot(seu, reduction = "umap", group.by = "predicted.celltype.l3_jordan") + ggtitle("predicted.celltype.l1")
p1
ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Annotation",
    "Automatic",
    "UMAP_predicted_celltype_JD.png"),
  plot = p1,
  width = 12,
  height = 6
)

p2 <- DimPlot(seu, reduction = "umap", group.by = "predicted.celltype.l3_jordan") + ggtitle("predicted.celltype.l3_jordan") + 
  facet_wrap(~predicted.celltype.l3_jordan, ncol = 6)
p2
ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Annotation",
    "Automatic",
    "UMAP_predicted_celltype_JD_facet.png"),
  plot = p2,
  width = 12,
  height = 12
)

p3 <- p1 | p2
p3
ggsave(
  filename = file.path(
    DIR_RESULT,
    "All_Tcells",
    "Annotation",
    "Automatic",
    "UMAP_predicted_celltype_JD_combined.pdf"),
  plot = p3,
  width = 12,
  height = 6
)
```

# Manual annotation {.tabset}

## Create my celltype signatures {.tabset}

### Run presto to find markers {.tabset}

```{r Run presto}
expr_matrix <- as.matrix(seu@assays$RNA@data)
labels <- seu$seurat_clusters

presto_markers <- wilcoxauc(expr_matrix, labels)
```

### Construction of the signature table {.tabset}

```{r Construct signature tibble}
signature_tbl <- tribble(
  ~feature, ~cell_type, ~class,
  
  # Activated
  "CD69", "Activated", "State",
  "TNFRSF9", "Activated", "State",
  "IL2RA", "Activated", "State",

  # Apoptotic / stress
  "MALAT1", "Apoptotic", "State",
  "NEAT1", "Apoptotic", "State",

  # CD4 lineage
  "CD3D", "CD4", "Identity",
  "CD3E", "CD4", "Identity",
  "CD4",  "CD4", "Identity",
  "TRBC1","CD4", "Identity",

  # CD8 lineage
  "FCGR3A","CD8", "Identity",
  "CD3D", "CD8", "Identity",
  "CD3E", "CD8", "Identity",
  "CD8A", "CD8", "Identity",
  "CD8B", "CD8", "Identity",
  "TRBC1","CD8", "Identity",
  "NCR1", "CD8", "Identity",

  # Cytotoxic
  "PRF1","Cytotoxic", "Function",
  "GZMB","Cytotoxic", "Function",
  "GNLY","Cytotoxic", "Function",
  "GZMH","Cytotoxic", "Function",
  "CCL5","Cytotoxic", "Function",
  "NKG7","Cytotoxic", "Function",
  "KLRG1","Cytotoxic", "Function",
  "GZMA","Cytotoxic", "Function",

  # Exhausted
  "PDCD1","Exhausted", "State",
  "LAG3","Exhausted", "State",
  "TOX","Exhausted", "State",
  "TIGIT","Exhausted", "State",
  "CD226","Exhausted", "State",
  "GZMK","Exhausted", "State",
  "HAVCR2","Exhausted", "State",
  "XCL1","Exhausted", "State",
  "XCL2","Exhausted", "State",
  "CD160","Exhausted", "State",

  # HLA
  "HLA-DRA","HLA", "Function",
  "HLA-DRB1","HLA", "Function",

  # IFN / ISG
  "IFNG","IFN", "Function",
  "IFIT2","IFN", "Function",
  "IFIT3","IFN", "Function",
  "ISG15","IFN", "Function",
  "STAT1","IFN", "Function",
  "IFI6","IFN", "Function",
  "IRF7","IFN", "Function",
  "MX1","IFN", "Function",
  "IFITM3","IFN", "Function",

  # Inflammatory / NFkB
  "TNFAIP3","Inflammatory", "Function",
  "NFKB1","Inflammatory", "Function",
  "NFKBIA","Inflammatory", "Function",
  "NFKBIZ","Inflammatory", "Function",
  "STAT4","Inflammatory", "Function",

  # Memory
  "CD44","Memory", "State",
  "PTPRC","Memory", "State",
  "CCR7","Memory", "State",
  "CD27","Memory", "State",
  "CD28","Memory", "State",

  # Naive
  "CCR7","Naif", "State",
  "SELL","Naif", "State",
  "CD27","Naif", "State",
  "CD28","Naif", "State",
  "TCF7","Naif", "State",
  "IL7R","Naif", "State",
  "LEF1","Naif", "State",

  # Proliferating
  "CENPV","Prolif", "State",
  "MKI67","Prolif", "State",
  "NCAPG","Prolif", "State",
  "KIFC1","Prolif", "State",
  "CENPE","Prolif", "State",
  "ORC6","Prolif", "State",

  # Resident (Trm)
  "CD69","Resident", "State",
  "CXCR6","Resident", "State",
  "NR4A2","Resident", "State",
  "RUNX3","Resident", "State",

  # Senescent
  "KLRG1","Senescent", "State",
  "B3GAT1","Senescent", "State",
  "CD27","Senescent", "State",
  "CD28","Senescent", "State",
  "LAMP1","Senescent", "State",
  "IL1A","Senescent", "State",
  "FAS","Senescent", "State",
  "TP53","Senescent", "State",
  "TP53BP1","Senescent", "State",
  "CDKN1A","Senescent", "State",
  "CDKN2A","Senescent", "State",
  "MKI67","Senescent", "State",

  # Tfh
  "IL21","Tfh", "Identity",
  # "FR4","Tfh", "Identity", # Only in mouse
  "CXCR5","Tfh", "Identity",
  "BCL6","Tfh", "Identity",
  "PDCD1","Tfh", "Identity",
  "ICOS","Tfh", "Identity",

  # Th1
  "CX3CR1","Th1", "Identity",
  "IFNG","Th1", "Identity",
  "TBX21","Th1", "Identity",
  "CXCR3","Th1", "Identity",
  "STAT4","Th1", "Identity",
  "IL12RB2","Th1", "Identity",

  # Th2
  "GATA3","Th2", "Identity",
  "CCR4","Th2", "Identity",
  "IL4","Th2", "Identity",
  "IL5","Th2", "Identity",
  "IL13","Th2", "Identity",

  # Th17
  "RORC","Th17", "Identity",
  "IL17A","Th17", "Identity",
  "CCR6","Th17", "Identity",
  "IL17F","Th17", "Identity",
  "IL23R","Th17", "Identity",

  # Treg
  "FOXP3","Treg", "Identity",
  "IL2RA","Treg", "Identity",
  "CTLA4","Treg", "Identity",
  "TIGIT","Treg", "Identity",
  "IKZF2","Treg", "Identity",
  "CD27","Treg", "Identity",
  "IL7R","Treg", "Identity"
)
```

### Visualization of all signatures together {.tabset}

```{r}
Tcell_signature <- "
^CD3D$|^CD3E$|^CD4$|^TRBC1$|
^FCGR3A$|^CD8A$|^CD8B$|^NCR1$|
^CX3CR1$|^IFNG$|^TBX21$|^CXCR3$|^STAT4$|^IL12RB2$|
^GATA3$|^CCR4$|^IL4$|^IL5$|^IL13$|
^RORC$|^IL17A$|^CCR6$|^IL17F$|^IL23R$|
^IL21$|^CXCR5$|^BCL6$|^PDCD1$|^ICOS$|
^FOXP3$|^IL2RA$|^CTLA4$|^TIGIT$|^IKZF2$|^CD27$|^IL7R$|
^PRF1$|^GZMB$|^GNLY$|^GZMH$|^CCL5$|^NKG7$|^KLRG1$|^GZMA$|
^LAG3$|^TOX$|^CD226$|^GZMK$|^HAVCR2$|^XCL1$|^XCL2$|^CD160$|
^B3GAT1$|^CD28$|^LAMP1$|^IL1A$|^FAS$|^TP53$|^TP53BP1$|^CDKN1A$|^CDKN2A$|^MKI67$|
^CD69$|^CXCR6$|^NR4A2$|^RUNX3$|^TNFRSF9$|
^CENPV$|^NCAPG$|^KIFC1$|^CENPE$|^ORC6$|
^CCR7$|^SELL$|^TCF7$|^LEF1$|^Tcells4$|^PTPRC$|
^MALAT1$|^NEAT1$|
^IFIT2$|^IFIT3$|^ISG15$|^STAT1$|^IFI6$|^IRF7$|^MX1$|^IFITM3$|
^TNFAIP3$|^NFKB1$|^NFKBIA$|^NFKBIZ$|
^HLA-DRA$|^HLA-DRB1$
"


selected_features <- grep(Tcell_signature, rownames(sce), value=TRUE)


m2 <- presto_markers %>%
  filter(feature %in% selected_features) %>%
  select(feature, group, avgExpr, pct_in) %>%
  left_join(signature_tbl, by = "feature") %>%
  mutate(signature_class = case_when(
  cell_type %in% c("CD4","CD8","Th1","Th2","Th17","Tfh","Treg") ~ "Identity",
  cell_type %in% c("Cytotoxic","IFN","Inflammatory","HLA") ~ "Function",
  cell_type %in% c("Naif","Memory","Activated","Exhausted",
                    "Resident","Prolif","Senescent","Apoptotic") ~ "State"
)) %>%
  filter(!is.na(cell_type)) %>%
  group_by(feature) %>%
  mutate(
    avgExpr = avgExpr - mean(avgExpr),
    avgExpr = scales::rescale(avgExpr, to = c(-2,2))
  ) %>%
  ungroup()

ggplot(m2, 
       aes(x = feature,
               y = fct_reorder(as.character(group), group),
               color = avgExpr,
               size = pct_in)) +
  geom_point() +
  scale_color_gradient2(low="blue", mid="white", high="red") +
  facet_wrap(~cell_type, scales="free", ncol=10) +
  theme(
    axis.text.x = element_text(angle=45, hjust=1),
    axis.text.y = element_text(angle=45, hjust=1)
  ) +
  labs(
    title = "LT Compartment - All Signatures",
    x = "Genes markers",
    y = "Clusters"
  )

dir.create(
   file.path(
     DIR_RESULT,
     "All_Tcells",
     "Annotation",
     "Manual"),
   recursive = TRUE,
   showWarnings = FALSE)

ggsave(
       filename = file.path(
         DIR_RESULT,
         "All_Tcells",
         "Annotation",
         "Manual",
         "Marker_Signatures_LT_Compartment_All_Signatures.png"),
       width = 20, height = 10)
)
```

### View of specific signatures {.tabset}

```{r Viz specific signature}
m3 <- presto_markers %>%
  filter(feature %in% selected_features) %>%
  select(feature, group, avgExpr, pct_in) %>%
  left_join(signature_tbl, by = "feature") %>%
  filter(!is.na(cell_type)) %>%
  group_by(feature) %>%
  mutate(
    avgExpr = avgExpr - mean(avgExpr),
    avgExpr = scales::rescale(avgExpr, to = c(-2,2))
  ) %>%
  ungroup()

for (i in unique(m3$class)) {
  p <- ggplot(m3 %>% filter(class == i), 
         aes(x=feature, y=fct_reorder(as.character(group), group), color = avgExpr, size = pct_in)) +
    geom_point()+
    scale_color_gradient2(low = "blue", mid="white",high = "red")+
    # theme_minimal() +
    facet_wrap(~ cell_type, scales = "free", ncol = 4) +
    theme(axis.text.y = element_text(angle=45, hjust=1), 
          axis.text.x = element_text(angle=45, hjust=1)) +
    labs(
    title = paste0("LT Compartment - ", i),
    x = "Genes markers",
    y = "Clusters"
  )
  
  print(p)
  
  dir.create(
    file.path(
      DIR_RESULT,
      "All_Tcells",
      "Annotation",
      "Manual"),
    recursive = TRUE,
    showWarnings = FALSE
  )
  
  ggsave(
    filename = file.path(
      DIR_RESULT,
      "All_Tcells",
      "Annotation",
      "Manual",
      paste0("Marker_Signatures_LT_Compartment_", i, ".png")),
    plot = p,
    width = 20, height = 10
  )
}
```

For now I think that we don't need to be to specific on pop annotation because we have all Tcells. For now we have to identified major populations (CD4, CD8) by combining annotation from me + automatic + from PPA to be sure that we analyse the right populations later.

And then we extract count for specific pop cd8 and cd4 cells and perform a new clustering/annotation on each of them separately to have more precise subpopulations.

```{r Combine annot}
cluster_annotation_list <- list(
  "0" = "CD4_naive_and_memory",
  "1" = "CD8_Cytotox_exhausted",
  "2" = "CD4_inflammed",
  "3" = "CD8_cytotox_low_resident_inflammed",
  "4" = "CD8_naive",
  "5" = "???",
  "6" = "Dying_Tcells",
  "7" = "CD8_IFN"
)

# Add annotations to Seurat object
seu$manual_annotation <- plyr::mapvalues(
  seu$seurat_clusters,
  from = names(cluster_annotation_list),
  to = unlist(cluster_annotation_list)
)

# Names of cells with matching annotations between automatic (mine and PPA SingleR predictions) and manual (mine) annotation
## But first let's harmonize the names between predicted.celltype.l3_jordan and predicted.celltype.l3
seu$predicted.celltype.l3_jordan <- plyr::mapvalues(
  seu$predicted.celltype.l3_jordan,
  from = c("Non-Vd2 gd T cells", "Naive CD4 T cells", "Follicular helper T cells", "Terminal effector CD8 T cells", "Vd2 gd T cells", "MAIT cells", "Terminal effector CD4 T cells", "Th1/Th17 cells", "Th17 cells", "Th1 cells", "Th2 cells", "Effector memory CD8 T cells", "T regulatory cells", "Central memory CD8 T cells", "Naive CD8 T cells", "Natural killer cells", "Myeloid dendritic cells", "Progenitor cells"),
  to = c("Non-Vd2 gd T cells", "Naive CD4 T cells", "Follicular helper T cells CD4", "Terminal effector CD8 T cells", "Vd2 gd T cells", "MAIT cells", "Terminal effector CD4 T cells", "Th1/Th17 cells CD4", "Th17 cells CD4", "Th1 cells CD4", "Th2 cells CD4", "Effector memory CD8 T cells", "T regulatory cells CD4", "Central memory CD8 T cells", "Naive CD8 T cells", "Natural killer cells", "Myeloid dendritic cells", "Progenitor cells")
)



matched_cells <- WhichCells(
  seu,
  expression =
    (
      grepl("CD4", predicted.celltype.l3_jordan) &
      grepl("CD4", predicted.celltype.l1) &
      grepl("CD4", manual_annotation)
    ) |
    (
      grepl("CD8", predicted.celltype.l3_jordan) &
      grepl("CD8", predicted.celltype.l1) &
      grepl("CD8", manual_annotation)
    )
)
length(matched_cells)
length(seu@meta.data$manual_annotation)
# Le nombre de cellules avec annotation "Natural killer cells" dans predicted.celltype.l3_jordan
length(
  predicted_cells <- WhichCells(
    seu,
    expression = (predicted.celltype.l1 == "other T")
))

```

There is a lot of cells annotate gdTcells but I will not focus on them for now as I want to focus on CD4 and CD8 Tcells, and there are too different from cd4+ and cd8+.

# Session Info {.tabset .unnumbered}

```{r Session Info}
sessionInfo()
```

::: callout-note
This is a note.
:::

::: callout-tip
This is a tip.
:::

::: callout-warning
This is a warning.
:::

::: callout-important
Important message.
:::

::: callout-caution
Caution!
:::

::: {.callout-tip title="Pro Tip" icon="lightbulb" collapse="true"}
Use `SCTransform` instead of log-normalization for variance stabilization.
:::

::: {.panel-tabset .panel-tabset-pills}
## SCTransform

Plot A

## Log-normalization

Plot B
:::

::: my-highlight
Some highlighted text.
:::
