---
title: "HH_vs_MM_scRNAseq_reanalysis"
author: "Jordan Dutel"
data: today
format:
  html:
    toc: true # table of contents
    toc-depth: 3
    number-sections: true
    theme: cosmo # theme Bootstrap
    code-fold: true # hide code by default
    code-tools: true # copy/display code button
    code-summary: "Show code"
    embed-resources: true # autonomous HTML
    toc-location: left

editor: visual
lang: en
title-block-banner: true
---

```{r setup, echo=FALSE}
############
# PACKAGES #
############

# BiocManager::install("SingleCellExperiment")
# install.packages("devtools")
# BiocManager::install("DropletUtils")
# devtools::install_github("immunogenomics/presto")
# BiocManager::install("scDblFinder")

library(DropletUtils)
library(SingleCellExperiment)
library(scater)
library(scran)
library(qs)
library(Seurat)
library(ggplot2)
library(patchwork)
library(clustree)
library(harmony)
library(patchwork)
library(dplyr)
library(ggridges)
library(readr)
library(knitr)
library(kableExtra)
library(SingleR)
library(celldex)
library(ComplexHeatmap)
library(circlize)
library(presto)
library(tidyverse)
library(tibble)
library(matrixStats)
library(viridis)
library(scuttle)
library(scDblFinder)
library(ggbeeswarm)





#########
# PATHS #
#########

knitr::opts_knit$set(
  root.dir = "~/Documents/Project/Thomas/20260116_scRNAseq_HD_MM_reanalysis")

seurat_path_Tcell_obj <- "/mnt/crct-share/crct13/PPA/PPA_work/project/01_HCvsMM/01A_HCvsMM_10x/results/Tcells_analysis/seurat_objects/QCed_seurat_Tcells_clean.qs"





##########
# COLORS #
##########

pal <-c('#e6194B', '#3cb44b', '#ffe119', '#4363d8', '#f58231', '#911eb4', '#42d4f4', '#f032e6', '#bfef45', '#fabed4', '#9A6324', '#469990', '#dcbeff', '#fffac8', '#800000', '#aaffc3', '#808000', '#ffd8b1', '#000075', '#a9a9a9', '#ffffff', '#000000')

# pal2 <- Seurat::DiscretePalette(length(unique(seu_CD4@meta.data$emulsion)))
```

# Load data {.tabset}

```{r}
seu <- qread(seurat_path_Tcell_obj)
```

# Normalization {.tabset}

## RNA matrix {.tabset}

```{r}
DefaultAssay(seu) <- "RNA"
seu <- NormalizeData(seu, assay = "RNA")
```

## ADT matrix {.tabset}

```{r}
DefaultAssay(seu) <- "ADT"
seu <- NormalizeData(seu, normalization.method = "CLR", margin = 2, assay = "ADT")
```

# Most Variable Features {.tabset}

## RNA matrix {.tabset}

```{r}
DefaultAssay(seu) <- "RNA"
seu <- FindVariableFeatures(seu, assay = "RNA")
```

## ADT matrix {.tabset}

```{r}
DefaultAssay(seu) <- "ADT"
VariableFeatures(seu) <- rownames(seu[["ADT"]])
```

# Scaling {.tabset}

## RNA matrix {.tabset}

```{r}
DefaultAssay(seu) <- "RNA"
seu <- ScaleData(seu, 
                 assay = "RNA",
                 vars.to.regress = c("nCount_RNA", "percent.mt"),
                 features = VariableFeatures(seu))
```

## ADT matrix {.tabset}

```{r}
DefaultAssay(seu) <- "ADT"
seu <- ScaleData(seu, assay = "ADT")
```

# PCA {.tabset}

## RNA matrix {.tabset}

```{r}
DefaultAssay(seu) <- "RNA"
seu <- RunPCA(seu, reduction.name = "pca")
ElbowPlot(seu, ndims = 50, reduction = "pca")
dir.create("03_result/PCA", showWarnings = FALSE)
ggsave("03_result/PCA/RNA_PCA_elbowplot.png")
```

## ADT matrix {.tabset}

```{r}
DefaultAssay(seu) <- "ADT"
seu <- RunPCA(seu, reduction.name = "apca")
ElbowPlot(seu, ndims = 13, reduction = "apca")
dir.create("03_result/PCA", showWarnings = FALSE)
ggsave("03_result/PCA/ADT_PCA_elbowplot.png")
```

# Batch correction (Harmony) {.tabset}

## RNA matrix {.tabset}

```{r}
DefaultAssay(seu) <- "RNA"
#batch correction with Harmony (RNA)
seu <- RunHarmony(seu, 
                  group.by.vars = "id", 
                  plot_convergence = TRUE)

# Save convergence plot
output_dir <- "03_result/Harmony"
dir.create(output_dir, showWarnings = FALSE)
ggsave(paste0(output_dir, "/RNA_Harmony_convergence_plot.png"))
```

# Weighted Nearest Neighbor (WNN) graph {.tabset}

```{r}
seu <- FindMultiModalNeighbors(seu,
                               reduction.list = list("harmony", "apca"),
                               dims.list = list(1:9, 1:13))
```

# UMAP {.tabset}

## Tuning {.tabset}

```{r}
# # Paramètres à tester
# min.dists <- c(0.04, 0.06, 0.08, 0.1, 0.12, 0.14, 0.16, 0.18, 0.2)
# spreads <- c(0.5, 1.0, 1.5, 2.0)
# n.neighbors <- c(40, 50, 60)
# # Dossier pour sauvegarder
# output_dir <- "../03_result/UMAP/Tuning"
# dir.create(output_dir, showWarnings = FALSE)
# 
# for (md in min.dists) {
#   for (sp in spreads) {
#     for (nn in n.neighbors) {
#       seu <- RunUMAP(seu,
#                      nn.name = "weighted.nn",
#                      min.dist = md,
#                      spread = sp,
#                      n.neighbors = 60,
#                      seed.use = 42,
#                      reduction.name = paste0("umap_md", md, "_sp", sp, "_nn", nn))
# 
#       p <- DimPlot(seu,
#                    reduction = paste0("umap_md", md, "_sp", sp, "_nn", nn),
#                    group.by = "predicted.celltype.l1") +
#         ggtitle(paste("min.dist =", md, "spread =", sp, "n.neighbors =", nn))
# 
#       # Nom de fichier
#       file_name <- paste0(output_dir, "/UMAP_md", md, "_sp", sp, "_nn", nn, ".png")
# 
#       ggsave(file_name, plot = p, width = 6, height = 5)
#     }
#   }
# }
```

## Run {.tabset}

```{r}
seu <- RunUMAP(seu,
               nn.name = "weighted.nn",
               min.dist = 0.06,
               spread = 1.0,
               n.neighbors = 50,
               seed.use = 42)
```

## Visualization {.tabset}

```{r}
# Visualization
metadata <- c("condition", "source", "id", "predicted.celltype.l1", "predicted.celltype.l2", "predicted.celltype.l3")

# Dossier pour sauvegarder
output_dir <- "03_result/UMAP/Metadata"
dir.create(output_dir, showWarnings = FALSE)

for (meta in metadata) {
  p <- DimPlot(seu, reduction = "umap", group.by = meta) + ggtitle(meta)
  
  print(p)
  
  # Nom de fichier
  file_name <- paste0(output_dir, "/UMAP_metadata_", meta, ".png")

  ggsave(file_name, plot = p, width = 6, height = 5)
}
```

# Clustering {.tabset}

## Test several clustering resolutions {.tabset}

```{r}
# # Clustering (commun RNA + ADT on WNN graph)
# Explore different resolutions
resolutions <- seq(0.01, 0.1, by = 0.01)
for (res in resolutions) {
  seu <- FindClusters(seu, graph.name = "wsnn", resolution = res, verbose = FALSE)
}
```

## Visualization {.tabset}

```{r}
clustree(seu, prefix = "wsnn_res.")

# Save the plot
dir.create("03_result/Clustering", showWarnings = FALSE)
ggsave("03_result/Clustering/Clustree_WNN.png", width = 10, height = 8)
```

## Optimal resolution {.tabset}

```{r}
# Choose resolution 0.06
seu$seurat_clusters <- seu$wsnn_res.0.1

# Reorganize clusters names
seu$seurat_clusters <- factor(seu$seurat_clusters, 
                              levels = names(sort(table(seu$seurat_clusters), 
                                                  decreasing = TRUE)))
```

## Clusters on UMAP {.tabset}

```{r}
DimPlot(seu, reduction = "umap", group.by = "seurat_clusters") + ggtitle("seurat_clusters")

# Save the plot
ggsave("03_result/Clustering/UMAP_seurat_clusters.png")
ggsave("03_result/UMAP/Metadata/UMAP_seurat_clusters.pdf")

### !!!!!!! ADD A FACET VERSION !!!!!!! ###
DimPlot(seu, reduction = "umap", group.by = "seurat_clusters") + ggtitle("seurat clusters") +
  facet_wrap(~seurat_clusters)

ggsave("03_result/Clustering/UMAP_seurat_clusters_facet.png")
ggsave("03_result/UMAP/Metadata/UMAP_seurat_clusters_facet.pdf")
```

!!!!!!! ADD A FACET VERSION !!!!!!! !!!!! and save !!!!!

## Clusters quality {.tabset}

### QC metrics per cluster {.tabset}

```{r}
VlnPlot(
  seu,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  group.by = "seurat_clusters",
  pt.size = 0,
  ncol = 3
)

ggsave("03_result/Clustering/Clusters_QC_metrics.png")
```

nCount_RNA: Library size

nFeature_RNA: Library complexity

percent.mt: Mitochondrial content

### Libraries size per cluster {.tabset}

```{r}
# Plot number of cells per cluster with percentage labels of total cells
cell_counts <- seu@meta.data %>%
  count(seurat_clusters, name = "Cell_Count") %>%
  mutate(
    Percentage = round(100 * Cell_Count / sum(Cell_Count), 6),
    label = paste0(Percentage, "%")
  )


ggplot(cell_counts, aes(y = seurat_clusters, x = Cell_Count, fill = seurat_clusters)) +
  geom_col() +
  geom_text(aes(label = label), hjust = -0.1, size = 3) +
  theme_minimal() +
  labs(
    title = "Number (and %) of cells per cluster",
    x = "Number of cells",
    y = "Cluster"
  ) +
  theme(legend.position = "none")

# Save the plot
ggsave("03_result/Clustering/Clusters_cell_counts.png")
```

## Cluster repartition per patients {.tabset}

```{r}
# Proportion of cells per cluster for each patient
cluster_patient <- seu@meta.data %>%
  group_by(id, seurat_clusters) %>%
  summarise(Cell_Count = n()) %>%
  group_by(id) %>%
  mutate(Percentage = Cell_Count / sum(Cell_Count) * 100)

ggplot(cluster_patient, aes(x = id, y = Percentage, fill = seurat_clusters)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Proportion of cells per cluster for each patient",
    x = "Patient ID",
    y = "Proportion of cells",
    fill = "Cluster"
  ) + 
  theme(
      axis.text.x = element_text(angle = 45, 
                                 hjust = 1, 
                                 vjust = 1))

# Save the plot
ggsave("03_result/Clustering/Clusters_repartition_per_patient.png")
```

### Mitochondrial content per cluster {.tabset}

```{r}
stats_mt <- seu@meta.data %>%
  group_by(seurat_clusters) %>%
  summarise(
    mean_pct_mt   = mean(percent.mt),
    median_pct_mt = median(percent.mt),
    sd_pct_mt     = sd(percent.mt)
  )

kable(stats_mt, format = "html") %>%
  kable_styling()

# Save the table
write.csv(stats_mt, 
          file = "03_result/Clustering/cluster_mitochondrial_content_stats.csv", 
          row.names = FALSE)
```

```{r}
ggplot(seu@meta.data,
       aes(
         x = percent.mt,
         y = seurat_clusters,
         fill = seurat_clusters
       )) +
  geom_density_ridges(
    alpha = 0.7,
    scale = 1.2,
    rel_min_height = 0.01,
    color = "white") +
  # theme_ridges() +
  # theme(
  #   legend.position = "none"
  # ) +
  # geom_vline(xintercept = 20, linetype = "dashed", color = "red") +
  labs(
    x = "% mitochondrial genes",
    y = "Clusters")

ggsave("03_result/Clustering/Clusters_mitochondrial_content_ridges.png")
```

Too few cells in clusters 10 to be seen

<!-- Mitcochondrial content on UMAP -->

<!-- ```{r} -->

<!-- FeaturePlot( -->

<!--   seu, -->

<!--   features = "percent.mt", -->

<!--   cols = c("lightgrey", "red"), -->

<!--   reduction = "umap") -->

<!-- ``` -->

## Remove bad clusters {.tabset}

```{r}
# Remove clusters 
seu <- subset(seu, 
              subset = !(seurat_clusters %in% c("8", "9","10")))

seu$seurat_clusters <- factor(seu$seurat_clusters)

```

```{r}
DimPlot(seu, reduction = "umap", group.by = "seurat_clusters") + ggtitle("seurat clusters after removing bad clusters")

ggsave("03_result/Clustering/UMAP_seurat_clusters_after_removing_bad_clusters.png")

DimPlot(seu, reduction = "umap", group.by = "seurat_clusters") + ggtitle("seurat clusters after removing bad clusters") +
  facet_wrap(~seurat_clusters, ncol = (length(unique(seu$seurat_clusters))/2) )

ggsave("03_result/Clustering/UMAP_seurat_clusters_after_removing_bad_clusters_facet.png")

### !!!!!!! ADD A FACET VERSION !!!!!!! ###
### !!!!! and save !!!!! ###
```

!!!!!!! ADD A FACET VERSION !!!!!!! !!!!! and save !!!!!

# Annotate clusters {.tabset}

## FeaturePlot {.tabset}

```{r}
dict <- list(
  "CD16-TotalSeqB" = "FCGR3A",
  "CD335-NKp46-TotalSeqB" = "NCR1",
  "CD197-CCR7-TotalSeqB" = "CCR7",
  "CD4-TotalSeqB" = "CD4",
  "CD8a-TotalSeqB" = "CD8A",
  "CD45RA-TotalSeqB" = "PTPRC",
  "CD62L-TotalSeqB" = "SELL",
  "CD127-IL-7Ra-TotalSeqB" = "IL7R",
  "CD25-TotalSeqB" = "IL2RA",
  "TIGIT-VSTM3-TotalSeqB" = "TIGIT",
  "CD226-DNAM-1-TotalSeqB" = "CD226",
  "CD38-TotalSeqB" = "CD38"
)

dir.create("03_result/FeaturePlot/Prot_RNA_Clusters_FeaturePlot", showWarnings = FALSE)

for (prot in names(dict)) {
  gene <- dict[[prot]]
  
  # ADT plot
  DefaultAssay(seu) <- "ADT"
  p1 <- FeaturePlot(seu, features = prot, reduction = "umap") + ggtitle(paste0(prot, " (ADT)"))
  
  # RNA plot
  DefaultAssay(seu) <- "RNA"
  # Vérifie que le gène existe dans RNA
  if (gene %in% rownames(seu[["RNA"]])) {
    p2 <- FeaturePlot(seu, features = gene, reduction = "umap") + ggtitle(paste0(gene, " (RNA)"))
  } else {
    p2 <- ggplot() + 
      ggtitle(paste0(gene, " (RNA) - Not found")) + 
      theme_void()
  }
  
  p3 <- DimPlot(seu, reduction = "umap", group.by = "seurat_clusters") + ggtitle("seurat_clusters")
  
  # Combine plots
  p <- (p1 | p2) / p3


  
  # Sauvegarde
  ggsave(filename = paste0("03_result/FeaturePlot/Prot_RNA_Clusters_FeaturePlot_", prot, "_", gene, "_featureplot.png"),
         plot = p, width = 12, height = 12)
}
```

# Clusters markers {.tabset}

## Find all markers {.tabset}

```{r}
markers <- FindAllMarkers(seu, only.pos = TRUE, min.pct = 0.25)

markers_filt <- markers %>%
  filter(
    p_val_adj < 0.05,
    avg_log2FC > 0.5,
    !grepl("^MT-|^RPL|^RPS", gene)
  )
```

## Top 10 markers per cluster {.tabset}

```{r}
genes.scaled <- rownames(seu[["RNA"]]@scale.data)

top10 <- markers_filt %>%
  filter(gene %in% genes.scaled) %>%
  group_by(cluster) %>%
  slice_max(avg_log2FC, n = 10)

top10 %>% arrange(cluster, desc(avg_log2FC))
```

## FeaturePlot of top10 markers per cluster {.tabset}

```{r}
# dir.create(
#   "../03_result/FeaturePlot/Top10_Markers_Per_Cluster",
#   recursive = TRUE,
#   showWarnings = FALSE
# )
# 
# for (i in unique(top10$cluster)) {
#   
#   cluster_dir <- paste0(
#     "../03_result/FeaturePlot/Top10_Markers_Per_Cluster/Cluster_", i
#   )
#   dir.create(cluster_dir, showWarnings = FALSE)
#   
#   genes_cluster <- top10 %>%
#     filter(cluster == i) %>%
#     pull(gene)
#   
#   p_list <- lapply(genes_cluster, function(gene) {
#     FeaturePlot(seu, features = gene, reduction = "umap") + 
#       ggtitle(paste0("Cluster ", i, ": ", gene))
#   })
#   
#   p_combined <- wrap_plots(p_list, ncol = 4)
#   
#   ggsave(filename = paste0("../03_result/FeaturePlot/Top10_Markers_Per_Cluster/Cluster_", i, "/Cluster_", i, "_Top10_Markers_featureplot.png"),
#          plot = p_combined, width = 16, height = 12)
#   
#   lapply(genes_cluster, function(gene) {
#     
#     gene_dir <- paste0(cluster_dir, "/Gene_", gene)
#     dir.create(gene_dir, showWarnings = FALSE)
#     
#     violin <- VlnPlot(seu, features = gene, group.by = "seurat_clusters", pt.size = 0) +
#       ggtitle(paste0("Cluster ", i, ": ", gene))
#     
#     ggsave(filename = paste0("../03_result/FeaturePlot/Top10_Markers_Per_Cluster/Cluster_", i, "/Gene_", gene, "/Cluster_", i, "_", gene, "_VlnPlot.png"),
#            plot = violin, width = 6, height = 4)
#     
#     feature <- FeaturePlot(seu, features = gene, reduction = "umap") + 
#       ggtitle(paste0("Cluster ", i, ": ", gene))
#     
#     ggsave(filename = paste0("../03_result/FeaturePlot/Top10_Markers_Per_Cluster/Cluster_", i, "/Gene_", gene, "/Cluster_", i, "_", gene, "_FeaturePlot.png"),
#            plot = feature, width = 6, height = 5)
#     
#     p_cluster <- DimPlot(seu, reduction = "umap", group.by = "seurat_clusters")
#     
#     p_violin_feature_cluster <- (violin | feature) / p_cluster
#     
#     ggsave(filename = paste0("../03_result/FeaturePlot/Top10_Markers_Per_Cluster/Cluster_", i, "/Gene_", gene, "/Cluster_", i, "_", gene, "_Vln_Feature_Cluster.png"),
#            plot = p_violin_feature_cluster, width = 8, height = 10)
#       
#   })
#     
# }
```

## Heatmap of markers per cluster {.tabset}

```{r}
genes <- unique(top10$gene)

# Extract scaled data for the selected genes
mat <- GetAssayData(
  seu,
  assay = "RNA",
  layer = "scale.data"
)[genes, ]

# Order cells by cluster
clusters <- seu$seurat_clusters
ord <- order(clusters)

mat <- mat[, ord]
clusters <- clusters[ord]

# Create annotation for clusters
cluster_ha <- HeatmapAnnotation(
  Cluster = clusters,
  col = list(
    Cluster = structure(
      Seurat::DiscretePalette(length(unique(clusters)), "polychrome"),
      names = levels(clusters)
    )
  )
)
# Create heatmap
heatmap <- Heatmap(
  mat,
  name = "Expression",
  top_annotation = cluster_ha,
  show_column_names = FALSE,
  show_row_names = TRUE,
  cluster_columns = FALSE,
  cluster_rows = TRUE,
  row_names_gp = grid::gpar(fontsize = 8),
  column_title = "Top 10 markers per cluster"
  )


# Save heatmap
dir.create("03_result/Heatmap", showWarnings = FALSE)
draw(heatmap)
png(
  filename = "03_result/Heatmap/Top10_Markers_Per_Cluster_Heatmap.png",
  width = 1200,
  height = 800
)



```

# Automatic annotation (SingleR) {.tabset}

## Make automatic predictions {.tabset}

```{r}
# Prepare SingleR input
sce <- as.SingleCellExperiment(seu, assay = "RNA")

# Reference dataset
ref <- fetchReference("monaco_immune", "2024-02-26")

# Run SingleR
predictions <- SingleR(
  test = sce,
  ref = ref,
  labels = ref$label.fine,
  assay.type.test = "logcounts")

# Add predictions to Seurat object
seu$predicted.celltype.l3_jordan <- predictions$pruned.labels
```

### Visualization of automatic annotations {.tabset}

```{r}
output_dir <- "03_result/Annotation/Automatic"
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

p1 <- DimPlot(seu, reduction = "umap", group.by = "predicted.celltype.l3_jordan") + ggtitle("predicted.celltype.l1")
p1
ggsave(paste0(output_dir, "/UMAP_predicted_celltype_JD.png"), p1)

p2 <- DimPlot(seu, reduction = "umap", group.by = "predicted.celltype.l3_jordan") + ggtitle("predicted.celltype.l3_jordan") + 
  facet_wrap(~predicted.celltype.l3_jordan, ncol = 6)
p2
ggsave(paste0(output_dir, "/UMAP_predicted_celltype_JD_facet.png"), p2)

p3 <- p1 | p2
p3
ggsave(paste0(output_dir, "/UMAP_predicted_celltype_JD_combined.png"), p3, width = 12, height = 6)

```

# Manual annotation {.tabset}

## Create my celltype signatures {.tabset}

### Run presto to find markers {.tabset}

```{r}
expr_matrix <- as.matrix(seu@assays$RNA@data)
labels <- seu$seurat_clusters

presto_markers <- wilcoxauc(expr_matrix, labels)
```

### Construction of the signature table {.tabset}

```{r}
signature_tbl <- tribble(
  ~feature, ~cell_type, ~class,
  
  # Activated
  "CD69", "Activated", "State",
  "TNFRSF9", "Activated", "State",
  "IL2RA", "Activated", "State",

  # Apoptotic / stress
  "MALAT1", "Apoptotic", "State",
  "NEAT1", "Apoptotic", "State",

  # CD4 lineage
  "CD3D", "CD4", "Identity",
  "CD3E", "CD4", "Identity",
  "CD4",  "CD4", "Identity",
  "TRBC1","CD4", "Identity",

  # CD8 lineage
  "FCGR3A","CD8", "Identity",
  "CD3D", "CD8", "Identity",
  "CD3E", "CD8", "Identity",
  "CD8A", "CD8", "Identity",
  "CD8B", "CD8", "Identity",
  "TRBC1","CD8", "Identity",
  "NCR1", "CD8", "Identity",

  # Cytotoxic
  "PRF1","Cytotoxic", "Function",
  "GZMB","Cytotoxic", "Function",
  "GNLY","Cytotoxic", "Function",
  "GZMH","Cytotoxic", "Function",
  "CCL5","Cytotoxic", "Function",
  "NKG7","Cytotoxic", "Function",
  "KLRG1","Cytotoxic", "Function",
  "GZMA","Cytotoxic", "Function",

  # Exhausted
  "PDCD1","Exhausted", "State",
  "LAG3","Exhausted", "State",
  "TOX","Exhausted", "State",
  "TIGIT","Exhausted", "State",
  "CD226","Exhausted", "State",
  "GZMK","Exhausted", "State",
  "HAVCR2","Exhausted", "State",
  "XCL1","Exhausted", "State",
  "XCL2","Exhausted", "State",
  "CD160","Exhausted", "State",

  # HLA
  "HLA-DRA","HLA", "Function",
  "HLA-DRB1","HLA", "Function",

  # IFN / ISG
  "IFNG","IFN", "Function",
  "IFIT2","IFN", "Function",
  "IFIT3","IFN", "Function",
  "ISG15","IFN", "Function",
  "STAT1","IFN", "Function",
  "IFI6","IFN", "Function",
  "IRF7","IFN", "Function",
  "MX1","IFN", "Function",
  "IFITM3","IFN", "Function",

  # Inflammatory / NFkB
  "TNFAIP3","Inflammatory", "Function",
  "NFKB1","Inflammatory", "Function",
  "NFKBIA","Inflammatory", "Function",
  "NFKBIZ","Inflammatory", "Function",
  "STAT4","Inflammatory", "Function",

  # Memory
  "CD44","Memory", "State",
  "PTPRC","Memory", "State",
  "CCR7","Memory", "State",
  "CD27","Memory", "State",
  "CD28","Memory", "State",

  # Naive
  "CCR7","Naif", "State",
  "SELL","Naif", "State",
  "CD27","Naif", "State",
  "CD28","Naif", "State",
  "TCF7","Naif", "State",
  "IL7R","Naif", "State",
  "LEF1","Naif", "State",

  # Proliferating
  "CENPV","Prolif", "State",
  "MKI67","Prolif", "State",
  "NCAPG","Prolif", "State",
  "KIFC1","Prolif", "State",
  "CENPE","Prolif", "State",
  "ORC6","Prolif", "State",

  # Resident (Trm)
  "CD69","Resident", "State",
  "CXCR6","Resident", "State",
  "NR4A2","Resident", "State",
  "RUNX3","Resident", "State",

  # Senescent
  "KLRG1","Senescent", "State",
  "B3GAT1","Senescent", "State",
  "CD27","Senescent", "State",
  "CD28","Senescent", "State",
  "LAMP1","Senescent", "State",
  "IL1A","Senescent", "State",
  "FAS","Senescent", "State",
  "TP53","Senescent", "State",
  "TP53BP1","Senescent", "State",
  "CDKN1A","Senescent", "State",
  "CDKN2A","Senescent", "State",
  "MKI67","Senescent", "State",

  # Tfh
  "IL21","Tfh", "Identity",
  # "FR4","Tfh", "Identity", # Only in mouse
  "CXCR5","Tfh", "Identity",
  "BCL6","Tfh", "Identity",
  "PDCD1","Tfh", "Identity",
  "ICOS","Tfh", "Identity",

  # Th1
  "CX3CR1","Th1", "Identity",
  "IFNG","Th1", "Identity",
  "TBX21","Th1", "Identity",
  "CXCR3","Th1", "Identity",
  "STAT4","Th1", "Identity",
  "IL12RB2","Th1", "Identity",

  # Th2
  "GATA3","Th2", "Identity",
  "CCR4","Th2", "Identity",
  "IL4","Th2", "Identity",
  "IL5","Th2", "Identity",
  "IL13","Th2", "Identity",

  # Th17
  "RORC","Th17", "Identity",
  "IL17A","Th17", "Identity",
  "CCR6","Th17", "Identity",
  "IL17F","Th17", "Identity",
  "IL23R","Th17", "Identity",

  # Treg
  "FOXP3","Treg", "Identity",
  "IL2RA","Treg", "Identity",
  "CTLA4","Treg", "Identity",
  "TIGIT","Treg", "Identity",
  "IKZF2","Treg", "Identity",
  "CD27","Treg", "Identity",
  "IL7R","Treg", "Identity"
)
```

### Visualization of all signatures together {.tabset}

```{r}
Tcell_signature <- "
^CD3D$|^CD3E$|^CD4$|^TRBC1$|
^FCGR3A$|^CD8A$|^CD8B$|^NCR1$|
^CX3CR1$|^IFNG$|^TBX21$|^CXCR3$|^STAT4$|^IL12RB2$|
^GATA3$|^CCR4$|^IL4$|^IL5$|^IL13$|
^RORC$|^IL17A$|^CCR6$|^IL17F$|^IL23R$|
^IL21$|^CXCR5$|^BCL6$|^PDCD1$|^ICOS$|
^FOXP3$|^IL2RA$|^CTLA4$|^TIGIT$|^IKZF2$|^CD27$|^IL7R$|
^PRF1$|^GZMB$|^GNLY$|^GZMH$|^CCL5$|^NKG7$|^KLRG1$|^GZMA$|
^LAG3$|^TOX$|^CD226$|^GZMK$|^HAVCR2$|^XCL1$|^XCL2$|^CD160$|
^B3GAT1$|^CD28$|^LAMP1$|^IL1A$|^FAS$|^TP53$|^TP53BP1$|^CDKN1A$|^CDKN2A$|^MKI67$|
^CD69$|^CXCR6$|^NR4A2$|^RUNX3$|^TNFRSF9$|
^CENPV$|^NCAPG$|^KIFC1$|^CENPE$|^ORC6$|
^CCR7$|^SELL$|^TCF7$|^LEF1$|^CD44$|^PTPRC$|
^MALAT1$|^NEAT1$|
^IFIT2$|^IFIT3$|^ISG15$|^STAT1$|^IFI6$|^IRF7$|^MX1$|^IFITM3$|
^TNFAIP3$|^NFKB1$|^NFKBIA$|^NFKBIZ$|
^HLA-DRA$|^HLA-DRB1$
"


selected_features <- grep(Tcell_signature, rownames(sce), value=TRUE)


m2 <- presto_markers %>%
  filter(feature %in% selected_features) %>%
  select(feature, group, avgExpr, pct_in) %>%
  left_join(signature_tbl, by = "feature") %>%
  mutate(signature_class = case_when(
  cell_type %in% c("CD4","CD8","Th1","Th2","Th17","Tfh","Treg") ~ "Identity",
  cell_type %in% c("Cytotoxic","IFN","Inflammatory","HLA") ~ "Function",
  cell_type %in% c("Naif","Memory","Activated","Exhausted",
                    "Resident","Prolif","Senescent","Apoptotic") ~ "State"
)) %>%
  filter(!is.na(cell_type)) %>%
  group_by(feature) %>%
  mutate(
    avgExpr = avgExpr - mean(avgExpr),
    avgExpr = scales::rescale(avgExpr, to = c(-2,2))
  ) %>%
  ungroup()

ggplot(m2, 
       aes(x = feature,
               y = fct_reorder(as.character(group), group),
               color = avgExpr,
               size = pct_in)) +
  geom_point() +
  scale_color_gradient2(low="blue", mid="white", high="red") +
  facet_wrap(~cell_type, scales="free", ncol=10) +
  theme(
    axis.text.x = element_text(angle=45, hjust=1),
    axis.text.y = element_text(angle=45, hjust=1)
  ) +
  labs(
    title = "LT Compartment - All Signatures",
    x = "Genes markers",
    y = "Clusters"
  )

dir.create("03_result/Annotation/Manual", showWarnings = FALSE)
ggsave(filename = "03_result/Annotation/Manual/Marker_Signatures_LT_Compartment_All_Signatures.png",
       width = 20, height = 10)
```

### View of specific signatures {.tabset}

```{r}
m3 <- presto_markers %>%
  filter(feature %in% selected_features) %>%
  select(feature, group, avgExpr, pct_in) %>%
  left_join(signature_tbl, by = "feature") %>%
  filter(!is.na(cell_type)) %>%
  group_by(feature) %>%
  mutate(
    avgExpr = avgExpr - mean(avgExpr),
    avgExpr = scales::rescale(avgExpr, to = c(-2,2))
  ) %>%
  ungroup()

for (i in unique(m3$class)) {
  p <- ggplot(m3 %>% filter(class == i), 
         aes(x=feature, y=fct_reorder(as.character(group), group), color = avgExpr, size = pct_in)) +
    geom_point()+
    scale_color_gradient2(low = "blue", mid="white",high = "red")+
    # theme_minimal() +
    facet_wrap(~ cell_type, scales = "free", ncol = 4) +
    theme(axis.text.y = element_text(angle=45, hjust=1), 
          axis.text.x = element_text(angle=45, hjust=1)) +
    labs(
    title = paste0("LT Compartment - ", i),
    x = "Genes markers",
    y = "Clusters"
  )
  
  print(p)
  
  dir.create("03_result/Annotation/Manual", showWarnings = FALSE)
  ggsave(filename = paste0("03_result/Annotation/Manual/Marker_Signatures_LT_Compartment_", i, ".png"),
         plot = p, width = 16, height = 8)
}
```

For now I think that we don't need to be to specific on pop annotation because we have all Tcells. For now we have to identified major populations (CD4, CD8) by combining annotation from me + automatic + from PPA to be sure that we analyse the right populations later.

And then we extract count for specific pop cd8 and cd4 cells and perform a new clustering/annotation on each of them separately to have more precise subpopulations.

```{r}
cluster_annotation_list <- list(
  "0" = "CD4_naive_and_memory",
  "1" = "CD8_Cytotox_exhausted",
  "2" = "CD4_inflammed",
  "3" = "CD8_cytotox_low_resident_inflammed",
  "4" = "CD8_naive",
  "5" = "???",
  "6" = "Dying_Tcells",
  "7" = "CD8_IFN"
)

# Add annotations to Seurat object
seu$manual_annotation <- plyr::mapvalues(
  seu$seurat_clusters,
  from = names(cluster_annotation_list),
  to = unlist(cluster_annotation_list)
)

# Names of cells with matching annotations between automatic (mine and PPA SingleR predictions) and manual (mine) annotation
## But first let's harmonize the names between predicted.celltype.l3_jordan and predicted.celltype.l3
seu$predicted.celltype.l3_jordan <- plyr::mapvalues(
  seu$predicted.celltype.l3_jordan,
  from = c("Non-Vd2 gd T cells", "Naive CD4 T cells", "Follicular helper T cells", "Terminal effector CD8 T cells", "Vd2 gd T cells", "MAIT cells", "Terminal effector CD4 T cells", "Th1/Th17 cells", "Th17 cells", "Th1 cells", "Th2 cells", "Effector memory CD8 T cells", "T regulatory cells", "Central memory CD8 T cells", "Naive CD8 T cells", "Natural killer cells", "Myeloid dendritic cells", "Progenitor cells"),
  to = c("Non-Vd2 gd T cells", "Naive CD4 T cells", "Follicular helper T cells CD4", "Terminal effector CD8 T cells", "Vd2 gd T cells", "MAIT cells", "Terminal effector CD4 T cells", "Th1/Th17 cells CD4", "Th17 cells CD4", "Th1 cells CD4", "Th2 cells CD4", "Effector memory CD8 T cells", "T regulatory cells CD4", "Central memory CD8 T cells", "Naive CD8 T cells", "Natural killer cells", "Myeloid dendritic cells", "Progenitor cells")
)



matched_cells <- WhichCells(
  seu,
  expression =
    (
      grepl("CD4", predicted.celltype.l3_jordan) &
      grepl("CD4", predicted.celltype.l1) &
      grepl("CD4", manual_annotation)
    ) |
    (
      grepl("CD8", predicted.celltype.l3_jordan) &
      grepl("CD8", predicted.celltype.l1) &
      grepl("CD8", manual_annotation)
    )
)
length(matched_cells)
length(seu@meta.data$manual_annotation)
# Le nombre de cellules avec annotation "Natural killer cells" dans predicted.celltype.l3_jordan
length(
  predicted_cells <- WhichCells(
    seu,
    expression = (predicted.celltype.l1 == "other T")
))

```

There is a lot of cells annotate gdTcells but I will not focus on them for now as I want to focus on CD4 and CD8 Tcells, and there are too different from cd4+ and cd8+.

# Specific analysis of CD8 Tcells {.tabset}

## Subset CD8 {.tabset}

```{r}
# Subset CD8 Tcells
seu_CD8 <- subset(
  seu,
  subset = (
    # grepl("CD8", predicted.celltype.l3_jordan) &
    # grepl("CD8", manual_annotation &
    grepl("CD8", predicted.celltype.l1) 
  )
)

length(Cells(seu_CD8))
```

```{r}
# Save qs object
qs::qsave(seu_CD8, "02_processed_data/HD_vs_MM_scRNAseq_reanalysis_SCE_CD8_subset.qs")
```

# Specific analysis of CD4 Tcells {.tabset}

## Subset CD4 {.tabset}

```{r}
# Subset CD4 Tcells
seu_CD4 <- subset(
  seu,
  subset = (
    # grepl("CD4", predicted.celltype.l3_jordan) &
    # grepl("CD4", manual_annotation) &
    grepl("CD4", predicted.celltype.l1)
  )
)

length(Cells(seu_CD4))
```

```{r}
# Save qs object
# qs::qsave(seu_CD4, "../02_processed_data/HD_vs_MM_scRNAseq_reanalysis_SCE_CD4_subset.qs")

# Load qs object
seu_CD4 <- qs::qread("02_processed_data/HD_vs_MM_scRNAseq_reanalysis_SCE_CD4_subset.qs")
```

## Reperform QC {.tabset}

### Calculations and vizualizations before cleaning {.tabset}

Maybe for each metrics, create a col in metatadata with status_metric (normal, low, high) based on 3*MAD from median per emulsion AND another col based on isOutlier. Then perform cutting/cutoff based on these status cols. 
Non Stringent QC : remove cells that are outliers only in both of the methods ; 
Stringent QC : remove cells that are outliers in at least one of the methods.

And then 
Global Stringent QC : remove cells that are outliers in at least one of the methods for at least one of the metrics.
Global Non Stringent QC : remove cells that are outliers only in both of the methods for all of the metrics (or critical metrics)

Maybe found critical metrics (not all metrics for QC stringence)

#### Cells count {.tabset}

##### Visualize {.tabset}

```{r}
seu_CD4@meta.data %>% 
  group_by(emulsion) %>%
  summarise(cell_count = n())

features <- c("condition", "id", "emulsion")

  
dir.create("03_result/QC/CD4_Tcells/Cell_Counts", showWarnings = FALSE)

for (feature in features) {
  p <- seu_CD4@meta.data %>%
    ggplot(aes(x = .data[[feature]], fill = .data[[feature]])) +
    geom_bar() +
    # scale_fill_manual(values = pal, drop = FALSE) +
    theme_classic() +
    theme(
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
      plot.title = element_text(hjust = 0.5, face = "bold")
    ) +
    ggtitle(paste("NCells -", feature)) +
    NoLegend()

  print(p)
  
  ggsave(
    filename = paste0(
      "03_result/QC/CD4_Tcells/Cell_Counts/Cell_Counts_By_",
      feature,
      "_Before_QC_CD4_Tcells.png"
    ),
    plot = p,
    width = 6,
    height = 4
  )
}

```

```{r}
DimPlot(seu_CD4, group.by = "emulsion") +
  facet_wrap(~emulsion, ncol = 4) +
  ggtitle("UMAP by emulsion before QC CD4 Tcells") + 
  NoLegend()

ggsave("03_result/QC/CD4_Tcells/Cell_Counts/UMAP_Cell_Counts_by_emulsion_Before_QC_CD4_Tcells.png", width = 10, height = 8)
```

#### PerCell QC Metrics {.tabset}

##### Run {.tabset}

```{r}
qc <- perCellQCMetrics(
  seu_CD4@assays$RNA@counts,
  subsets = list(
    mito = grep("^MT-", rownames(seu_CD4@assays$RNA@counts), value = TRUE),
    ribo = grep("^RPL|^RPS", rownames(seu_CD4@assays$RNA@counts), value = TRUE)
  )
)
seu_CD4 <- AddMetaData(seu_CD4, as.data.frame(qc))
```

##### Mitochondria {.tabset}

###### Visualize {.tabset}

```{r}
ggplot(seu_CD4@meta.data, aes(x = percent.mt, y = emulsion, fill = after_stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01,
    size = 0.2
  ) +
  scale_fill_viridis(
    alpha = 0.9,
    name = "percent.mt",
    option = "C",
    guide = "none"
  ) +
  labs(
    title = "Mitochondrial content by emulsion",
    x = "Percent mitochondrial reads",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.spacing = unit(0.1, "lines"),
    strip.text.x = element_text(size = 8),
    plot.title = element_text(face = "bold")
  )

dir.create("03_result/QC", showWarnings = FALSE)
dir.create("03_result/QC/CD4_Tcells", showWarnings = FALSE)
dir.create("03_result/QC/CD4_Tcells/Mitochondria", showWarnings = FALSE)
ggsave(
  filename = "03_result/QC/CD4_Tcells/Mitochondria/Ridgeplot_mitochondrial_content_by_emulsion_before_QC_CD4_Tcells.png",
  width = 8,
  height = 5
)
```

```{r}
FeaturePlot(
  seu_CD4,
  features = "percent.mt",
  reduction = "umap",
  order = TRUE,
  cols = c("lightgrey", "red")
)

ggsave(
  filename = "03_result/QC/CD4_Tcells/Mitochondria/UMAP_mitochondrial_content_before_QC_CD4_Tcells.png",
  width = 6,
  height = 5
)
```
There are some cells with high mito content in some emulsions.

#### Library size (nCountRNA) {.tabset}

##### Visualize {.tabset}

```{r}
qc_lib_size <- seu_CD4@meta.data %>%
  dplyr::group_by(emulsion) %>%
  dplyr::mutate(
    libsize_status = dplyr::case_when(
      nCount_RNA < median(nCount_RNA) - 3 * mad(nCount_RNA) ~ "low",
      nCount_RNA > median(nCount_RNA) + 3 * mad(nCount_RNA) ~ "high",
      TRUE ~ "normal"
    )
  ) %>%
  dplyr::ungroup() %>% 
  filter(nCount_RNA > 0)

ggplot(qc_lib_size, aes(x = nCount_RNA)) +
  geom_density(fill = "grey80", alpha = 0.6) +
  geom_rug(
    aes(color = libsize_status),
    alpha = 0.6,
    sides = "b"
  ) +
  scale_x_log10() +
  scale_color_manual(
    values = c(
      low = "blue",
      normal = "green",
      high = "red"
    )
  ) +
  facet_wrap(~emulsion) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  ) +
  labs(
    color = "Library size QC",
    x = "nCount_RNA (log10)",
    y = "Density"
  )


dir.create("03_result/QC/CD4_Tcells/Library_Size", showWarnings = FALSE)
ggsave(
  filename = "03_result/QC/CD4_Tcells/Library_Size/DensityPlot_library_size_density_by_emulsion.png",
  width = 10,
  height = 6
)
```

```{r}
ggplot(qc_lib_size, aes(x = emulsion, y = nCount_RNA, color = libsize_status)) +
  geom_violin(fill = "grey90", color = NA) +
  geom_quasirandom(alpha = 0.5, size = 0.8) +
  scale_y_log10() +
  scale_color_manual(
    values = c(low = "blue", normal = "green", high = "red")
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))



ggsave(
  filename = "03_result/QC/CD4_Tcells/Library_Size/ViolinPlot_library_size_by_emulsion.png",
  width = 10,
  height = 6
)
```

#### Library diversity/heterogeneity (nFeatureRNA) {.tabset}

##### Visualize {.tabset}

```{r}
qc_lib_diversity <- seu_CD4@meta.data %>%
  dplyr::group_by(emulsion) %>%
  dplyr::mutate(
    libdiv_status = dplyr::case_when(
      nFeature_RNA < median(nFeature_RNA) - 3 * mad(nFeature_RNA) ~ "low",
      nFeature_RNA > median(nFeature_RNA) + 3 * mad(nFeature_RNA) ~ "high",
      TRUE ~ "normal"
    )
  ) %>%
  dplyr::ungroup() %>% 
  filter(nFeature_RNA > 0)

ggplot(qc_lib_diversity, aes(x = nFeature_RNA)) +
  geom_density(fill = "grey80", alpha = 0.6) +
  geom_rug(
    aes(color = libdiv_status),
    alpha = 0.6,
    sides = "b"
  ) +
  scale_x_log10() +
  scale_color_manual(
    values = c(
      low = "blue",
      normal = "green",
      high = "red"
    )
  ) +
  facet_wrap(~emulsion) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  ) +
  labs(
    color = "Library diversity QC",
    x = "nFeatureRNA (log10)",
    y = "Density"
  )


dir.create("03_result/QC/CD4_Tcells/Library_Diversity", showWarnings = FALSE)
ggsave(
  filename = "03_result/QC/CD4_Tcells/Library_Diversity/DensityPlot_library_diversity_density_by_emulsion.png",
  width = 10,
  height = 6
)
```

```{r}
# ggplot(qc_lib_diversity, aes(x = emulsion, y = nFeature_RNA, color = qc_lib_diversity)) +
#   geom_violin(fill = "grey90", color = NA) +
#   geom_quasirandom(alpha = 0.5, size = 0.8) +
#   scale_y_log10() +
#   scale_color_manual(
#     values = c(low = "blue", normal = "green", high = "red")
#   ) +
#   theme_classic() +
#   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
# 
# ggsave(
#   filename = "03_result/QC/CD4_Tcells/Library_Diversity/ViolinPlot_library_diversity_by_emulsion.png",
#   width = 10,
#   height = 6
# )
```

#### Librairy complexity {.tabset}

##### Run {.tabset}

```{r}
seu_CD4$log10_Complexity <- log10(seu_CD4$nFeature_RNA) / log10(seu_CD4$nCount_RNA)

seu_CD4$low_complexity_outliers <- isOutlier(
  seu_CD4$log10_Complexity,
  nmads = 3,
  type = "lower"
)
seu_CD4$high_complexity_outliers <- isOutlier(
  seu_CD4$log10_Complexity,
  nmads = 3,
  type = "higher"
)

qc_lib_complexity <- seu_CD4@meta.data %>%
  dplyr::group_by(emulsion) %>%
  dplyr::mutate(
    libcomplex_status = dplyr::case_when(
      low_complexity_outliers == TRUE ~ "low",
      high_complexity_outliers == TRUE ~ "high",
      TRUE ~ "normal"
    )
  ) %>%
  dplyr::ungroup() %>% 
  filter(nCount_RNA > 0 & nFeature_RNA > 0)


```

##### Visualize {.tabset}

```{r}
low_quantil_complexity <- quantile(seu_CD4$log10_Complexity, 0.05, na.rm = TRUE)

ggplot(seu_CD4@meta.data, aes(x = log10_Complexity)) +
  geom_density(fill = "lightgreen", alpha = 0.6) +
  geom_vline(xintercept = low_quantil_complexity, linetype = "dashed", color = "red") +
  labs(
    title = "Library complexity (log10GenesPerUMI)",
    subtitle = paste0("Red dashed line = 5% quantile (", round(low_quantil_complexity, 2), ")"),
    x = "log10GenesPerUMI"
  ) +
  theme_classic()

dir.create("03_result/QC/CD4_Tcells/Library_Complexity", showWarnings = FALSE)
ggsave(
  filename = "03_result/QC/CD4_Tcells/Library_Complexity/DensityPlot_library_complexity_before_QC_CD4_Tcells.png",
  width = 8,
  height = 5
)
```

```{r}
ggplot(seu_CD4@meta.data, aes(x = nCount_RNA, y = nFeature_RNA)) +
  geom_point(
    data = subset(seu_CD4@meta.data, !low_complexity_outliers),
    color = "grey70",
    alpha = 0.25,
    size = 0.4
  ) +
  geom_point(
    data = subset(seu_CD4@meta.data, low_complexity_outliers),
    color = "blue3",
    alpha = 0.7,
    size = 0.6
  ) +
  geom_smooth(method = "lm", color = "black", linewidth = 0.6) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = "Library complexity",
    subtitle = "Genes detected vs UMIs (log–log)",
    x = "UMIs per cell",
    y = "Genes per cell",
    color = "QC status"
  ) +
  theme_classic(base_size = 11)

ggsave(
  filename = "03_result/QC/CD4_Tcells/Library_Complexity/ScatterPlot_library_complexity_before_QC_CD4_Tcells.png",
  width = 6,
  height = 5
)

```

```{r}
ggplot(seu_CD4@meta.data,
       aes(x = log10_Complexity, fill = emulsion)) +
  geom_density(alpha = 0.4) +
  geom_vline(
    seu_CD4@meta.data,
    xintercept = median(seu_CD4@meta.data$log10_Complexity) - 3 * mad(seu_CD4@meta.data$log10_Complexity),
    linetype = "dashed",
    color = "blue3"
  ) +
  geom_vline(
    seu_CD4@meta.data,
    xintercept = median(seu_CD4@meta.data$log10_Complexity) + 3 * mad(seu_CD4@meta.data$log10_Complexity),
    linetype = "dashed",
    color = "red3"
  ) +
  facet_wrap(~ emulsion, ncol = 4) +
  labs(
    title = "Library complexity by emulsion",
    x = "log10GenesPerUMI"
  ) +
  theme_classic() +
  theme(legend.position = "none")

ggsave(
  filename = "03_result/QC/CD4_Tcells/Library_Complexity/DensityPlot_library_complexity_by_emulsion_before_QC_CD4_Tcells.png",
  width = 10,
  height = 6
)
```

```{r}
ggplot(seu_CD4@meta.data,
       aes(x = emulsion, y = log10_Complexity)) +
  geom_violin(fill = "grey90", color = NA) +
  geom_quasirandom(alpha = 0.4, size = 0.6) +
  labs(
    title = "Library complexity per emulsion",
    y = "log10GenesPerUMI",
    x = NULL
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(
  filename = "03_result/QC/CD4_Tcells/Library_Complexity/ViolinPlot_library_complexity_by_emulsion_before_QC_CD4_Tcells.png",
  width = 10,
  height = 6
)
```

```{r}
seu_CD4$qc_low_complexity <- isOutlier(
  seu_CD4$log10_Complexity,
  nmads = 3,
  type = "lower"
)

ggplot(seu_CD4@meta.data,
       aes(x = log10_Complexity, fill = qc_low_complexity)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("FALSE" = "grey70", "TRUE" = "red")) +
  theme_classic()

ggsave(
  filename = "03_result/QC/CD4_Tcells/Library_Complexity/DensityPlot_library_complexity_outliers_before_QC_CD4_Tcells.png",
  width = 8,
  height = 5
)
```

#### Cell Cycle Scoring {.tabset}

##### Run {.tabset}

```{r}
seu_CD4 <- CellCycleScoring(
  seu_CD4,
  s.features = cc.genes$s.genes,
  g2m.features = cc.genes$g2m.genes,
  set.ident = FALSE
)
```

#### Outlier detection {.tabset}

##### Run {.tabset}

```{r}
out_low_complexity <- isOutlier(qc$detected, nmads = 3, type = "lower")
out_high_mito <- isOutlier(qc$subsets_mito_percent, nmads = 3, type = "higher")
out_combined <- out_low_complexity | out_high_mito
seu_CD4$out_low_complexity <- out_low_complexity
seu_CD4$out_high_mito <- out_high_mito
seu_CD4$out_combined <- out_combined
```

##### Visualize {.tabset}

```{r}
ggplot(seu_CD4@meta.data, aes(x = detected, y = percent.mt, color = out_combined)) +
  geom_point(alpha = 0.5) +
  theme_classic() +
  labs(color = "Outlier QC") +
  ggtitle("Détection des outliers QC : complexité et mito")


VlnPlot(seu_CD4, features = c("nFeature_RNA", "percent.mt"), group.by = "out_combined", pt.size=0)
```

#### Doublet detection {.tabset}

##### Run {.tabset}

```{r}
sce <- as.SingleCellExperiment(seu_CD4, assay = "RNA")
sce <- scDblFinder(seu_CD4@assays$RNA@counts)
seu_CD4$doublet_score <- sce$scDblFinder.score
seu_CD4$doublet_class <- sce$scDblFinder.class
```

### Cutoff {.tabset}

Cut off BY TECHNICAL BATCH (more qualitative QC than a global cut)

#### Mitochondria {.tabset}

```{r}
cut_mt <- tapply(
  seu_CD4$percent.mt,
  seu_CD4$emulsion,
  function(x) {
    mad_cut <- median(x) + 3 * mad(x) # 3 MADs above the median cutoff
    q_cut   <- quantile(x, 0.95) # 95th percentile cutoff
    min(mad_cut, q_cut) # Choose the more stringent cutoff
  }
)
```

```{r}
cut_mt <- pmin(pmax(cut_mt, 5), 20)  # Ensure cutoffs are between 5% and 20%

seu_CD4$mt_cut <- cut_mt[seu_CD4$emulsion]
seu_CD4 <- subset(seu_CD4, subset = percent.mt <= mt_cut)

table(seu_CD4$emulsion)
```

If the batch is over 15-20% mt, we eliminate this batch

#### Visualization AFTER cutoff {.tabset}

##### Mitochondria {.tabset}

```{r}
ggplot(seu_CD4@meta.data, aes(x = percent.mt, y = emulsion, fill = after_stat(x))) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  scale_fill_viridis(option = "C", guide = "none") +
  # geom_vline(
  #   data = data.frame(
  #     emulsion = names(cut_mt),
  #     cut = cut_mt
  #   ),
  #   aes(xintercept = cut),
  #   linetype = "dashed"
  # ) +
  theme_minimal() +
  labs(
    title = "Post-QC mitochondrial content by batch",
    x = "% mitochondrial reads",
    y = NULL
  )

ggsave(
  filename = "03_result/QC/CD4_Tcells/Mitochondria/Ridgeplot_mitochondrial_content_by_emulsion_after_QC_CD4_Tcells.png",
  width = 8,
  height = 5
)
```

```{r}
FeaturePlot(
  seu_CD4,
  features = "percent.mt",
  reduction = "umap",
  order = TRUE,
  cols = c("lightgrey", "red")
)

ggsave(
  filename = "03_result/QC/CD4_Tcells/Mitochondria/UMAP_mitochondrial_content_after_QC_CD4_Tcells.png",
  width = 6,
  height = 5
)
```

# Session Info {.tabset}

```{r}
sessionInfo()
```
